---
title: "R Notebook"
output: html_notebook
---


Surveys were conducted in 2006/2007 (2007 data), 2007/2008 (2008 data), 2009/2010 (2010 data), 2010/2011 (2011 data), 2014/2015 (2015 data), 2015/2016 (2016 data), and 2019/2020 (2020 data). So, this one (2016), it should be the 6th dataset. They should put Year in their dataset.

Get data and start looking at them:

```{r}
rm(list = ls())
library(tidyverse)
library(lubridate)
library(readr)
source("Granite_Canyon_Counts_fcns.R")

# this is for the 6th data set in the analysis (2016)
YEAR <- 2016
idx.yr <- 6

# # periods for the 6 survey years
periods <-c(136, 135, 164, 178, 179, 151)

#Watch start times, as fraction of a day - stored in a different file
begin <- as.matrix(read.table("Data/begin.txt", 
                              header=T, 
                              nrows = max(periods)))

#watch end times
end <- as.matrix(read.table("Data/end.txt", 
                            header=T,
                            nrows = max(periods)))

obs.list <- read.csv("Data/Observer list.csv", header = T) 
colnames(obs.list) <- c("obs", "ID")

# this file contains all input data for WinBUGS.
v1.out <- readRDS("RData/2006-2019_GC_Formatted_Data.RDS")

# Pull out the information for 2016
periods.2016 <- v1.out$periods[idx.yr]
n.2016 <- v1.out$n[1:periods.2016,,idx.yr]
n.com.2016 <- v1.out$n.com[1:periods.2016,,idx.yr]
n.sp.2016 <- v1.out$n.sp[1:periods.2016,,idx.yr]
obs.2016 <- v1.out$obs[1:periods.2016,,idx.yr]

vs.2016 <- v1.out$vs[1:periods.2016,idx.yr]
bf.2016 <- v1.out$bf[1:periods.2016,idx.yr]
day.2016 <- v1.out$day[1:periods.2016,idx.yr]

FinalData.v1 <- data.frame(begin = begin[1:periods[idx.yr], idx.yr],
                           end = end[1:periods[idx.yr], idx.yr],
                           bf = bf.2016,
                           vs = vs.2016,
                           n = n.2016[,1],
                           obs = obs.2016[,1],
                           BeginDay = day.2016,
                           v = "V1")

```


```{r}
# This contains the results from my version
v2.out <- readRDS("RData/out_2016_Tomo_v2.rds")
FinalData.v2 <- v2.out$FinalData %>% 
  mutate(v = "V2") %>% 
  left_join(obs.list, by = "obs") %>%
  select(-c(dur, ff, i, BeginHr)) 

# find if there is NA in ID - not in the look up table  
ID.NA <- filter(FinalData.v2, is.na(ID))

unique.ID.NA <- unique(ID.NA$obs)

if (length(unique.ID.NA) > 0){
  for (k in 1:length(unique.ID.NA)){
    FinalData.v2[FinalData.v2$obs == unique.ID.NA[k], "ID"] <- max(obs.list$ID) + k
    
  }
  
}

# replace column names
FinalData.v2 %>% select(-obs) %>%
  mutate(obs = ID) %>%
  select(-ID) -> FinalData.v2

# rearrange the columns to match v1
FinalData.v2 <- FinalData.v2[, names(FinalData.v1)]
FinalData.Both <- rbind(FinalData.v2, FinalData.v1)


```


Compare how...


```{r}
ggplot(FinalData.Both) +
  geom_point(aes(x = begin, y = n, color = v),
             alpha = 0.5)
```


```{r}
min.begin <- min(floor(FinalData.Both$begin))
max.begin <- max(ceiling(FinalData.Both$begin))

time.steps <- min.begin:max.begin
difs <- data.frame(begin = double(),
                   end = double(),
                   min.begin = double(), 
                   max.end = double(), 
                   n.periods = integer(), 
                   max.bf = integer(), 
                   max.vs = integer(), 
                   total.whales = integer(),
                   time.step = integer(),
                   stringsAsFactors = F)

c <- k <- 1
for (k in 1:(length(time.steps)-1)){
  tmp <- filter(FinalData.Both, begin >= time.steps[k] & begin < time.steps[k+1])
  if (nrow(tmp) > 0){
    tmp %>% filter(v == "V1") -> tmp.1
    tmp %>% filter(v == "V2") -> tmp.2
    
    difs[c,] <- c(min(tmp$begin), 
                  max(tmp$end),
                  min(tmp.1$begin) - min(tmp.2$begin), 
                  max(tmp.1$end) - max(tmp.2$end),
                  nrow(tmp.1) - nrow(tmp.2),
                  max(tmp.1$bf) - max(tmp.1$bf),
                  max(tmp.1$vs) - max(tmp.1$vs),
                  sum(tmp.1$n) - sum(tmp.2$n),
                  time.steps[k])
    c <- c + 1
    
  }
  
}


```


There are some differences in the number of whales detected per day. Look at one at a time to see what made the difference.

```{r}
difs %>% filter(n.periods != 0 | total.whales != 0) -> difs.1
FinalData.Both %>% mutate(time.steps = floor(FinalData.Both$begin)) -> FinalData.Both

v2.out$Data_Out %>% 
  mutate(time.steps = floor(v2.out$Data_Out$begin)) -> Data_Out.v2 

v2.out$CorrectLength %>%
  mutate(time.steps = floor(v2.out$CorrectLength$begin)) -> CorrectLength.v2 

difs.1[1,]
```

First one has one more period in V1 than in V2.

```{r}
idx <- 1
FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V1") -> tmp.V1

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# The first and last shifts are missing in V2.
Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

# the begin date is different; 30.3125 in V1, whereas it is 30.35417 in V2
data <- get.data("Data/", YEAR, ff = 1)

# The first line is 30.35417... where did 30.3125 come from?
# Find the corresponding calendar date:
TempHr <- (tmp.V1$begin[1] - floor(tmp.V1$begin[1])) * 24
TempDay <- (tmp.V1$begin[1] - (tmp.V1$begin[1] - floor(tmp.V1$begin[1]))) + mdy(paste0("11/30/", (YEAR - 1)))

# Start Hr is 7.5 or 7:30... That's not right. 
CorrectLength.v2 %>% 
    filter(time.steps == difs.1[idx, "time.step"]) 

# so, the first shift lasted only 30 minutes, which was kicked out by the "correct length" filter.
# Observers changed but the shift continued. Having different observers made it to not work in my
# version... I may remove the line 013 (P) to combine the two shifts.


```


```{r}
difs.1[2,]
```


This has been fixed...
The second one has 1 more whales in V2 than in V1. 

<!-- ```{r} -->
<!-- idx <- 2   -->
<!-- FinalData.Both %>%  -->
<!--   filter(time.steps == difs.1[idx, "time.step"]) %>%  -->
<!--   filter(v == "V1") -> tmp.V1 -->

<!-- FinalData.Both %>%  -->
<!--   filter(time.steps == difs.1[idx, "time.step"]) %>%  -->
<!--   filter(v == "V2") -> tmp.V2 -->

<!-- # The sixth shift has one less whale in V1 (19) than in V2 (20) -->
<!-- Data_Out.v2 %>%  -->
<!--   filter(time.steps == difs.1[idx, "time.step"])  -->

<!-- # ff = 3 -->
<!-- data <- get.data("Data/", YEAR, ff = 3) -->

<!-- # i = 6: -->
<!-- out.shift <- get.shift(YEAR, data, ff=2, i=6) -->

<!-- out.shift$data %>% -->
<!--   select(V5, V9) %>% -->
<!--   mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>% -->
<!--   group_by(V5) %>% -->
<!--   summarise(max.n = max(V9), -->
<!--             last.n = last(V9), -->
<!--             d.max.last = max(V9) - last(V9)) -> data.summary.6.v2 -->
<!-- # It's not anything to do with max vs last...  -->

<!-- # Get the raw data files -->
<!-- data %>% filter(begin >= tmp.V1[6,"begin"] & begin <= tmp.V1[6, "end"]) %>% -->
<!-- #data.3.5 %>%  -->
<!--   filter(V2 == "S") %>% -->
<!--   select(V5, V9) %>% -->
<!--   mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>% -->
<!--   group_by(V5) %>% -->
<!--   summarise(max.n = max(V9), -->
<!--             last.n = last(V9), -->
<!--             d.max.last = max(V9) - last(V9)) -> data.summary.6.v1 -->

<!-- # There were only 12 groups sighted in V1, whereas 13 groups were sighted in V2. -->
<!-- # Group ID 69 is missing from V1, which was sighted 1 second prior to the end of -->
<!-- # the shift... should that be excluded? -->
<!-- out.shift$data.shift -->

<!-- ``` -->


```{r}
idx <- 2
difs.1[idx,]
```

one more period in V1 than in V2 with 19 more whales in V1 than in V2

```{r}
FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V1") -> tmp.V1

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# V2 is missing the first shift
Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 
FF <- 4
# There were four shifts recorded, where the first and last had shorter than 1.5 hrs
# So, those were excluded when filtering for the shift duration
CorrectLength.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

data <- get.data("Data/", YEAR, ff = FF)

shift <-  1
out.shift.1 <- get.shift(YEAR, data, ff=FF, i=shift)


shift <-  4
out.shift.4 <- get.shift(YEAR, data, ff=FF, i=shift)

out.shift.1$out.df

# So... somehow shift 1 was included even though the shift only lasted < 1 hr. 
# The raw file shows the shift started at 8 a.m.and ended at 8:58:47. The second shift
# started at 9:00:39 and ended 10:30:36. So, somehow, the number of hours for the first
# shift was inflated to be 1.5 hrs! 
```



```{r}
idx <- 3
difs.1[idx,]
```

The fourth one has 2 more periods in V1 than V2 with 8 more whales.

```{r}

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V1") -> tmp.V1

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# V2 is missing the first and second shifts
Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

# The first and second shifts are too short... 
FF <- 5

CorrectLength.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

data <- get.data("Data/", YEAR, ff = FF)

shift <-  1
out.shift.1 <- get.shift(YEAR, data, ff=FF, i=shift)
# The first shift started at 8:30:06, ended at 8:56:51, so < 1 hr

shift <-  2
out.shift.2 <- get.shift(YEAR, data, ff=FF, i=shift)
# Started at 10:34:53 and ended 10:56:57, so < 1 hr

```



```{r}
idx <- 4
difs.1[idx,]
```

1 more period in V2 than V1 with 14 more whales.

```{r}

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V1") -> tmp.V1

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# V1 is missing the third shift and one less whale in the last (6th) shift
Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

# V2 is missing the first and second shifts
Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 
# all shifts have enough time

# The first and second shifts are too short... 
FF <- 6

data <- get.data("Data/", YEAR, ff = FF)

shift <-  3
out.shift.3 <- get.shift(YEAR, data, ff=FF, i=shift)
# I don't know why this shift was thrown out... 

shift <-  6
out.shift.6 <- get.shift(YEAR, data, ff=FF, i=shift)

# Get the raw data files
data %>% filter(begin >= tmp.V1[5,"begin"] & begin <= tmp.V1[5, "end"]) %>% 
#data.3.5 %>% 
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  # select(V5) %>% 
  # unique() -> group.ID.V1
  summarise(max.n = max(V9),
            last.n = last(V9),
            d.max.last = max(V9) - last(V9)) -> data.summary.6.v1

# not max vs last problem.

# The last group  is 73 in V1
out.shift.6$data %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  #unique() -> group.ID.V2
  group_by(V5) %>%
  summarise(max.n = max(V9),
            last.n = last(V9),
            d.max.last = max(V9) - last(V9)) -> data.summary.6.v2

data.summary.6.v1 %>% right_join(data.summary.6.v2, by = "V5") %>%
  mutate(d.max.n = max.n.x - max.n.y,
         d.last.n = last.n.x - last.n.y) %>%
  select(d.max.n, d.last.n) -> data.summary.6

# They have the same numbers of whales per group... then why are there one less
# whale in V1? When I add max.n or last.n in the raw data file (data.summary.6.v1),
# they add up to 38... 

# Not sure what's going on here... 
```





```{r}
idx <- 5
difs.1[idx,]
```

1 more whale in V2 than V1.

```{r}

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V1") -> tmp.V1

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# V1 is missing the third shift and one less whale in the last (6th) shift
Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

FF <- 11

# the last shift (6) has one more whale in V2 than in V1

data <- get.data("Data/", YEAR, ff = FF)

shift <-  6
out.shift.6 <- get.shift(YEAR, data, ff=FF, i=shift)
# I don't know why this shift was thrown out... 

data %>% 
  filter(begin >= tmp.V1$begin[shift] & begin <= tmp.V1$end[shift]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarise(last.n = last(V9),
            max.n = max(V9)) -> summary.11.6.V1
  
c(sum(summary.11.6.V1$last.n), sum(summary.11.6.V1$max.n))

# No difference in max vs last counts
# There were 23 (length(unique(summary.11.6.V1$V5))) groups in raw data

out.shift.6$data %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarise(last.n = last(V9),
            max.n = max(V9)) -> summary.11.6.V2

c(sum(summary.11.6.V2$last.n), sum(summary.11.6.V2$max.n))
# There were 23 (length(unique(summary.11.6.V2$V5))) groups from V2 extraction
# I have no idea why V1 has 1 less whale...
  
```



```{r}
idx <- 6
difs.1[idx,]
```

2 more whale in V2 than V1.

```{r}

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V1") -> tmp.V1

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# The last shift (6) has 2 more whales in V2. Another last shift... something is off...
# I can't reproduce their results because the old script does not run any longer.

Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

FF <- 12

# the last shift (6) has one more whale in V2 than in V1

data <- get.data("Data/", YEAR, ff = FF)

shift <-  6
out.shift.6 <- get.shift(YEAR, data, ff=FF, i=shift)

data %>% 
  filter(begin >= tmp.V1$begin[shift] & begin <= tmp.V1$end[shift]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarise(last.n = last(V9),
            max.n = max(V9)) -> summary.12.6.V1
  
c(sum(summary.12.6.V1$last.n), sum(summary.12.6.V1$max.n))

# No difference in max vs last counts (both are 32) - why did v1 ended up 
# with 30 whales? 

# There were 22 (length(unique(summary.12.6.V1$V5))) groups in raw data

out.shift.6$data %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarise(last.n = last(V9),
            max.n = max(V9)) -> summary.12.6.V2

c(sum(summary.12.6.V2$last.n), sum(summary.12.6.V2$max.n))
# There were 22 (length(unique(summary.12.6.V2$V5))) groups from V2 extraction
# I have no idea why V1 has 2 less whales...

summary.12.6.V2 %>% 
  left_join(summary.12.6.V1, by = "V5") %>%
  mutate(d.max.n = max.n.x - max.n.y,
         d.last.n = last.n.x - last.n.y) %>%
  select(V5, d.max.n, d.last.n) -> summary.12.6

# there is no difference here either... 
```




```{r}
idx <- 7
difs.1[idx,]
```

1 more shift and 27 more whales in V1 than V2. Nothing returns from V2

```{r}

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V1") -> tmp.V1


FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

CorrectLength.v2 %>% 
    filter(time.steps == difs.1[idx, "time.step"]) 
# There were two shifts but one has short period and the other had vs = 5. So,
# nothing should come back from this day. 

Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

FF <- 14

# the last shift (6) has one more whale in V2 than in V1

data <- get.data("Data/", YEAR, ff = FF)

shift <-  1
out.shift.1 <- get.shift(YEAR, data, ff=FF, i=shift)

shift <-  2
out.shift.2 <- get.shift(YEAR, data, ff=FF, i=shift)

# Raw data file has the starting time of 13:40:05, which was changed to
# 13:30:00 to include this shift in the analysis. 

# Changing the "grace period" to 10 minutes would include this shift without
# modifying the data.
```




```{r}
idx <- 8
difs.1[idx,]
```

1 more shift and 19 more whales in V1 than V2. 

```{r}

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V1") -> tmp.V1


FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

CorrectLength.v2 %>% 
    filter(time.steps == difs.1[idx, "time.step"]) 

# The begin time for the first shift was modified. 
# Data file indicates the start time was 7:43:13 but in the data, it was changed
# to 7:30:00 (fractional_Day2YMDhms(tmp.V1$begin[1],YEAR))

# Observer code was different (40 in V1 vs 42 in V2)

FF <- 17

data <- get.data("Data/", YEAR, ff = FF)

shift <-  1
out.shift.1 <- get.shift(YEAR, data, ff=FF, i=shift)


```


V1 has 13 more whales than V1 - Wrong entries of "South" in "EDITED GW160202_071118.dat" (FF=22)
Some lines were fixed to align entries. No VS were entered in rows from 005 to 024. NA entered. This has been fixed.



```{r}
idx <- 9
difs.1[idx,]
```


V1 has one more period and one more whale than V2.

```{r}

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V1") -> tmp.V1


FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# the fourth period is missing in V2

Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

FF <- 29
data <- get.data("Data/", YEAR, ff = FF)

shift <-  4
out.shift.4 <- get.shift(YEAR, data, ff=FF, i=shift)
# this shift lasted only 34 minutes (12:00:17 to 12:34.28) but was included
# in the analysis?

```




<!-- ```{r} -->
<!-- idx <- 11 -->
<!-- difs.1[idx,] -->
<!-- ``` -->

The tenth one has 12 more whales in V1 than V2. This has been fixed by modifying the data file. It seems that wrong information is recorded in V13, which is supposed to be  visibility (VS) but "South" is entered in many... that may be the problem

I edited the data file to enter NA where no VS was entered. The file was renamed with _TE added to the end of the file name. It's possible that this may have been caused also by misalignment of entries, where reticle readings (V7) did not have sufficient digits in (e.g., 0.8 instead of 0.80). 

Now V2 has one more whales than V1.

<!-- ```{r} -->

<!-- FinalData.Both %>%  -->
<!--   filter(time.steps == difs.1[idx, "time.step"]) %>%  -->
<!--   filter(v == "V1") -> tmp.V1 -->


<!-- FinalData.Both %>%  -->
<!--   filter(time.steps == difs.1[idx, "time.step"]) %>%  -->
<!--   filter(v == "V2") -> tmp.V2 -->

<!-- # the last shift (6) has one more whale in V2. Again the last shift!  -->

<!-- Data_Out.v2 %>%  -->
<!--   filter(time.steps == difs.1[idx, "time.step"])  -->

<!-- FF <- 22 -->

<!-- data <- get.data("Data/", YEAR, ff = FF) -->

<!-- shift <-  6 -->
<!-- out.shift.6 <- get.shift(YEAR, data, ff=FF, i=shift) -->

<!-- data %>%  -->
<!--   filter(begin >= tmp.V1$begin[shift] & begin <= tmp.V1$end[shift]) %>% -->
<!--   filter(V2 == "S") %>% -->
<!--   select(V5, V9) %>% -->
<!--   mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>% -->
<!--   group_by(V5) %>% -->
<!--   summarise(last.n = last(V9), -->
<!--             max.n = max(V9)) -> summary.22.6.V1 -->

<!-- c(sum(summary.22.6.V1$last.n), sum(summary.22.6.V1$max.n)) -->

<!-- out.shift.6$data %>% -->
<!--   select(V5, V9) %>% -->
<!--   mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>% -->
<!--   group_by(V5) %>% -->
<!--   summarise(last.n = last(V9), -->
<!--             max.n = max(V9)) -> summary.22.6.V2 -->

<!-- c(sum(summary.22.6.V2$last.n), sum(summary.22.6.V2$max.n)) -->

<!-- summary.22.6.V2 %>%  -->
<!--   left_join(summary.22.6.V1, by = "V5") %>% -->
<!--   mutate(d.max.n = max.n.x - max.n.y, -->
<!--          d.last.n = last.n.x - last.n.y) %>% -->
<!--   select(V5, d.max.n, d.last.n) -> summary.22.6 -->

<!-- # group 59 was not included in V1, which was observed at 16:30:03, which was a tiny -->
<!-- # bit (3 seconds) after the supposed end of the shift...  -->
<!-- ``` -->


<!-- ```{r} -->
<!-- idx <- 12 -->
<!-- difs.1[idx,] -->
<!-- ``` -->

Three more whales in V2 than in V1

<!-- ```{r} -->

<!-- FinalData.Both %>%  -->
<!--   filter(time.steps == difs.1[idx, "time.step"]) %>%  -->
<!--   filter(v == "V1") -> tmp.V1 -->

<!-- FinalData.Both %>%  -->
<!--   filter(time.steps == difs.1[idx, "time.step"]) %>%  -->
<!--   filter(v == "V2") -> tmp.V2 -->

<!-- # One less whale in shift 1 and two less whales in shift 5 in V1 -->

<!-- Data_Out.v2 %>%  -->
<!--   filter(time.steps == difs.1[idx, "time.step"])  -->

<!-- FF <- 23 -->
<!-- data <- get.data("Data/", YEAR, ff = FF) -->

<!-- shift <-  1 -->
<!-- out.shift.1 <- get.shift(YEAR, data, ff=FF, i=shift) -->

<!-- out.shift.1$data %>% -->
<!--   select(V5, V9) %>% -->
<!--   mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>% -->
<!--   group_by(V5) %>% -->
<!--   summarise(last.n = last(V9), -->
<!--             max.n = max(V9)) -> summary.23.1.V2 -->

<!-- data %>%  -->
<!--   filter(begin >= tmp.V1$begin[shift] & begin <= tmp.V1$end[shift]) %>% -->
<!--   filter(V2 == "S") %>% -->
<!--   select(V5, V9) %>% -->
<!--   mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>% -->
<!--   group_by(V5) %>% -->
<!--   summarise(last.n = last(V9), -->
<!--             max.n = max(V9)) -> summary.23.1.V1 -->
<!-- # This returns 19... rather than 18 as in tmp.V1 -->

<!-- shift <-  5 -->
<!-- out.shift.5 <- get.shift(YEAR, data, ff=FF, i=shift) -->

<!-- out.shift.5$data %>% -->
<!--   select(V5, V9) %>% -->
<!--   mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>% -->
<!--   group_by(V5) %>% -->
<!--   summarise(last.n = last(V9), -->
<!--             max.n = max(V9)) -> summary.23.5.V2 -->

<!-- data %>%  -->
<!--   filter(begin >= tmp.V1$begin[shift] & begin <= tmp.V1$end[shift]) %>% -->
<!--   filter(V2 == "S") %>% -->
<!--   select(V5, V9) %>% -->
<!--   mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>% -->
<!--   group_by(V5) %>% -->
<!--   summarise(last.n = last(V9), -->
<!--             max.n = max(V9)) -> summary.23.5.V1 -->

<!-- #V1 HAS MORE WHALES THAN V2...  -->

<!-- summary.22.6.V2 %>%  -->
<!--   left_join(summary.22.6.V1, by = "V5") %>% -->
<!--   mutate(d.max.n = max.n.x - max.n.y, -->
<!--          d.last.n = last.n.x - last.n.y) %>% -->
<!--   select(V5, d.max.n, d.last.n) -> summary.22.6 -->




<!-- ``` -->

```{r}


# the 4th shift did not have long enough period (0.0306 days = 0.73 hrs)
# the 4th and 5th shifts were combined in V1. I'm going to fix the data file.
# The new data file will have _TE at the end. 
# 044 E	02/11/2016	12:44:54	
# 046 B	02/11/2016	12:54:35
# 047 P	02/11/2016	12:54:35	JWG	MSL	N
# were deleted.

```

