---
title: "R Notebook"
output: html_notebook
---

Compares output from WinBugs runs in January and July. Should we expect to see differences in estimates when we add another year's worth of data?


```{r}
rm(list=ls())
library(tidyverse)
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

compute.Nhat <- function(GW_Nmix, seasons, data.str){
  # Daily.Est <- GW_Nmix$sims.list$Daily.Est
  # sp <- GW_Nmix$sims.list$sp
  # com <- GW_Nmix$sims.list$com

  Corrected.Est <- GW_Nmix$sims.list$Corrected.Est
  abundance.df <- data.frame(Season = seasons, 
                             Mean = apply(Corrected.Est,
                                          FUN = mean,
                                          MARGIN = 2),
                              Median = apply(Corrected.Est, 
                                             FUN = median, 
                                             MARGIN = 2),
                              LCL = apply(Corrected.Est, 
                                          MARGIN = 2, 
                                          FUN = quantile, 0.025),
                              UCL = apply(Corrected.Est, 
                                          MARGIN = 2, 
                                          FUN = quantile, 0.975),
                             year = seq(1, length(seasons)),
                             Data = data.str)
  return(abundance.df)

  # stats.list <- vector(mode = "list",
  #                          length = dim(Daily.Est)[3])
  # 
  # for (k in 1:dim(Daily.Est)[3]){
  # 
  #   stats.list[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est[,,k], 2,
  #                                                          median),
  #                                 Daily.Est.LCL = apply(Daily.Est[,,k], 2,
  #                                                       quantile,0.275),
  #                                 Daily.Est.UCL = apply(Daily.Est[,,k], 2,
  #                                                       quantile,0.975),
  #                                 sp.median = apply(exp(sp[,,k]), 2,
  #                                                   median),
  #                                 sp.LCL = apply(exp(sp[,,k]), 2,
  #                                                quantile,0.025),
  #                                 sp.UCL = apply(exp(sp[,,k]), 2,
  #                                                quantile,0.975),
  #                                 com.median = apply(exp(com[,,k]), 2,
  #                                                    median),
  #                                 com.LCL = apply(exp(com[,,k]), 2,
  #                                                 quantile,0.025),
  #                                 com.UCL = apply(exp(com[,,k]), 2,
  #                                                 quantile,0.975),
  #                                 #total.median = apply(exp(sp[,,k]), 1, sum),
  #                                 days = 1:dim(Daily.Est)[2],
  #                                 year = k)
  # }
  # 
  # all.stats <- do.call("rbind", stats.list) %>% group_by(year)
  # return(all.stats)
}

# results from Jan 2022
# comes from GW Rebuild for WinBUGS TE.Rmd
load("RData/GW BUGS 6yr 100k.Rdata")

abundance.df.6yr.v1 <- compute.Nhat(GW_Nmix, seasons = seasons[1:6], data.str = "v1_6yr")


```


7 year data

```{r}
rm(list = ls()[!(ls() %in% c("abundance.df.6yr.v1", "compute.Nhat", "seasons"))])

load("RData/GW BUGS 7yr 100k.Rdata")

abundance.df.7yr.v0 <- compute.Nhat(GW_Nmix, seasons = seasons[1:7], data.str = "v0_7yr")


```


Move to 8-yr data

```{r}
rm(list = ls()[!(ls() %in% c("abundance.df.6yr.v1", 
                             "abundance.df.7yr.v0",
                             "compute.Nhat", "seasons"))])

# Comes from WinBUGS Ver2.Rmd
# the RDS file is a list and BUGS_out object in the list should be the same
# structure as GW_Nmix object in the .RData objects above. 
out.8yr.v1 <- readRDS("RData/WinBUGS_8yr_v1.rds")

GW_Nmix <- out.8yr.v1$BUGS_out
abundance.df.8yr.v1 <- compute.Nhat(GW_Nmix, seasons, data.str = "v1_8yr")


```



```{r}
rm(list = ls()[!(ls() %in% c("abundance.df.6yr.v1", "abundance.df.7yr.v0",
                             "abundance.df.8yr.v1", "compute.Nhat", "seasons"))])

# Comes from WinBUGS Ver2.Rmd
# the RDS file is a list and BUGS_out object in the list should be the same
# structure as GW_Nmix object in the .RData objects above. 
out.8yr.v2 <- readRDS("RData/WinBUGS_8yr_v2.rds")

GW_Nmix <- out.8yr.v2$BUGS_out
abundance.df.8yr.v2 <- compute.Nhat(GW_Nmix, seasons, data.str = "v2_8yr")

```

7 yr - this is actually 6 years worth of data. 

```{r}
rm(list = ls()[!(ls() %in% c("abundance.df.6yr.v1", "abundance.df.7yr.v0",
                             "abundance.df.8yr.v1",
                             "abundance.df.8yr.v2", "compute.Nhat", "seasons"))])

out <- readRDS("RData/WinBUGS_4yr_v1.rds")

GW_Nmix <- out$BUGS_out
abundance.df.4yr.v1 <- compute.Nhat(GW_Nmix, seasons[1:4], data.str = "v1_4yr")

```



```{r}
rm(list = ls()[!(ls() %in% c("abundance.df.6yr.v1",
                             "abundance.df.7yr.v0",
                             "abundance.df.8yr.v1",
                             "abundance.df.8yr.v2", 
                             "abundance.df.4yr.v1", 
                             "compute.Nhat", "seasons"))])

out <- readRDS("RData/WinBUGS_4yr_v2.rds")

GW_Nmix <- out$BUGS_out
abundance.df.4yr.v2 <- compute.Nhat(GW_Nmix, seasons[1:4], data.str = "v2_4yr")

```


```{r}
rm(list = ls()[!(ls() %in% c("abundance.df.6yr.v1",
                             "abundance.df.8yr.v1",
                             "abundance.df.8yr.v2",
                             #"abundance.df.7yr",
                             "abundance.df.4yr.v1",
                             "abundance.df.4yr.v2",
                             "compute.Nhat", "seasons"))])

load("RData/GW BUGS 4yr 100k v0.RData")

abundance.df.4yr.v0 <- compute.Nhat(GW_Nmix, seasons[1:4], data.str = "v0_4yr")

```


```{r}
rm(list = ls()[!(ls() %in% c("abundance.df.6yr.v1",
                             "abundance.df.7yr.v0",
                             "abundance.df.8yr.v1",
                             "abundance.df.8yr.v2",
                             #"abundance.df.7yr",
                             "abundance.df.4yr.v1",
                             "abundance.df.4yr.v2",
                             "abundance.df.4yr.v0",
                             "compute.Nhat", "seasons"))])

load("RData/GW BUGS 6yr 100k v0.RData")
abundance.df.6yr.v0 <- compute.Nhat(GW_Nmix, seasons[1:6], data.str = "v0_6yr")

```



```{r}
# # seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
# #              "2014/2015", "2015/2016", "2019/2020", "2021/2022")
# 
# # summary.fcn <- function(x, seasons){
# #   x %>% 
# #     group_by(year) %>%
# #     summarize(Nhat = sum(Daily.Est.median),
# #               LCL = sum(Daily.Est.LCL),
# #               UCL = sum(Daily.Est.UCL)) %>%
# #     mutate(season = seasons) -> out.df
# #   return(out.df)
# # }
# 
# 
# annual.Nhat.6yr.v1 <- abundance.df.6yr %>%
#   mutate(Data = "v1_6yr")
# annual.Nhat.6yr.v0 <- summary.fcn(abundance.df.6yr.v0,
#                                    seasons[1:max(abundance.df.6yr.v0$year)])%>%
#   mutate(Data = "v0_6yr")
# annual.Nhat.8yr.v1 <- summary.fcn(abundance.df.8yr.v1,
#                                    seasons[1:max(abundance.df.8yr.v1$year)]) %>%
#   mutate(Data = "v1_8yr")
# annual.Nhat.8yr.v2 <- summary.fcn(abundance.df.8yr.v2,
#                                    seasons[1:max(abundance.df.8yr.v2$year)])%>%
#   mutate(Data = "v2_8yr")
# annual.Nhat.4yr.v1 <- summary.fcn(abundance.df.4yr.v1,
#                                    seasons[1:max(abundance.df.4yr.v1$year)])%>%
#   mutate(Data = "v1_4yr")
# annual.Nhat.4yr.v0 <- summary.fcn(abundance.df.4yr.v0,
#                                    seasons[1:max(abundance.df.4yr.v0$year)])%>%
#   mutate(Data = "v0_4yr")
# annual.Nhat.4yr.v2 <- summary.fcn(abundance.df.4yr.v2,
#                                    seasons[1:max(abundance.df.4yr.v2$year)])%>%
#   mutate(Data = "v2_4yr")
# 
# # annual.Nhat.7yr <- summary.fcn(abundance.df.7yr,
# #                                    seasons[1:max(abundance.df.7yr$year)])

durban.data <- data.frame(Season = c("2006/2007", "2007/2008", 
                                     "2009/2010", "2010/2011"),
                          Median = c(20750, 17820, 21210, 20990),
                          LCL = c(18860, 16150, 19420, 19230),
                          UCL = c(23320, 19920, 23250, 22900), 
                          year = c(1,2,3,4),
                          
                          Data = "Durban")
```


All 6-yr analysis results in the same results, as expected. Results from 8-yr analyses are different between v1 and v2, as well as from the 6-yr analysis. I need to compare 6-yr analysis and 7-yr analysis. 

```{r}
abundance.df <- rbind(abundance.df.4yr.v0 %>% dplyr::select(!Mean),
                    abundance.df.4yr.v1 %>% dplyr::select(!Mean),
                    abundance.df.4yr.v2 %>% dplyr::select(-Mean),
                    abundance.df.6yr.v0 %>% dplyr::select(-Mean),
                    abundance.df.6yr.v1 %>% dplyr::select(-Mean),
                    abundance.df.8yr.v1 %>% dplyr::select(-Mean),
                    abundance.df.8yr.v2 %>% dplyr::select(-Mean), 
                    durban.data)


ggplot(data = abundance.df, 
       aes(x = Season, y = Median, color = Data, group = Data)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(x = Season, ymin = LCL, ymax = UCL, color = Data)) 

```



