---
title: "R Notebook"
output: html_notebook
---


Changed the file name to v0_v2 from v1_v2. 2022-03-25

Surveys were conducted in 2006/2007 (2007 data), 2007/2008 (2008 data), 2009/2010 (2010 data), 2010/2011 (2011 data), 2014/2015 (2015 data), 2015/2016 (2016 data), and 2019/2020 (2020 data). So, this one (2015), it should be the 5th dataset. They should put Year in their dataset.

Get data and start looking at them:

```{r}
rm(list = ls())
library(tidyverse)
library(lubridate)
library(readr)
source("Granite_Canyon_Counts_fcns.R")

# this is for the 5th data set in the analysis (2015)
YEAR <- 2015
idx.yr <- 5
obs.list <- read.csv("Data/Observer list.csv", header = T) 
colnames(obs.list) <- c("obs", "ID")

# # periods for the 6 survey years
periods <-c(136, 135, 164, 178, 179, 151)

out.list <- compare.V0.V2.BUGSinput(YEAR, idx.yr, periods, obs.list)


```


There are some differences in the number of whales detected per day. Look at one at a time to see what made the difference.


```{r}
idx <- 1
out.list$difs.1[idx,]
```

Three more whales in V2 than in V0.

```{r}

out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

out.list$Data_Out.v2 %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) 

# From the output, I see ff = 1.

FF <- 1

data <- get.data("Data/", YEAR, ff = FF)

#The last shift (4) has 3 more whales in V2 than in V2#

shift <- 4
out.shift.4 <- get.shift(YEAR, data, ff=FF, i=shift)

data %>% 
  filter(begin >= tmp.V0[shift,"begin"] & begin <= tmp.V0[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V0

data %>% 
  filter(begin >= tmp.V2[shift,"begin"] & begin <= tmp.V2[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V2

# These add up to 25, which is 3 more than V0 contained... no difference here...

```




```{r}
idx <- 2
out.list$difs.1[idx,]
```

3 more whales in v.2 than in v.0

```{r}
out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# comparing tmp.V0 and tmp.V2 results in 3 more whales in the last shift (6)
# V0 = 25 and V2 = 28

out.list$Data_Out.v2 %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) 

# output shows the file id is 9

FF <- 9
data <- get.data("Data/", YEAR, ff = FF)

shift <- 6
out.shift.6 <- get.shift(YEAR, data, ff=FF, i=shift)

# V0
data %>% 
  filter(begin >= tmp.V0[shift,"begin"] & begin <= tmp.V0[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V0

# V2
data %>% 
  filter(begin >= tmp.V2[shift,"begin"] & begin <= tmp.V2[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V2

```

The difference comes from having the difference in the end time, where V0 ended at `r fractional_Day2YMDhms(tmp.V0[6,"end"], 2015)$hms` and missed the last sighting `r summary.V2[nrow(summary.V2), "V5"]` of `r summary.V2[nrow(summary.V2), "n"]` whales. 

```{r}
idx <- 3
out.list$difs.1[idx,]
```

One more period and 12 more whales in V2 than in V0.

```{r}

out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# Shift 2 is missing in V0
out.list$Data_Out.v2 %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) 

FF <- 10 
data <- get.data("Data/", YEAR, ff = FF)

shift <- 2
out.shift.2 <- get.shift(YEAR, data, ff=FF, i=shift)

data %>% 
  filter(begin >= tmp.V2[shift,"begin"] & 
           begin <= tmp.V2[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V2

out.shift.2$data.shift

# I don't know why this one was eliminated in V0. It's possible that BFT and
# VS was 5 at the end of the period (> 30 seconds from the end), which 
# triggered the exclusion... That's exactly what happened. 
```

It's possible that BFT and VS was 5 at the end of the period (> 30 seconds from the end), which  triggered the exclusion... That's exactly what happened. 


```{r}
idx <- 4
out.list$difs.1[idx,]
```

One more shift and 10 more whales in V2 than in V0.

```{r}
# This file contained the following problem first. 
# The shift started at 7:30:36 and ended 10:26:44, which was way too long... 
# There should have been a shift change at 9 a.m., which was not recorded.
# This was fixed with _TE added to the file name (2/3/2015)

# After fixing the problem, V2 now has an extra shift with 10 whales.
# I don't know how these were treated in V0... 

out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# Comparing tmp.V0 and tmp.V2, the second shift (65.37554 - 65.43750) is missing
# in V0 
out.list$Data_Out.v2 %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) 
# FF is 26

FF <- 26
data <- get.data("Data/", YEAR, ff = FF)

shift <- 2
out.shift.2 <- get.shift(YEAR, data, ff=FF, i=shift)

data %>% 
  filter(begin >= tmp.V2[shift,"begin"] & begin <= tmp.V2[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V2

# I don't know why this period was removed in V0. - because of the shift that was too long.

```



```{r}
idx <- 5
out.list$difs.1[idx,]
```

One more period and 4 more whales in V0 than in V2.

```{r}

out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# The first shift with 4 whales is missing in V2
out.list$Data_Out.v2 %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) 
# FF is 30

# The data shows that first shift was from 71.31314 to 71.31346 (7:31:22). The second shift
# was from 71.41697 (10:00) to 71.43763. The summarized data from .RData has the first shift
# to be from 71.3750 (9:00) to 71.4375 (10:30). In the raw data file, there is no information
# between 7:31:23 and 10:00:26. So, the first shift in V0 was fabricated... 
FF <- 30
data <- get.data("Data/", YEAR, ff = FF)

shift <- 1

data %>% 
  filter(begin >= tmp.V0[shift,"begin"] & begin <= tmp.V0[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V0

# shift <- 2
out.shift.1 <- get.shift(YEAR, data, ff=FF, i=shift)

# The problem was changing Bft from 4 to 5 at 10:01:36, which was only one hour
# into the shift. The shift continued to 10:30:51, but no sighting happened 
# after the change in Bft. So, this should have been excluded from the analysis.
# The new extraction code needs to be fixed to include Bft that is too high
# but the most of the shift is completed, but exclude Bft change during the sfhit. 

```



```{r}
idx <- 6
out.list$difs.1[idx,]
```

One more whale in V0 than in V2. THIS NEEDS TO BE EXAMINED MORE. 2022-03-30

```{r}
out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

out.list$FinalData.Both %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# The second shift has one more whale (14 vs 13) in V0 
out.list$Data_Out.v2 %>% 
  filter(time.steps == out.list$difs.1[idx, "time.step"]) 
# FF is 31

FF <- 31
data <- get.data("Data/", YEAR, ff = FF)

shift <- 2
out.shift.2 <- get.shift(YEAR, data, ff=FF, i=shift)

out.shift.2$data %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarise(last.n = last(V9),
            max.n = max(V9)) -> summary.31.2.V2

# sum of last.n's in summary.31.2.V2 is 13. How did V0 come up with 14?

data %>% 
  filter(begin >= tmp.V0[shift,"begin"] & begin <= tmp.V0[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V0


data %>% 
  filter(begin >= tmp.V2[shift,"begin"] & begin <= tmp.V2[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V2


```

I think what happened here is that Group 22 (n = 1 for southbound and n = 2 for northbound) was included in V0 (just the southbound recording), whereas it was excluded in V2. 

That's it for 2015. 


