---
title: "R Notebook"
output: html_notebook
---

Compares output from WinBugs runs in January and July. Should we expect to see differences in estimates when we add another year's worth of data?


```{r}
rm(list=ls())
library(tidyverse)

compute.stats <- function(GW_Nmix){
  Daily.Est <- GW_Nmix$sims.list$Daily.Est
  sp <- GW_Nmix$sims.list$sp
  com <- GW_Nmix$sims.list$com
  
  stats.list <- vector(mode = "list",
                           length = dim(Daily.Est)[3])

  for (k in 1:dim(Daily.Est)[3]){
  
    stats.list[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est[,,k], 2,
                                                           median),
                                  Daily.Est.LCL = apply(Daily.Est[,,k], 2,
                                                        quantile,0.275),
                                  Daily.Est.UCL = apply(Daily.Est[,,k], 2,
                                                        quantile,0.975),
                                  sp.median = apply(exp(sp[,,k]), 2,
                                                    median),
                                  sp.LCL = apply(exp(sp[,,k]), 2,
                                                 quantile,0.025),
                                  sp.UCL = apply(exp(sp[,,k]), 2,
                                                 quantile,0.975),
                                  com.median = apply(exp(com[,,k]), 2,
                                                     median),
                                  com.LCL = apply(exp(com[,,k]), 2,
                                                  quantile,0.025),
                                  com.UCL = apply(exp(com[,,k]), 2,
                                                  quantile,0.975),
                                  #total.median = apply(exp(sp[,,k]), 1, sum),
                                  days = 1:dim(Daily.Est)[2],
                                  year = k)
  }
  
  all.stats <- do.call("rbind", stats.list) %>% group_by(year)
  return(all.stats)
}

# results from Jan 2022
# comes from GW Rebuild for WinBUGS TE.Rmd
load("RData/GW BUGS 6yr 100k.Rdata")

all.stats.6yr <- compute.stats(GW_Nmix)


```



Move to 8-yr data

```{r}
rm(list = ls()[!(ls() %in% c("all.stats.6yr", "compute.stats"))])

# Comes from WinBUGS Ver2.Rmd
# the RDS file is a list and BUGS_out object in the list should be the same
# structure as GW_Nmix object in the .RData objects above. 
out.8yr.v1 <- readRDS("RData/WinBUGS_8yr_v1.rds")

GW_Nmix <- out.8yr.v1$BUGS_out
all.stats.8yr.v1 <- compute.stats(GW_Nmix)


```



```{r}
rm(list = ls()[!(ls() %in% c("all.stats.6yr",
                             "all.stats.8yr.v1", "compute.stats"))])

# Comes from WinBUGS Ver2.Rmd
# the RDS file is a list and BUGS_out object in the list should be the same
# structure as GW_Nmix object in the .RData objects above. 
out.8yr.v2 <- readRDS("RData/WinBUGS_8yr_v2.rds")

GW_Nmix <- out.8yr.v2$BUGS_out
all.stats.8yr.v2 <- compute.stats(GW_Nmix)

```

7 yr - this is actually 6 years worth of data. 

```{r}
rm(list = ls()[!(ls() %in% c("all.stats.6yr",
                             "all.stats.8yr.v1",
                             "all.stats.8yr.v2", "compute.stats"))])
# results from July 2022
# comes from GW Rebuild for WinBUGS TE v2.Rmd

```



```{r}
rm(list = ls()[!(ls() %in% c("all.stats.6yr",
                             "all.stats.8yr.v1",
                             "all.stats.8yr.v2", "compute.stats"))])

out <- readRDS("RData/WinBUGS_4yr_v1.rds")

GW_Nmix <- out$BUGS_out
all.stats.4yr.v1 <- compute.stats(GW_Nmix)

```



```{r}
rm(list = ls()[!(ls() %in% c("all.stats.6yr",
                             "all.stats.8yr.v1",
                             "all.stats.8yr.v2", 
                             "all.stats.4yr.v1", 
                             "compute.stats"))])

out <- readRDS("RData/WinBUGS_4yr_v2.rds")

GW_Nmix <- out$BUGS_out
all.stats.4yr.v2 <- compute.stats(GW_Nmix)

```


```{r}
rm(list = ls()[!(ls() %in% c("all.stats.6yr",
                             "all.stats.8yr.v1",
                             "all.stats.8yr.v2",
                             #"all.stats.7yr",
                             "all.stats.4yr.v1",
                             "all.stats.4yr.v2",
                             "compute.stats"))])

load("RData/GW BUGS 4yr 100k v0.RData")
all.stats.4yr.v0 <- compute.stats(GW_Nmix)

```


```{r}
rm(list = ls()[!(ls() %in% c("all.stats.6yr",
                             "all.stats.8yr.v1",
                             "all.stats.8yr.v2",
                             #"all.stats.7yr",
                             "all.stats.4yr.v1",
                             "all.stats.4yr.v2",
                             "all.stats.4yr.v0",
                             "compute.stats"))])

load("RData/GW BUGS 6yr 100k v0.RData")
all.stats.6yr.v0 <- compute.stats(GW_Nmix)

```



```{r}
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

summary.fcn <- function(x, seasons){
  x %>% 
    group_by(year) %>%
    summarize(Nhat = sum(Daily.Est.median),
              LCL = sum(Daily.Est.LCL),
              UCL = sum(Daily.Est.UCL)) %>%
    mutate(season = seasons) -> out.df
  return(out.df)
}


annual.Nhat.6yr.v1 <- summary.fcn(all.stats.6yr,
                                   seasons[1:max(all.stats.6yr$year)])%>%
  mutate(Data = "v1_6yr")
annual.Nhat.6yr.v0 <- summary.fcn(all.stats.6yr.v0,
                                   seasons[1:max(all.stats.6yr.v0$year)])%>%
  mutate(Data = "v0_6yr")
annual.Nhat.8yr.v1 <- summary.fcn(all.stats.8yr.v1,
                                   seasons[1:max(all.stats.8yr.v1$year)]) %>%
  mutate(Data = "v1_8yr")
annual.Nhat.8yr.v2 <- summary.fcn(all.stats.8yr.v2,
                                   seasons[1:max(all.stats.8yr.v2$year)])%>%
  mutate(Data = "v2_8yr")
annual.Nhat.4yr.v1 <- summary.fcn(all.stats.4yr.v1,
                                   seasons[1:max(all.stats.4yr.v1$year)])%>%
  mutate(Data = "v1_4yr")
annual.Nhat.4yr.v0 <- summary.fcn(all.stats.4yr.v0,
                                   seasons[1:max(all.stats.4yr.v0$year)])%>%
  mutate(Data = "v0_4yr")
annual.Nhat.4yr.v2 <- summary.fcn(all.stats.4yr.v2,
                                   seasons[1:max(all.stats.4yr.v2$year)])%>%
  mutate(Data = "v2_4yr")

# annual.Nhat.7yr <- summary.fcn(all.stats.7yr,
#                                    seasons[1:max(all.stats.7yr$year)])

durban.data <- data.frame(year = c(1,2,3,4),
                          Nhat = c(20750, 17820, 21210, 20990),
                          LCL = c(18860, 16150, 19420, 19230),
                          UCL = c(23320, 19920, 23250, 22900), 
                          season = c("2006/2007", "2007/2008", 
                                     "2009/2010", "2010/2011"),
                          Data = "Durban")
```


All 6-yr analysis results in the same results, as expected. Results from 8-yr analyses are different between v1 and v2, as well as from the 6-yr analysis. I need to compare 6-yr analysis and 7-yr analysis. 

```{r}
annual.Nhat <- rbind(annual.Nhat.4yr.v0,
                    annual.Nhat.4yr.v1,
                    annual.Nhat.4yr.v2,
                    annual.Nhat.6yr.v0,
                    annual.Nhat.6yr.v1,
                    annual.Nhat.8yr.v1,
                    annual.Nhat.8yr.v2, 
                    durban.data)


ggplot(data = annual.Nhat) +
  geom_point(aes(x = season, y = Nhat, color = Data)) +
  geom_errorbar(aes(x = season, ymin = LCL, ymax = UCL, color = Data)) 

```



