---
title: "R Notebook"
output: html_notebook
---


Get data and start looking at them:

```{r}
rm(list = ls())
source("Granite_Canyon_Counts_fcns.R")
library(tidyverse)
library(lubridate)

YEAR <- 2022

#GO THROUGH THIS FILE ONCE MORE TO MAKE SURE EVERYTHING LOOKS OKAY. 2022-03-21

v3.out <- readRDS("RData/out_2022_Joshs.rds")
v2.out <- readRDS("RData/out_2022_Tomo_v2.rds")
v1.out <- readRDS("RData/out_2022_Tomo_v1.rds")
FinalData.v2 <- v2.out$FinalData %>% mutate(v = "V2")%>% select(-dur)
FinalData.v1 <- v1.out$FinalData %>% mutate(v = "V1") 
FinalData.v3 <- v3.out$FinalData %>% mutate(v = "V3") 
FinalData.all <- rbind(FinalData.v1, FinalData.v2, FinalData.v3)

# dim(FinalData.v1)
# dim(FinalData.v2)

FinalData.v1 %>% 
  group_by(ff) %>% 
  summarize(nrow = n()) -> v1.summary

FinalData.v2 %>% 
  group_by(ff) %>% 
  summarize(nrow = n()) -> v2.summary

FinalData.v3 %>% 
  group_by(ff) %>% 
  summarize(nrow = n()) -> v3.summary

v1.summary %>% 
  left_join(v2.summary, by = "ff") %>% 
  left_join(v3.summary, by = "ff") %>%
  mutate(dif_1vs2 = nrow.x - nrow.y,
         dif_1vs3 = nrow.x - nrow,
         dif_2vs3 = nrow.y - nrow) -> v1Vsv2VsV3_n


```


ff = 12 has different numbers between 1vs2 and 2vs 3, meaning 2 is different from 1 and 3. Looking at the vX.summary objects, there were 6 rows (shifts) in ff=12 according to v1 and v3. v2 returned only 5 rows. ff = 12 is for 2022-01-12. Using FinalData.all %>% filter(ff = 12) revealed that v2 missed the second shift that started at 9. 

```{r}

data.12 <- get.data("Data/", YEAR = 2022, ff = 12)
data.12.2 <- get.shift(2022, data.12, ff = 12, i = 2)

```

These two functions return the right set of shifts, even the second shift that started at 9. The difference is that in the old version (Josh's and V1), visibility code was taken only from sightings (Column 13), whereas the new version takes it from the visibility lines (event = V). This does not affect too many but for this instance, VS was 5 at the beginning of the shift, which was changed to 4 at 9:29:55. The first sighting happened at 9:35:47. So, using the old method, VS was always 4. I'm not sure which is the right way to do it. 

In my opinion, first 30 min should be removed from the analysis and the shift starts at 9:29:55. 

```{r}
obs.list <- read.csv("Data/Observer list.csv", header = T) 
colnames(obs.list) <- c("obs", "ID")

FinalData.v3 <- v3.out$FinalData %>% mutate(v = "V3") 
FinalData.v2 <- v2.out$FinalData %>% mutate(v = "V2") %>% select(-dur)
FinalData.v1 <- v1.out$FinalData %>% mutate(v = "V1") 

FinalData.v2 %>% 
  left_join(obs.list, by = "obs") -> FinalData.v2

# find if there is NA in ID - not in the look up table  
ID.NA <- filter(FinalData.v2, is.na(ID))

unique.ID.NA <- unique(ID.NA$obs)

if (length(unique.ID.NA) > 0){
  for (k in 1:length(unique.ID.NA)){
    FinalData.v2[FinalData.v2$obs == unique.ID.NA[k], "ID"] <- max(obs.list$ID) + k
    
  }
  
}

# replace column names
FinalData.v2 %>% select(-obs) %>%
  mutate(obs = ID) %>%
  select(-ID) -> FinalData.v2

# rearrange the columns to match v1
FinalData.v2 <- FinalData.v2[, names(FinalData.v1)]
FinalData.v3 <- FinalData.v3[, names(FinalData.v1)]
FinalData.Both <- rbind(FinalData.v2, FinalData.v1)
FinalData.All <- rbind(FinalData.v1, FinalData.v2, FinalData.v3)
```



```{r}
min.begin <- min(floor(FinalData.Both$begin))
max.begin <- max(ceiling(FinalData.Both$begin))

time.steps <- min.begin:max.begin
difs <- data.frame(begin = double(),
                   end = double(),
                   min.begin = double(), 
                   max.end = double(), 
                   n.periods = integer(), 
                   max.bf = integer(), 
                   max.vs = integer(), 
                   total.whales = integer(),
                   time.step = integer(),
                   stringsAsFactors = F)

c <- k <- 1
for (k in 1:(length(time.steps)-1)){
  tmp <- filter(FinalData.Both, begin >= time.steps[k] & begin < time.steps[k+1])
  if (nrow(tmp) > 0){
    tmp %>% filter(v == "V1") -> tmp.1
    tmp %>% filter(v == "V2") -> tmp.2
    
    difs[c,] <- c(min(tmp$begin), 
                  max(tmp$end),
                  min(tmp.1$begin) - min(tmp.2$begin), 
                  max(tmp.1$end) - max(tmp.2$end),
                  nrow(tmp.1) - nrow(tmp.2),
                  max(tmp.1$bf) - max(tmp.1$bf),
                  max(tmp.1$vs) - max(tmp.1$vs),
                  sum(tmp.1$n) - sum(tmp.2$n),
                  time.steps[k])
    c <- c + 1
    
  }
  
}
```



```{r}
difs %>% filter(n.periods != 0 | total.whales != 0) -> difs.1
FinalData.Both %>% mutate(time.steps = floor(FinalData.Both$begin)) -> FinalData.Both

v2.out$Data_Out %>% 
  mutate(time.steps = floor(v2.out$Data_Out$begin)) -> Data_Out.v2 

v2.out$CorrectLength %>%
  mutate(time.steps = floor(v2.out$CorrectLength$begin)) -> CorrectLength.v2 

```

There is only one. 3 more whales in V2 than in V1.


```{r}
idx <- 1
difs.1[idx,]
```



```{r}
FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V1") -> tmp.V1

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# Nothing returned from V1 or V2


FF <- 13
data <- get.data("Data/", YEAR, ff = FF)
# 1/13/2022 - wrong data entry on line 89 - fixed and saved with _TE
# Also changed the "E" entry time from 15:00:00 to 15:00:40 because
# there was a sighting at 15:00:38

shift <- 5
out.shift.5 <- get.shift(YEAR, data, ff = FF, i = shift)


data %>%  
  filter(begin >= tmp.V1[5,"begin"] & begin <= tmp.V1[5, "end"]) %>% 
  filter(V2 == "S") %>% 
  select(V5, V9) %>% 
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>% 
  group_by(V5) %>% 
  summarize(n = last(V9)) -> summary.13.5.V1

# there are 31 whales here, and so as V2. Why did V1 compute as 28 whales?


out.shift.5$data %>% group_by(V5) %>%
  mutate(V9 = as.numeric(V9)) %>%
  summarize(n = last(V9)) -> summary.13.5.V2

# Group 53 (3) should not come in because it is found in the next shift, which was
# too short to be included. 2022-03-21
```


