---
title: "R Notebook"
output: html_notebook
---


Changed the file name to v0_v2 from v1_v2. 2022-03-25

Surveys were conducted in 2006/2007 (2007 data), 2007/2008 (2008 data), 2009/2010 (2010 data), 2010/2011 (2011 data), 2014/2015 (2015 data), 2015/2016 (2016 data), and 2019/2020 (2020 data). So, this one (2015), it should be the 5th dataset. They should put Year in their dataset.

Get data and start looking at them:

```{r}
rm(list = ls())
library(tidyverse)
library(lubridate)
library(readr)
source("Granite_Canyon_Counts_fcns.R")

# this is for the 5th data set in the analysis (2015)
YEAR <- 2015
idx.yr <- 5

# # periods for the 6 survey years
periods <-c(136, 135, 164, 178, 179, 151)

#Watch start times, as fraction of a day - stored in a different file
begin <- as.matrix(read.table("Data/begin.txt", 
                              header=T, 
                              nrows = max(periods)))

#watch end times
end <- as.matrix(read.table("Data/end.txt", 
                            header=T,
                            nrows = max(periods)))

obs.list <- read.csv("Data/Observer list.csv", header = T) 
colnames(obs.list) <- c("obs", "ID")

# this file contains all input data for WinBUGS.
V0.out <- readRDS("RData/2006-2019_GC_Formatted_Data.RDS")

# Pull out the information for 2015
periods.2015 <- V0.out$periods[idx.yr]
n.2015 <- V0.out$n[1:periods.2015,,idx.yr]
n.com.2015 <- V0.out$n.com[1:periods.2015,,idx.yr]
n.sp.2015 <- V0.out$n.sp[1:periods.2015,,idx.yr]
obs.2015 <- V0.out$obs[1:periods.2015,,idx.yr]

vs.2015 <- V0.out$vs[1:periods.2015,idx.yr]
bf.2015 <- V0.out$bf[1:periods.2015,idx.yr]
day.2015 <- V0.out$day[1:periods.2015,idx.yr]

FinalData.V0 <- data.frame(begin = begin[1:periods[idx.yr], idx.yr],
                           end = end[1:periods[idx.yr], idx.yr],
                           bf = bf.2015,
                           vs = vs.2015,
                           n = n.2015[,1],
                           obs = obs.2015[,1],
                           BeginDay = day.2015,
                           v = "V0")

```


```{r}
# This contains the results from my version
v2.out <- readRDS("RData/out_2015_Tomo_v2.rds")
FinalData.v2 <- v2.out$FinalData %>% 
  mutate(v = "V2") %>% 
  left_join(obs.list, by = "obs") %>%
  select(-c(dur, ff, i, BeginHr)) 

# find if there is NA in ID - not in the look up table  
ID.NA <- filter(FinalData.v2, is.na(ID))

unique.ID.NA <- unique(ID.NA$obs)

if (length(unique.ID.NA) > 0){
  for (k in 1:length(unique.ID.NA)){
    FinalData.v2[FinalData.v2$obs == unique.ID.NA[k], "ID"] <- max(obs.list$ID) + k
    
  }
  
}

# replace column names
FinalData.v2 %>% select(-obs) %>%
  mutate(obs = ID) %>%
  select(-ID) -> FinalData.v2

# rearrange the columns to match V0
FinalData.v2 <- FinalData.v2[, names(FinalData.V0)]
FinalData.Both <- rbind(FinalData.v2, FinalData.V0)

v2.out$Data_Out %>% 
  mutate(time.steps = floor(v2.out$Data_Out$begin)) -> Data_Out.v2 

v2.out$CorrectLength %>%
  mutate(time.steps = floor(v2.out$CorrectLength$begin)) -> CorrectLength.v2 

```


```{r}
min.begin <- min(floor(FinalData.Both$begin))
max.begin <- max(ceiling(FinalData.Both$begin))

time.steps <- min.begin:max.begin
difs <- data.frame(begin = double(),
                   end = double(),
                   min.begin = double(), 
                   max.end = double(), 
                   n.periods = integer(), 
                   max.bf = integer(), 
                   max.vs = integer(), 
                   total.whales = integer(),
                   time.step = integer(),
                   stringsAsFactors = F)

c <- k <- 1
for (k in 1:(length(time.steps)-1)){
  tmp <- filter(FinalData.Both, begin >= time.steps[k] & begin < time.steps[k+1])
  if (nrow(tmp) > 0){
    tmp %>% filter(v == "V0") -> tmp.1
    tmp %>% filter(v == "V2") -> tmp.2
    
    difs[c,] <- c(min(tmp$begin), 
                  max(tmp$end),
                  min(tmp.1$begin) - min(tmp.2$begin), 
                  max(tmp.1$end) - max(tmp.2$end),
                  nrow(tmp.1) - nrow(tmp.2),
                  max(tmp.1$bf) - max(tmp.1$bf),
                  max(tmp.1$vs) - max(tmp.1$vs),
                  sum(tmp.1$n) - sum(tmp.2$n),
                  time.steps[k])
    c <- c + 1
    
  }
  
}


```


There are some differences in the number of whales detected per day. Look at one at a time to see what made the difference.

```{r}
difs %>% filter(n.periods != 0 | total.whales != 0) -> difs.1
FinalData.Both %>% mutate(time.steps = floor(FinalData.Both$begin)) -> FinalData.Both


```



```{r}
idx <- 1
difs.1[idx,]
```

Three more whales in V2 than in V0.

```{r}

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2


Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

# From the output, I see ff = 1.

FF <- 1

data <- get.data("Data/", YEAR, ff = FF)

#The last shift (4) has 3 more whales in V2 than in V2#

shift <- 4
out.shift.4 <- get.shift(YEAR, data, ff=FF, i=shift)

data %>% 
  filter(begin >= tmp.V0[shift,"begin"] & begin <= tmp.V0[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V0

data %>% 
  filter(begin >= tmp.V2[shift,"begin"] & begin <= tmp.V2[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V2

# These add up to 25, which is 3 more than V0 contained... no difference here...

```




```{r}
idx <- 2
difs.1[idx,]
```

3 more whales in v.2 than in v.0

```{r}
FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# comparing tmp.V0 and tmp.V2 results in 3 more whales in the last shift (6)
# V0 = 25 and V2 = 28

Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

# output shows the file id is 9

FF <- 9
data <- get.data("Data/", YEAR, ff = FF)

shift <- 6
out.shift.6 <- get.shift(YEAR, data, ff=FF, i=shift)

# V0
data %>% 
  filter(begin >= tmp.V0[shift,"begin"] & begin <= tmp.V0[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V0

# V2
data %>% 
  filter(begin >= tmp.V2[shift,"begin"] & begin <= tmp.V2[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V2

```

The difference comes from having the difference in the end time, where V0 ended at `r fractional_Day2YMDhms(tmp.V0[6,"end"], 2015)$hms` and missed the last sighting `r summary.V2[nrow(summary.V2), "V5"]` of `r summary.V2[nrow(summary.V2), "n"]` whales. 

```{r}
idx <- 3
difs.1[idx,]
```

One more period and 12 more whales in V2 than in V0.

```{r}

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# Shift 2 is missing in V0
Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 

FF <- 10 
data <- get.data("Data/", YEAR, ff = FF)

shift <- 2
out.shift.2 <- get.shift(YEAR, data, ff=FF, i=shift)

data %>% 
  filter(begin >= tmp.V2[shift,"begin"] & 
           begin <= tmp.V2[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V2

out.shift.2$data.shift

# I don't know why this one was eliminated in V0. It's possible that BFT and
# VS was 5 at the end of the period (> 30 seconds from the end), which 
# triggered the exclusion... That's exactly what happened. 
```

It's possible that BFT and VS was 5 at the end of the period (> 30 seconds from the end), which  triggered the exclusion... That's exactly what happened. 


```{r}
idx <- 4
difs.1[idx,]
```

One more shift and 10 more whales in V2 than in V0.

```{r}
# This file containd the following problem first. 
# The shift started at 7:30:36 and ended 10:26:44, which was way too long... 
# There should have been a shift change at 9 a.m., which was not recorded.
# This was fixed with _TE added to the file name (2/3/2015)

# After fixing the problem, V2 now has an extra shift with 10 whales.
# I don't know how these were treated in V0... 

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# Comparing tmp.V0 and tmp.V2, the second shift (65.37554 - 65.43750) is missing
# in V0 
Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 
# FF is 26

FF <- 26
data <- get.data("Data/", YEAR, ff = FF)

shift <- 2
out.shift.2 <- get.shift(YEAR, data, ff=FF, i=shift)

data %>% 
  filter(begin >= tmp.V2[shift,"begin"] & begin <= tmp.V2[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V2

# I don't know why this period was removed in V0. 
```



```{r}
idx <- 5
difs.1[idx,]
```

One more period and 4 more whales in V0 than in V2.

```{r}

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# The first shift with 4 whales is missing in V2
Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 
# FF is 30

# The data shows that first shift was from 71.31314 to 71.31346 (7:31:22). The second shift
# was from 71.41697 (10:00) to 71.43763. The summarized data from .RData has the first shift
# to be from 71.3750 (9:00) to 71.4375 (10:30). In the raw data file, there is no information
# between 7:31:23 and 10:00:26. So, the first shift in V0 was fabricated... 
FF <- 30
data <- get.data("Data/", YEAR, ff = FF)

shift <- 1

data %>% 
  filter(begin >= tmp.V0[shift,"begin"] & begin <= tmp.V0[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V0

# shift <- 2
out.shift.1 <- get.shift(YEAR, data, ff=FF, i=shift)

# The problem was changing Bft from 4 to 5 at 10:01:36, which was only one hour
# into the shift. The shift continued to 10:30:51, but no sighting happened 
# after the change in Bft. So, this should have been excluded from the analysis.
# The new extraction code needs to be fixed to include Bft that is too high
# but the most of the shift is completed, but exclude Bft change during the sfhit. 

```



```{r}
idx <- 6
difs.1[idx,]
```

One more whale in V0 than in V2. THIS NEEDS TO BE EXAMINED MORE. 2022-03-30

```{r}
FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V0") -> tmp.V0

FinalData.Both %>% 
  filter(time.steps == difs.1[idx, "time.step"]) %>% 
  filter(v == "V2") -> tmp.V2

# The second shift has one more whale (14 vs 13) in V0 
Data_Out.v2 %>% 
  filter(time.steps == difs.1[idx, "time.step"]) 
# FF is 31

FF <- 31
data <- get.data("Data/", YEAR, ff = FF)

shift <- 2
out.shift.2 <- get.shift(YEAR, data, ff=FF, i=shift)

out.shift.2$data %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarise(last.n = last(V9),
            max.n = max(V9)) -> summary.28.2.V2

data %>% 
  filter(begin >= tmp.V0[shift,"begin"] & begin <= tmp.V0[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V0


data %>% 
  filter(begin >= tmp.V2[shift,"begin"] & begin <= tmp.V2[shift, "end"]) %>%
  filter(V2 == "S") %>%
  select(V5, V9) %>%
  mutate(V5 = as.numeric(V5), V9 = as.numeric(V9)) %>%
  group_by(V5) %>%
  summarize(n = last(V9)) -> summary.V2


```


That's it for 2015. 


