---
title: "R Notebook"
output: html_notebook
---

Compares output from WinBugs runs in January and July. Should we expect to see differences in estimates when we add another year's worth of data?


```{r}
rm(list=ls())
library(tidyverse)
# results from Jan 2022
# comes from GW Rebuild for WinBUGS TE.Rmd
load("RData/GW BUGS 6yr 100k.Rdata")

Daily.Est.Jan2022 <- GW_Nmix$sims.list$Daily.Est
sp.Jan2022 <- GW_Nmix$sims.list$sp
com.Jan2022 <- GW_Nmix$sims.list$com

# Each one of them is (# samples) x (90 days) x (# years)
# To plot them using ggplot's facet, I need to convert
# these into 2D dataframes of statistics (upper and lower 
# CIs, median, etc.)
# Daily.Est.list <- sp.list <- com.list <- vector(mode = "list", 
#                                                 length = dim(Daily.Est)[3])
# 
# Daily.Est.UCIs <- Daily.Est.LCIs <- vector(mode = "list",
#                                            length = dim(Daily.Est)[3])

stats.list.Jan2022 <- vector(mode = "list",
                     length = dim(Daily.Est.Jan2022)[3])

for (k in 1:dim(Daily.Est.Jan2022)[3]){
  
  stats.list.Jan2022[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est.Jan2022[,,k], 2,
                                                         median),
                                Daily.Est.LCL = apply(Daily.Est.Jan2022[,,k], 2,
                                                      quantile,0.275),
                                Daily.Est.UCL = apply(Daily.Est.Jan2022[,,k], 2,
                                                      quantile,0.975),
                                sp.median = apply(exp(sp.Jan2022[,,k]), 2,
                                                  median),
                                sp.LCL = apply(exp(sp.Jan2022[,,k]), 2,
                                               quantile,0.025),
                                sp.UCL = apply(exp(sp.Jan2022[,,k]), 2,
                                               quantile,0.975),
                                com.median = apply(exp(com.Jan2022[,,k]), 2,
                                                   median),
                                com.LCL = apply(exp(com.Jan2022[,,k]), 2,
                                                quantile,0.025),
                                com.UCL = apply(exp(com.Jan2022[,,k]), 2,
                                                quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est.Jan2022)[2],
                                year = k)
}

all.stats.Jan2022 <- do.call("rbind", stats.list.Jan2022) %>% group_by(year)

```




```{r}
rm(list = ls()[!(ls() %in% c("all.stats.Jan2022"))])
# results from July 2022
# comes from GW Rebuild for WinBUGS TE.Rmd - ran in a different time
load("RData/GW BUGS 6yr 100k July2022.Rdata")

Daily.Est.Jul2022 <- GW_Nmix$sims.list$Daily.Est
sp.Jul2022 <- GW_Nmix$sims.list$sp
com.Jul2022 <- GW_Nmix$sims.list$com

# Each one of them is (# samples) x (90 days) x (# years)
# To plot them using ggplot's facet, I need to convert
# these into 2D dataframes of statistics (upper and lower 
# CIs, median, etc.)
# Daily.Est.list <- sp.list <- com.list <- vector(mode = "list", 
#                                                 length = dim(Daily.Est)[3])
# 
# Daily.Est.UCIs <- Daily.Est.LCIs <- vector(mode = "list",
#                                            length = dim(Daily.Est)[3])

stats.list.Jul2022 <- vector(mode = "list",
                     length = dim(Daily.Est.Jul2022)[3])

for (k in 1:dim(Daily.Est.Jul2022)[3]){
  # Daily.Est.list[[k]] <- Daily.Est[,,k]
  # Daily.Est.UCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.975)
  # Daily.Est.LCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.275)
  # 
  # sp.list[[k]] <- sp[,,k]
  # com.list[[k]] <- com[,,k]
  
  stats.list.Jul2022[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est.Jul2022[,,k], 2,
                                                         median),
                                Daily.Est.LCL = apply(Daily.Est.Jul2022[,,k], 2,
                                                      quantile,0.275),
                                Daily.Est.UCL = apply(Daily.Est.Jul2022[,,k], 2,
                                                      quantile,0.975),
                                sp.median = apply(exp(sp.Jul2022[,,k]), 2,
                                                  median),
                                sp.LCL = apply(exp(sp.Jul2022[,,k]), 2,
                                               quantile,0.025),
                                sp.UCL = apply(exp(sp.Jul2022[,,k]), 2,
                                               quantile,0.975),
                                com.median = apply(exp(com.Jul2022[,,k]), 2,
                                                   median),
                                com.LCL = apply(exp(com.Jul2022[,,k]), 2,
                                                quantile,0.025),
                                com.UCL = apply(exp(com.Jul2022[,,k]), 2,
                                                quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est.Jul2022)[2],
                                year = k)
}

all.stats.Jul2022 <- do.call("rbind", stats.list.Jul2022) %>% group_by(year)

```

They are identical... these are for six years. 

```{r}
rm(list = ls()[!(ls() %in% c("all.stats.Jan2022", "all.stats.Jul2022"))])
# results from July 2022
# comes from GW Rebuild for WinBUGS TE v2.Rmd
load("RData/GW BUGS 6yr 100ka.Rdata")

Daily.Est.Apr2022 <- GW_Nmix$sims.list$Daily.Est
sp.Apr2022 <- GW_Nmix$sims.list$sp
com.Apr2022 <- GW_Nmix$sims.list$com

# Each one of them is (# samples) x (90 days) x (# years)
# To plot them using ggplot's facet, I need to convert
# these into 2D dataframes of statistics (upper and lower 
# CIs, median, etc.)
# Daily.Est.list <- sp.list <- com.list <- vector(mode = "list", 
#                                                 length = dim(Daily.Est)[3])
# 
# Daily.Est.UCIs <- Daily.Est.LCIs <- vector(mode = "list",
#                                            length = dim(Daily.Est)[3])

stats.list.Apr2022 <- vector(mode = "list",
                     length = dim(Daily.Est.Apr2022)[3])

for (k in 1:dim(Daily.Est.Apr2022)[3]){
  # Daily.Est.list[[k]] <- Daily.Est[,,k]
  # Daily.Est.UCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.975)
  # Daily.Est.LCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.275)
  # 
  # sp.list[[k]] <- sp[,,k]
  # com.list[[k]] <- com[,,k]
  
  stats.list.Apr2022[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est.Apr2022[,,k], 2,
                                                         median),
                                Daily.Est.LCL = apply(Daily.Est.Apr2022[,,k], 2,
                                                      quantile,0.275),
                                Daily.Est.UCL = apply(Daily.Est.Apr2022[,,k], 2,
                                                      quantile,0.975),
                                sp.median = apply(exp(sp.Apr2022[,,k]), 2,
                                                  median),
                                sp.LCL = apply(exp(sp.Apr2022[,,k]), 2,
                                               quantile,0.025),
                                sp.UCL = apply(exp(sp.Apr2022[,,k]), 2,
                                               quantile,0.975),
                                com.median = apply(exp(com.Apr2022[,,k]), 2,
                                                   median),
                                com.LCL = apply(exp(com.Apr2022[,,k]), 2,
                                                quantile,0.025),
                                com.UCL = apply(exp(com.Apr2022[,,k]), 2,
                                                quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est.Apr2022)[2],
                                year = k)
}

all.stats.Apr2022 <- do.call("rbind", stats.list.Apr2022) %>% group_by(year)
```

They are all identical... 

Move to 8-yr data

```{r}
rm(list = ls()[!(ls() %in% c("all.stats.Jan2022", 
                             "all.stats.Jul2022",
                             "all.stats.Apr2022"))])

# Comes from WinBUGS Ver2.Rmd
# the RDS file is a list and BUGS_out object in the list should be the same
# structure as GW_Nmix object in the .RData objects above. 
out.8yr.v1 <- readRDS("RData/WinBUGS_8yr_v1.rds")

GW_Nmix <- out.8yr.v1$BUGS_out
Daily.Est.8yr.v1 <- GW_Nmix$sims.list$Daily.Est
sp.8yr.v1 <- GW_Nmix$sims.list$sp
com.8yr.v1 <- GW_Nmix$sims.list$com

stats.list.8yr.v1 <- vector(mode = "list",
                            length = dim(Daily.Est.8yr.v1)[3])

for (k in 1:dim(Daily.Est.8yr.v1)[3]){
  
  stats.list.8yr.v1[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est.8yr.v1[,,k], 2,
                                                         median),
                                Daily.Est.LCL = apply(Daily.Est.8yr.v1[,,k], 2,
                                                      quantile,0.275),
                                Daily.Est.UCL = apply(Daily.Est.8yr.v1[,,k], 2,
                                                      quantile,0.975),
                                sp.median = apply(exp(sp.8yr.v1[,,k]), 2,
                                                  median),
                                sp.LCL = apply(exp(sp.8yr.v1[,,k]), 2,
                                               quantile,0.025),
                                sp.UCL = apply(exp(sp.8yr.v1[,,k]), 2,
                                               quantile,0.975),
                                com.median = apply(exp(com.8yr.v1[,,k]), 2,
                                                   median),
                                com.LCL = apply(exp(com.8yr.v1[,,k]), 2,
                                                quantile,0.025),
                                com.UCL = apply(exp(com.8yr.v1[,,k]), 2,
                                                quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est.8yr.v1)[2],
                                year = k)
}

all.stats.8yr.v1 <- do.call("rbind", stats.list.8yr.v1) %>% 
  group_by(year)


# all.stats.8yr.v2.Apr2022 <- data.frame(out.8yr.v2$BUGS_out$summary) %>% 
#   rownames_to_column(var = "parameters")
# 
# Nhats <- all.stats.8yr.Apr2022[grep("Corrected.Est", all.stats.8yr.Apr2022$parameters),]

```



```{r}
rm(list = ls()[!(ls() %in% c("all.stats.Jan2022", 
                             "all.stats.Jul2022",
                             "all.stats.Apr2022",
                             "all.stats.8yr.v1"))])

# Comes from WinBUGS Ver2.Rmd
# the RDS file is a list and BUGS_out object in the list should be the same
# structure as GW_Nmix object in the .RData objects above. 
out.8yr.v2 <- readRDS("RData/WinBUGS_8yr_v2.rds")

GW_Nmix <- out.8yr.v2$BUGS_out
Daily.Est.8yr.v2 <- GW_Nmix$sims.list$Daily.Est
sp.8yr.v2 <- GW_Nmix$sims.list$sp
com.8yr.v2 <- GW_Nmix$sims.list$com

stats.list.8yr.v2 <- vector(mode = "list",
                            length = dim(Daily.Est.8yr.v2)[3])

for (k in 1:dim(Daily.Est.8yr.v2)[3]){
  
  stats.list.8yr.v2[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est.8yr.v2[,,k], 2,
                                                         median),
                                Daily.Est.LCL = apply(Daily.Est.8yr.v2[,,k], 2,
                                                      quantile,0.275),
                                Daily.Est.UCL = apply(Daily.Est.8yr.v2[,,k], 2,
                                                      quantile,0.975),
                                sp.median = apply(exp(sp.8yr.v2[,,k]), 2,
                                                  median),
                                sp.LCL = apply(exp(sp.8yr.v2[,,k]), 2,
                                               quantile,0.025),
                                sp.UCL = apply(exp(sp.8yr.v2[,,k]), 2,
                                               quantile,0.975),
                                com.median = apply(exp(com.8yr.v2[,,k]), 2,
                                                   median),
                                com.LCL = apply(exp(com.8yr.v2[,,k]), 2,
                                                quantile,0.025),
                                com.UCL = apply(exp(com.8yr.v2[,,k]), 2,
                                                quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est.8yr.v2)[2],
                                year = k)
}

all.stats.8yr.v2 <- do.call("rbind", stats.list.8yr.v2) %>% 
  group_by(year)


# all.stats.8yr.v2.Apr2022 <- data.frame(out.8yr.v2$BUGS_out$summary) %>% 
#   rownames_to_column(var = "parameters")
# 
# Nhats <- all.stats.8yr.Apr2022[grep("Corrected.Est", all.stats.8yr.Apr2022$parameters),]

```


```{r}
rm(list = ls()[!(ls() %in% c("all.stats.Jan2022", 
                             "all.stats.Jul2022",
                             "all.stats.Apr2022",
                             "all.stats.8yr.v1",
                             "all.stats.8yr.v2"))])

summary.fcn <- function(x){
  x %>% 
  group_by(year) %>%
  summarize(Nhat = sum(Daily.Est.median),
            LCL = sum(Daily.Est.LCL),
            UCL = sum(Daily.Est.UCL)) -> out.df
  return(out.df)
}

annual.Nhat.Jan2022 <- summary.fcn(all.stats.Jan2022)
annual.Nhat.Apr2022 <- summary.fcn(all.stats.Apr2022)
annual.Nhat.Jul2022 <- summary.fcn(all.stats.Jul2022)
annual.Nhat.8yr.v1 <- summary.fcn(all.stats.8yr.v1)
annual.Nhat.8yr.v2 <- summary.fcn(all.stats.8yr.v2)

```


All 6-yr analysis results in the same results, as expected. Results from 8-yr analyses are different between v1 and v2, as well as from the 6-yr analysis. I need to compare 6-yr analysis and 7-yr analysis. 

```{r}



```



