---
title: "Gray whale counts analysis for Granite Canyon"
output: html_notebook
---

This notebook contains the analysis of gray whale count data from Granite Canyon,  CA. It was first developed by John Durban, which was later rewritten by Josh Stewart. I'm just translating Josh's code into a notebook file so that I can annotate better. Furthermore, I will use OpenBUGS rather than WinBUGS in order to make this run on my Linux computer. 

Updated 2022-02-07: Parallel computing did not improve the speed, at least on the linux laptop. I should try it on my Windows computer to see if it speeds up... 

Load libraries:

```{r}
rm(list=ls())
library(R2jags)
library(abind)
library(R2OpenBUGS)
library(snow)
library(snowfall)

```

Get necessary data and initial values files:

```{r}
#Number of watch periods in each year's survey
periods <-c(136, 135, 164, 178, 179, 151)

#Watch start times, as fraction of a day
begin <- as.matrix(read.table("Data/begin.txt", 
                              header=T, 
                              nrows = max(periods)))

#watch end times
end <- as.matrix(read.table("Data/end.txt", 
                            header=T,
                            nrows = max(periods)))

#whale counts
n <- as.matrix(read.table("Data/n.txt",
                          header=T,
                          nrows = max(periods)))

dim(n) <- c(179,2,6) #convert this back to a 3D array
#n <- abind(n,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the sightings array (this is for the day 1 and day 90 zero-whale anchor points)


n1 <- as.matrix(read.table("Data/n1.txt",
                           header=T,
                          nrows = max(periods)))      #These aren't needed; used in the WinBUGS GUI version previously. Here we can just re-use n
dim(n1) <- c(179,2,6) #convert this back to a 3D array


n2 <- as.matrix(read.table("Data/n2.txt",
                           header=T,
                          nrows = max(periods)))      #These aren't needed
dim(n2) <- c(179,2,6) #convert this back to a 3D array


u <- as.matrix(read.table("Data/u.txt",
                          header=T,
                          nrows = max(periods)))
dim(u) <- c(179,2,6) #convert this back to a 3D array
#u <- abind(u,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the effort on/off array
#for(i in 1:length(periods)){ #Place 1's for 'effort on' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
#  u[(periods[i]+1):(periods[i]+2),,i] <- 1
#}


#visibility
vs <- as.matrix(read.table("Data/vs.txt",
                           header=T,
                          nrows = max(periods)))
#vs <- rbind(vs,matrix(NA,nrow=2,ncol=6)) #Add two trailing NAs (this is for the day 1 and day 90 zero-whale anchor points)

#beaufort
bf <- as.matrix(read.table("Data/bf.txt",
                           header=T,
                          nrows = max(periods)))
#bf <- rbind(bf,matrix(NA,nrow=2,ncol=6)) #Add two trailing NAs (this is for the day 1 and day 90 zero-whale anchor points)

#observer numbers
obs <- as.matrix(read.table("Data/obs.txt",
                            header=T,
                          nrows = max(periods)))
dim(obs) <- c(179,2,6) #convert this back to a 3D array
#obs <- abind(obs,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the effort on/off array
#for(i in 1:length(periods)){ #Place 36s for 'no observer' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
#  obs[(periods[i]+1):(periods[i]+2),,i] <- 36 #this will force it to the mean observation probability with no observer effect
#}

N_inits <- as.matrix(read.table("Data/Initial values/N_inits.txt",header=T))

for(i in 1:length(periods)){
  N_inits[(periods[i]+1):(periods[i]+2),i] <- NA #we're going to make N a partially observed data object with anchor points at day 1 and 90
}

#The 'data' has to be the inverse of the inits, 
# with NAs for all of the estimated Ns, and 0s for the days 1 and 90
N <- matrix(NA,nrow=181,ncol=6) 

for(i in 1:length(periods)){
  N[(periods[i]+1):(periods[i]+2),i] <- 0 #True number of whales passing fixed at 0 for day 1 and 90
}

```

Create data lists:

```{r}

#specify survey days associated with each watch period
day <- floor(begin)
t <- round((begin+end)/2)
#Add a couple of extra rows of NAs to the end of the day index reference to match up with the fixed 0s in N (above), assigning them to days 1 and 90
day <- rbind(as.matrix(day),matrix(NA,nrow=2,ncol=6))
for(i in 1:length(periods)){ #Set the anchor points: days 1 and 90
  day[(periods[i]+1):(periods[i]+2),i] <- c(1,90)
}

t <- rbind(as.matrix(t),matrix(NA,nrow=2,ncol=6))
for(i in 1:length(periods)){ #Set the anchor points: days 1 and 90
  t[(periods[i]+1):(periods[i]+2),i] <- c(1,90)
}

#### Two separate models each fit to data, 
# and then a third replicate of data to select
# 'best fit' (same structure as Durban et al)

# Shorten data to first 4 years only to replicate 
# the analysis in Durban et al 2016:
n.short <- n[,,1:4]
obs.short <- obs[,,1:4]
periods.short <- periods[1:4]
u.short <- u[,,1:4]
vs.short <- vs[,1:4]
bf.short <- bf[,1:4]
day.short <-day[,1:4]
t.short <- t[,1:4]
N.short <- N[,1:4]
N_inits.short <- N_inits[,1:4]

Watch.Length <- rbind(end,matrix(NA,nrow=2,ncol=6)) - rbind(begin,matrix(NA,nrow=2,ncol=6))
for(i in 1:length(periods)){ #Place 36s for 'no observer' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
  Watch.Length[(periods[i]+1):(periods[i]+2),i] <- 1 #this will force it to the mean observation probability with no observer effect
}
Watch.Length.short <- Watch.Length[,1:4]

# jags.data.short <- list(n=n.short,
#                         n.com=n.short,
#                         n.sp=n.short,
#                         n.station = dim(n.short)[2],
#                         n.year = dim(n.short)[3],
#                         n.obs = max(obs.short),
#                         periods = periods.short,
#                         obs=obs.short,
#                         #Watch.Length = 0.0625,
#                         u=u.short,
#                         vs=vs.short,
#                         bf=bf.short,
#                         day=t.short,
#                         #day=day.short,
#                         N=N.short,
#                         N.com=N.short,
#                         N.sp=N.short,
#                         knot=c(-1.46,-1.26,-1.02,-0.78,
#                                -0.58,-0.34,-0.10,0.10,
#                                0.34,0.57,0.78,1.02,1.26,1.46),
#                         n.knots=14,
#                         Watch.Length=Watch.Length.short)
# 
# jags.inits.short <- function() list(mean.prob = 0.5,
#                                     BF.Fixed = 0,
#                                     VS.Fixed = 0,
#                                     mean.prob.sp = 0.5,
#                                     BF.Fixed.sp = 0,
#                                     VS.Fixed.sp = 0,
#                                     mean.prob.com = 0.5,
#                                     BF.Fixed.com = 0,
#                                     VS.Fixed.com = 0,
#                                     mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
#                                     beta.sigma = c(0.5,0.5,0.5),#beta.sigma = c(7,7,7),
#                                     BF.Switch = 1,
#                                     VS.Switch = 1,
#                                     OBS.Switch = 1,
#                                     sigma.Obs = 1,
#                                     BF.Switch.sp = 1,
#                                     VS.Switch.sp = 1,
#                                     OBS.Switch.sp = 1,
#                                     sigma.Obs.sp = 1,
#                                     BF.Switch.com = 1,
#                                     VS.Switch.com = 1,
#                                     OBS.Switch.com = 1,
#                                     sigma.Obs.com = 1,
#                                     N = N_inits.short,
#                                     N.com = N_inits.short,
#                                     N.sp = N_inits.short,
#                                     #z = matrix(1,nrow=90,ncol=6),
#                                     beta.sp = array(data=0,dim=c(2,4)),
#                                     sd.b.sp = c(1,1,1,1),
#                                     z=matrix(1,nrow=90,ncol=4))

jags.data <- list(n=n,
                  n.com=n,
                  n.sp=n,
                  n.station = dim(n)[2],
                  n.year = dim(n)[3],
                  n.obs = max(obs),
                  periods = periods,
                  obs=obs,
                  #Watch.Length = 0.0625,
                  u=u,
                  vs=vs,
                  bf=bf,
                  #day=day,
                  day=t,
                  N=N,
                  N.com=N,
                  N.sp=N,
                  knot=c(-1.46,-1.26,-1.02,-0.78,
                         -0.58,-0.34,-0.10,0.10,
                         0.34,0.57,0.78,1.02,1.26,1.46),
                  n.knots=14,
                  #begin=begin,
                  #end=end,
                  Watch.Length=Watch.Length)

jags.inits <- function() list(mean.prob = 0.5,
                              BF.Fixed = 0,
                              VS.Fixed = 0,
                              mean.prob.sp = 0.5,
                              BF.Fixed.sp = 0,
                              VS.Fixed.sp = 0,
                              mean.prob.com = 0.5,
                              BF.Fixed.com = 0,
                              VS.Fixed.com = 0,
                              mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
                              beta.sigma = c(1,1,1),#beta.sigma = c(7,7,7),
                              BF.Switch = 1,
                              VS.Switch = 1,
                              OBS.Switch = 1,
                              sigma.Obs = 1,
                              BF.Switch.sp = 1,
                              VS.Switch.sp = 1,
                              OBS.Switch.sp = 1,
                              sigma.Obs.sp = 1,
                              BF.Switch.com = 1,
                              VS.Switch.com = 1,
                              OBS.Switch.com = 1,
                              sigma.Obs.com = 1,
                              N = N_inits,
                              N.com = N_inits,
                              N.sp = N_inits,
                              #z = matrix(1,nrow=90,ncol=6),
                              beta.sp = array(data=0,dim=c(2,6)),
                              sd.b.sp = c(1,1,1,1,1,1),
                              z=matrix(1,nrow=90,ncol=6))

```


Then run BUGS:

```{r}
#### To run 2006-2019 data, load .RDS object and name it jags.data, 
# then update intitial values:
# Files are 2006-2019_GC_Formatted_Data and 2006-2019_GC_N_inits

# WinBUGS gives errors when N inits are set to 0. 
# Try setting them to 1 instead (seems to work):
N_inits[which(N_inits==0,arr.ind = T)] <- 1

parameters <- c("lambda","OBS.RF","OBS.Switch",
                "BF.Switch","BF.Fixed","VS.Switch",
                "VS.Fixed","mean.prob","mean.prob.com",
                "mean.prob.sp","BF.Fixed.com",
                "BF.Fixed.sp","VS.Fixed.com",
                "VS.Fixed.sp",
                "Corrected.Est","Raw.Est","z",
                "com","sp","Daily.Est","mean.beta",
                "beta.sigma","beta","beta.sp","b.sp","sd.b.sp")

# ni <- 100000
# nt <- 80
# nb <- 60000
# nc <- 3

n.iter <- 100000 #500 #
n.thin <- 80 #2 #
n.burn <- 60000 #200 #
n.chain <- 3

MCMC.params <- list(n.iter = n.iter,
                    n.thin = n.thin,
                    n.burn = n.burn,
                    n.chain = n.chain)

# make it parallel using snow and snowfall
# Modified from https://oliviergimenez.github.io/blog/run_openbugs_parallel/

sfInit(parallel=TRUE, cpus=n.chain)
sfLibrary(R2OpenBUGS)

# creating separate directory for each CPU process
folders <- list()
for (k in 1:n.chain){
  folders[[k]] <- paste0(getwd(), "/chain", k)
  if (!dir.exists(folders[[k]])) dir.create(folders[[k]])
}

# sinking the model into a file in each directory
for (folder in folders){
  if (!file.exists(paste0(folder, "/GW_Nmix_Orig.txt")))
    file.copy(from = "models/GW_Nmix_Orig.txt",
              to = paste0(folder, "/GW_Nmix_Orig.txt"))
}

# Change the number of rows/columns of z and beta.sp with new data. 
parallel.bugs <- function(chain, x.data, N_inits, params, MCMC.params){
  # a. defining directory for each CPU
  sub.folder <- paste(getwd(),"/chain", chain, sep="")
 
  # b. specifying the initial MCMC values
  inits <- function() list(mean.prob = 0.5,
                              BF.Fixed = 0,
                              VS.Fixed = 0,
                              mean.prob.sp = 0.5,
                              BF.Fixed.sp = 0,
                              VS.Fixed.sp = 0,
                              mean.prob.com = 0.5,
                              BF.Fixed.com = 0,
                              VS.Fixed.com = 0,
                              mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
                              beta.sigma = c(1,1,1),#beta.sigma = c(7,7,7),
                              BF.Switch = 1,
                              VS.Switch = 1,
                              OBS.Switch = 1,
                              sigma.Obs = 1,
                              BF.Switch.sp = 1,
                              VS.Switch.sp = 1,
                              OBS.Switch.sp = 1,
                              sigma.Obs.sp = 1,
                              BF.Switch.com = 1,
                              VS.Switch.com = 1,
                              OBS.Switch.com = 1,
                              sigma.Obs.com = 1,
                              N = N_inits,
                              N.com = N_inits,
                              N.sp = N_inits,
                              beta.sp = array(data=0,dim=c(2,6)),
                              sd.b.sp = c(1,1,1,1,1,1),
                              z=matrix(1,nrow=90,ncol=6))
  
  # c. calling OpenBugs
  bugs(data = x.data,
       inits = inits,
       parameters.to.save = params,
       model.file = "GW_Nmix_Orig.txt",
       working.directory = sub.folder,
       n.chains = MCMC.params$n.chain,
       n.iter = MCMC.params$n.iter, 
       n.burnin = MCMC.params$n.burn, 
       n.thin = MCMC.params$n.thin)
  
}
 
Start_Time<-Sys.time()

sfLapply(1:n.chain, 
         fun=parallel.bugs, 
         x.data=jags.data, 
         N_inits = N_inits,
         params=parameters, 
         MCMC.params= MCMC.params)

#Run time: 
Run_Time <- Sys.time() - Start_Time

# locating position of each CODA chain - each folder has all chains.
chain1 <- paste(folders[[1]], "/CODAchain1.txt", sep="")
chain2 <- paste(folders[[1]], "/CODAchain2.txt", sep="")
chain3 <- paste(folders[[1]], "/CODAchain3.txt", sep="")
 
# and, finally, getting the results
res <- read.bugs(c(chain1, chain2, chain3), quiet = T)
sfStop()

save.image("RData/GW BUGS 7yr 100k parallel.RData")

```



Make some plots:

```{r}
par(mfrow=c(2,2))
year <- 4

#John's plots:

UCIs <- apply(Daily.Est[,,year],2,quantile,0.975)
LCIs <- apply(Daily.Est[,,year],2,quantile,0.025)
plot(apply(exp(sp[,,year]),2,median), 
     type='l',
     ylim=c(0,max(UCIs)+50), 
     xlab="Days since 1 December", 
     ylab = "Whales per day")
lines(apply(exp(com[,,year]),2,median),type='l',lty=2)
segments(x0=1:90,y0=LCIs,y1=UCIs)

#Com vs Sp
plot(apply(exp(sp[,,year]),2,quantile,0.975),type='l',lty=2)
lines(apply(exp(sp[,,year]),2,median),type='l')
lines(apply(exp(sp[,,year]),2,quantile,0.025),type='l',lty=2)

plot(apply(exp(com[,,year]),2,quantile,0.975),type='l',lty=2)
lines(apply(exp(com[,,year]),2,median),type='l')
lines(apply(exp(com[,,year]),2,quantile,0.025),type='l',lty=2)

plot(apply(Daily.Est[,,year],2,quantile,0.975),type='l',lty=2)
lines(apply(Daily.Est[,,year],2,median),type='l')
lines(apply(Daily.Est[,,year],2,quantile,0.025),type='l',lty=2)

median(apply(exp(sp[,,4]),1,sum))

# PARAMETER NAME TRANSFERS:

# JOSH     JOHN
#    Spline 
# b.sp     b1 
# beta.sp  beta1
# X.sp     X1


```

