---
title: "R Notebook"
output: html_notebook
---

This is a notebook version of GW Rebuild for WinBUGS and using the data in "RData/2006-2019_GC_Formatted_Data.RDS". I did not get raw data for these years, except 2015, 2016, 2019, and 2022. There is a need to examine the data because I found some errors in Josh's data extraction code.

The stored data does not include years... 

Set up libraries and get data:

```{r}
rm(list=ls())
library(R2jags)
library(abind)
library(R2WinBUGS)
library(tidyverse)
# PARAMETER NAME TRANSFERS:

# JOSH     JOHN
#    Spline 
# b.sp     b1 
# beta.sp  beta1
# X.sp     X1

dat <- readRDS("RData/2006-2019_GC_Formatted_Data.RDS")

# the number of years to be included in the analysis.
# maximum is 7. the dat object contains 7 years worth of data.
x <- 7
```

Get data:

```{r}
#Number of watch periods in each year's survey
periods <-dat$periods[1:x]

#Watch start times, as fraction of a day
begin <- as.matrix(read.table("Data/begin.txt", 
                              header=T, 
                              nrows = max(periods)))

#watch end times
end <- as.matrix(read.table("Data/end.txt", 
                            header=T,
                            nrows = max(periods)))

#whale counts
n <- as.matrix(read.table("Data/n.txt",
                          header=T,
                          nrows = max(periods)))

dim(n) <- c(179,2,x) #convert this back to a 3D array
#n <- abind(n,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the sightings array (this is for the day 1 and day 90 zero-whale anchor points)


n1 <- as.matrix(read.table("Data/n1.txt",
                           header=T,
                          nrows = max(periods)))      #These aren't needed; used in the WinBUGS GUI version previously. Here we can just re-use n
dim(n1) <- c(179,2,x) #convert this back to a 3D array


n2 <- as.matrix(read.table("Data/n2.txt",
                           header=T,
                          nrows = max(periods)))      #These aren't needed
dim(n2) <- c(179,2,x) #convert this back to a 3D array


u <- as.matrix(read.table("Data/u.txt",
                          header=T,
                          nrows = max(periods)))
dim(u) <- c(179,2,x) #convert this back to a 3D array
#u <- abind(u,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the effort on/off array
#for(i in 1:length(periods)){ #Place 1's for 'effort on' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
#  u[(periods[i]+1):(periods[i]+2),,i] <- 1
#}


#visibility
vs <- as.matrix(read.table("Data/vs.txt",
                           header=T,
                          nrows = max(periods)))
#vs <- rbind(vs,matrix(NA,nrow=2,ncol=6)) #Add two trailing NAs (this is for the day 1 and day 90 zero-whale anchor points)

#beaufort
bf <- as.matrix(read.table("Data/bf.txt",
                           header=T,
                          nrows = max(periods)))
#bf <- rbind(bf,matrix(NA,nrow=2,ncol=6)) #Add two trailing NAs (this is for the day 1 and day 90 zero-whale anchor points)

#observer numbers
obs <- as.matrix(read.table("Data/obs.txt",
                            header=T,
                          nrows = max(periods)))
dim(obs) <- c(179,2,x) #convert this back to a 3D array
#obs <- abind(obs,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the effort on/off array
#for(i in 1:length(periods)){ #Place 36s for 'no observer' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
#  obs[(periods[i]+1):(periods[i]+2),,i] <- 36 #this will force it to the mean observation probability with no observer effect
#}

N_inits <- as.matrix(read.table("Data/Initial Values/N_inits.txt",
                                header=T))

for(i in 1:length(periods)){
  N_inits[(periods[i]+1):(periods[i]+2),i] <- NA #we're going to make N a partially observed data object with anchor points at day 1 and 90
}

#The 'data' has to be the inverse of the inits, 
# with NAs for all of the estimated Ns, and 0s for the days 1 and 90
N <- matrix(NA,nrow=181,ncol=x) 

for(i in 1:length(periods)){
  N[(periods[i]+1):(periods[i]+2),i] <- 0 #True number of whales passing fixed at 0 for day 1 and 90
}
```

Set up BUGS:

```{r}

#specify survey days associated with each watch period
day <- floor(begin)
t <- round((begin+end)/2)
#Add a couple of extra rows of NAs to the end of the day index reference to match up with the fixed 0s in N (above), assigning them to days 1 and 90
day <- rbind(as.matrix(day),matrix(NA,nrow=2,ncol=x))
for(i in 1:length(periods)){ #Set the anchor points: days 1 and 90
  day[(periods[i]+1):(periods[i]+2),i] <- c(1,90)
}

t <- rbind(as.matrix(t),matrix(NA,nrow=2,ncol=x))
for(i in 1:length(periods)){ #Set the anchor points: days 1 and 90
  t[(periods[i]+1):(periods[i]+2),i] <- c(1,90)
}

#### Two separate models each fit to data, 
# and then a third replicate of data to select
# 'best fit' (same structure as Durban et al)

# Shorten data to first 4 years only to replicate 
# the analysis in Durban et al 2016:
n.short <- n[,,1:4]
obs.short <- obs[,,1:4]
periods.short <- periods[1:4]
u.short <- u[,,1:4]
vs.short <- vs[,1:4]
bf.short <- bf[,1:4]
day.short <-day[,1:4]
t.short <- t[,1:4]
N.short <- N[,1:4]
N_inits.short <- N_inits[,1:4]

Watch.Length <- rbind(end,matrix(NA,nrow=2,ncol=x)) - rbind(begin,matrix(NA,nrow=2,ncol=x))
for(i in 1:length(periods)){ #Place 36s for 'no observer' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
  Watch.Length[(periods[i]+1):(periods[i]+2),i] <- 1 #this will force it to the mean observation probability with no observer effect
}

jags.data <- list(n=n,
                  n.com=n,
                  n.sp=n,
                  n.station = dim(n)[2],
                  n.year = dim(n)[3],
                  n.obs = max(obs),
                  periods = periods,
                  obs=obs,
                  #Watch.Length = 0.0625,
                  u=u,
                  vs=vs,
                  bf=bf,
                  #day=day,
                  day=t,
                  N=N,
                  N.com=N,
                  N.sp=N,
                  knot=c(-1.46,-1.26,-1.02,-0.78,
                         -0.58,-0.34,-0.10,0.10,
                         0.34,0.57,0.78,1.02,1.26,1.46),
                  n.knots=14,
                  #begin=begin,
                  #end=end,
                  Watch.Length=Watch.Length)

jags.inits <- function() list(mean.prob = 0.5,
                              BF.Fixed = 0,
                              VS.Fixed = 0,
                              mean.prob.sp = 0.5,
                              BF.Fixed.sp = 0,
                              VS.Fixed.sp = 0,
                              mean.prob.com = 0.5,
                              BF.Fixed.com = 0,
                              VS.Fixed.com = 0,
                              mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
                              beta.sigma = c(1,1,1),#beta.sigma = c(7,7,7),
                              BF.Switch = 1,
                              VS.Switch = 1,
                              OBS.Switch = 1,
                              sigma.Obs = 1,
                              BF.Switch.sp = 1,
                              VS.Switch.sp = 1,
                              OBS.Switch.sp = 1,
                              sigma.Obs.sp = 1,
                              BF.Switch.com = 1,
                              VS.Switch.com = 1,
                              OBS.Switch.com = 1,
                              sigma.Obs.com = 1,
                              N = N_inits,
                              N.com = N_inits,
                              N.sp = N_inits,
                              #z = matrix(1,nrow=90,ncol=6),
                              beta.sp = array(data=0,dim=c(2,x)),
                              sd.b.sp = c(1,1,1,1,1,1),
                              z=matrix(1,nrow=90,ncol=x))

out.inits <- list(fun = jags.inits,
                  N_inits = N_inits)
#### To run 2006-2019 data, load .RDS object and name it jags.data, 
# then update intitial values:
# Files are 2006-2019_GC_Formatted_Data and 2006-2019_GC_N_inits

# WinBUGS gives errors when N inits are set to 0. 
# Try setting them to 1 instead (seems to work):
N_inits[which(N_inits==0,arr.ind = T)] <- 1

parameters <- c("lambda","OBS.RF","OBS.Switch",
                "BF.Switch","BF.Fixed","VS.Switch",
                "VS.Fixed","mean.prob","mean.prob.com",
                "mean.prob.sp","BF.Fixed.com",
                "BF.Fixed.sp","VS.Fixed.com",
                "VS.Fixed.sp",
                "Corrected.Est","Raw.Est","z",
                "com","sp","Daily.Est","mean.beta",
                "beta.sigma","beta","beta.sp","b.sp","sd.b.sp")

ni <- 100000
nt <- 80
nb <- 60000
nc <- 3

# ni <- 100
# nt <- 2
# nb <- 60
# nc <- 3
```

Run WinBUGS if it hasn't been run..

```{r}
if (!file.exists(paste0("RData/GW BUGS ", x, "yr 100ka.RData"))){
  
  #Run time: 
  Start_Time<-Sys.time()
  
  GW_Nmix <- bugs(data = jags.data,
                  inits = jags.inits,
                  parameters = parameters,
                  model.file="GW_Nmix_Orig.bugs",
                  n.chains = nc,
                  n.iter = ni, 
                  n.burnin = nb, 
                  n.thin = nt,
                  debug=F,
                  bugs.directory = "C:/Users/tomo.eguchi/Documents/WinBUGS14")
  
  Run_Time <- Sys.time() - Start_Time
  save.image(paste0("RData/GW BUGS ", x, "yr 100ka.RData"))
  
} else {
  load(paste0("RData/GW BUGS ", x, "yr 100ka.RData"))
}
```


Make some plots:

```{r}
# Extract estimated counts
Daily.Est <- GW_Nmix$sims.list$Daily.Est
sp <- GW_Nmix$sims.list$sp
com <- GW_Nmix$sims.list$com

# Each one of them is (# samples) x (90 days) x (# years)
# To plot them using ggplot's facet, I need to convert
# these into 2D dataframes of statistics (upper and lower 
# CIs, median, etc.)
# Daily.Est.list <- sp.list <- com.list <- vector(mode = "list", 
#                                                 length = dim(Daily.Est)[3])
# 
# Daily.Est.UCIs <- Daily.Est.LCIs <- vector(mode = "list",
#                                            length = dim(Daily.Est)[3])

stats.list <- vector(mode = "list",
                     length = dim(Daily.Est)[3])

for (k in 1:dim(Daily.Est)[3]){
  # Daily.Est.list[[k]] <- Daily.Est[,,k]
  # Daily.Est.UCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.975)
  # Daily.Est.LCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.275)
  # 
  # sp.list[[k]] <- sp[,,k]
  # com.list[[k]] <- com[,,k]
  
  stats.list[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est[,,k], 2,
                                                         median),
                                Daily.Est.LCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.275),
                                Daily.Est.UCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.975),
                                sp.median = apply(exp(sp[,,k]), 2,
                                                  median),
                                sp.LCL = apply(exp(sp[,,k]), 2,
                                               quantile,0.025),
                                sp.UCL = apply(exp(sp[,,k]), 2,
                                               quantile,0.975),
                                com.median = apply(exp(com[,,k]), 2,
                                                   median),
                                com.LCL = apply(exp(com[,,k]), 2,
                                                quantile,0.025),
                                com.UCL = apply(exp(com[,,k]), 2,
                                                quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est)[2],
                                year = k)
}

all.stats <- do.call("rbind", stats.list) %>% group_by(year)

ggplot(data = all.stats) + 
  geom_line(aes(x = days, y = sp.median)) + 
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", 
              alpha = 0.5) +
  geom_line(aes(x = days, y = com.median),
            color = "red") +
  facet_wrap(vars(year)) +
  xlab("Days since December 1") + 
  ylab("Whales per day")

```



```{r}

#Com vs Sp
ggplot(data = all.stats) +
  geom_line(aes(x = days, y = sp.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day")



```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = com.LCL, 
                  ymax = com.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day")

```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = Daily.Est.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = Daily.Est.LCL, 
                  ymax = Daily.Est.UCL),
              fill = "orange", 
              alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day")


```

Total abundance
```{r}
abundance.list <- vector(mode = "list", length = dim(sp)[3])
for (y in 1:dim(sp)[3]){
  abundance.list[[y]] <- data.frame(median = median(apply(exp(sp[,,y]), 1, sum)),
                                    LCL = quantile(apply(exp(sp[,,y]), 1, sum), 0.025),
                                    UCL = quantile(apply(exp(sp[,,y]), 1, sum), 0.975),
                                    year = y)
  
}

abundance.df <- do.call("rbind", abundance.list)

ggplot(data = abundance.df) + 
  geom_line(aes(x = year, y = median)) + 
  geom_errorbar(aes(x = year, ymin = LCL, ymax = UCL))

#median(apply(exp(sp[,,4]),1,sum))

```

