---
title: "R Notebook"
output: html_notebook
---

This notebook contains an effort to impute unobserved gray whale counts using the discrete Fourier series (DFS). This approach has been tried for imputation of the numbers of nesting leatherback turtles and green turtle hatchlings before (although unpublished as of Aug 2022).


Set up libraries

```{r}
rm(list=ls())
library(abind)
library(jagsUI)
library(tidyverse)
library(bayesplot)
library(mgcv)


# # periods for the 6 survey years
periods <-c(136, 135, 164, 178, 179, 151)

# Observer list gets updated as new observers are added. 
# obs.list <- read.csv("Data/Observer list.csv", header = T) 
# colnames(obs.list) <- c("obs", "ID")
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

n.seasons <- length(seasons)

MCMC.params <- list(n.samples = 100000,
                   n.thin = 80,
                   n.burnin = 60000,
                   n.chains = 5)

```

The daily mean is a smooth function of days (1:90)

```{r setup-data}
#Ver2.results <- readRDS(paste0("RData/WinBUGS_", x, "yr_v2.rds"))
Ver2.results <- readRDS(paste0("RData/WinBUGS_", 
                               n.seasons, "yr_100k_v2.rds"))

daily.effort <- daily.counts <- daily.N <- u <- matrix(nrow = 90, ncol = n.seasons)
bf <- vs <- obs <- matrix(nrow = 90, ncol = n.seasons)
Watch.Length <- Ver2.results$jags.data$Watch.Length
day.mat <- Ver2.results$jags.data$day
ns <- rbind(Ver2.results$jags.data$n[,1,], matrix(NA, nrow = 2, ncol = n.seasons))
BF <-  rbind(Ver2.results$jags.data$bf, matrix(NA, nrow = 2, ncol = n.seasons))
VS <-  rbind(Ver2.results$jags.data$vs, matrix(NA, nrow = 2, ncol = n.seasons))
OBS <- rbind(Ver2.results$jags.data$obs[,1,], matrix(NA, nrow = 2, ncol = n.seasons))

for (y in 1:n.seasons){
  for (j in 1:90){
    if (j == 1 | j == 90){
      daily.effort[j,y] <- 0
      daily.counts[j,y] <- 0
      daily.N[j,y] <- 0
      u[j,y] <- 0
      bf[j,y] <- vs[j,y] <- 99
      obs[j,y] <- max(OBS, na.rm = T) + 1 
    } else {
      daily.effort[j, y] <- sum(Watch.Length[day.mat[,y] == j, y], na.rm = T)
      daily.counts[j, y] <- sum(ns[day.mat[,y] == j, y], na.rm = T)
      daily.N[j, y] <- ceiling(daily.counts[j,y]/daily.effort[j,y])
      
      u[j,y] <- ifelse(sum(Watch.Length[day.mat[,y] == j, y], na.rm = T) > 0, 1, 0)
      bf.tmp <- BF[day.mat[,y] == j, y]
      if (sum(!is.na(bf.tmp)) > 0){
        bf[j,y] <- max(BF[day.mat[,y] == j, y], na.rm = T)
        vs[j,y] <- max(VS[day.mat[,y] == j, y], na.rm = T)
      } else {
        bf[j,y] <- 99
        vs[j,y] <- 99
      }
      
      obs[j,y] <- first(OBS[day.mat[,y] == j, y])
    }
  }
}
  
daily.N[is.na(daily.N)] <- 0
bf[daily.effort == 0] <- 99
vs[daily.effort == 0] <- 99
obs[daily.effort == 0] <- max(OBS, na.rm = T) + 1

# daily.N.df <- data.frame(daily.N) 
# colnames(daily.N.df) <- seasons
# daily.N.df %>% 
#   pivot_longer(everything(),
#                names_to = "Season", 
#                values_to = "N") %>%  
#   arrange(Season) -> daily.N.df.long
# 
# daily.N.df.long$days <- rep(1:90, times = n.seasons)
# daily.N.df.long$year <- factor(rep(1:n.seasons, each = 90))
# 
# daily.N.df.long %>% 
#   select(-Season) %>%
#   na.omit() -> daily.N.df.long.noNA
# 
# # To use non-Gaussian family, I add 1 to all counts:
# daily.N.df.long.noNA$Nplus1 <- daily.N.df.long.noNA$N + 1  

jags.data <- list(y = log(daily.N+0.01),
                  u = u,
                  bf = bf,
                  vs = vs,
                  obs = obs,
                  period = dim(daily.N)[1],
                  n.steps = dim(daily.N)[1],
                  n.years = dim(daily.N)[2],
                  n.obs = max(obs, na.rm = T),
                  pi = pi)

jags.params <- c("beta.cos", "beta.sin",
                 "sigma.cos", "sigma.sin",
                 "sigma.y", "mean.prob",
                 "BF.fixed", "VS.fixed",
                 "OBS.RF", "Raw.Est",
                 "Corrected.Est",
                 "y", "N", "deviance")
```



```{r run-jags}

jags.model <- paste0("models/model_DFS_beta_yr_bino.txt")

jm <- jags(jags.data,
             inits = NULL,
             parameters.to.save= jags.params,
             model.file = jags.model,
             n.chains = MCMC.params$n.chains,
             n.burnin = MCMC.params$n.burnin,
             n.thin = MCMC.params$n.thin,
             n.iter = MCMC.params$n.samples,
             DIC = T, parallel=T)

mcmc_trace(jm$samples, c("Corrected.Est[1]", "Corrected.Est[2]",
                         "Corrected.Est[3]", "Corrected.Est[4]",
                         "Corrected.Est[5]", "Corrected.Est[6]",
                         "Corrected.Est[7]", "Corrected.Est[8]"))


mcmc_dens(jm$samples, c("Corrected.Est[1]", "Corrected.Est[2]",
                         "Corrected.Est[3]", "Corrected.Est[4]",
                         "Corrected.Est[5]", "Corrected.Est[6]",
                         "Corrected.Est[7]", "Corrected.Est[8]"))
  
```


```{r}
N.mean <- exp(jm$mean$N) %>%
  as_tibble()

df_colnames <- tibble(
  "new_name" = seasons,
  "old_name" = colnames(N.mean)
)

var_names <- deframe(df_colnames)
N.mean %>% 
  rename(!!!var_names) %>%
  mutate(Day = 1:90) -> N.mean
 
N.mean %>% 
  select(-Day) %>%
  pivot_longer(cols = everything(),
               names_to = "Season",
               values_to = "N") %>%
  arrange(Season) %>%
  mutate(Day = rep(1:90, length(seasons))) -> N.mean.long
  

Corrected.Est.mean <- jm$mean$Corrected.Est

ggplot(N.mean.long) +
  geom_point(aes(x = Day, y = N, color = Season))
  #facet_grid(rows = vars(Season))


```

Looks nice but probably not very useful in estimating the total abundance with these symmetrical curves.

