---
title: "R Notebook"
output: html_notebook
---

This notebook includes the part that a GAM is fitted to the observed daily counts of gray whales. In this version (Ver3), I use jagam function in mgcv to create the input data but they are used to run Model_DailyN_GAM_kx_Nmix_JAGS.txt. 

THIS IS STILL INCOMPLETE. I DON'T KNOW HOW TO PREDICT COUNTS ONCE BETAS ARE ESTIMATED, WITHOUT USING THE PREDICT.GAM FUNCTION IN MGCV. I WANT TO DO THAT IN THE JAGS CODE DIRECTLY USING THE POSTERIORS OF BETAS. I PROBABLY HAVE TO READ SIMON WOOD'S BOOK TO FIGURE OUT THE NUTS AND BOLTS... 2022-07-15

Set up libraries

```{r}
rm(list=ls())
library(abind)
library(jagsUI)
library(tidyverse)
library(bayesplot)
library(mgcv)


# # periods for the 6 survey years
periods <-c(136, 135, 164, 178, 179, 151)

# Observer list gets updated as new observers are added. 
# obs.list <- read.csv("Data/Observer list.csv", header = T) 
# colnames(obs.list) <- c("obs", "ID")
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

n.seasons <- length(seasons)

# CHANGE THIS LINE BACK
#Ver2.results <- readRDS(paste0("RData/WinBUGS_", x, "yr_v2.rds"))
Ver2.results <- readRDS(paste0("RData/WinBUGS_", 
                               n.seasons, "yr_100k_v2.rds"))

#jags.data <- Ver2.results$jags.data
jags.params <- c("lambda","OBS.RF","OBS.Switch",
                "BF.Switch","BF.Fixed","VS.Switch",
                "VS.Fixed","mean.prob","mean.prob.com",
                "mean.prob.sp","BF.Fixed.com",
                "BF.Fixed.sp","VS.Fixed.com",
                "VS.Fixed.sp",
                "Corrected.Est","Raw.Est","z",
                "com","sp","Daily.Est","mean.beta",
                "beta.sigma","beta","beta.sp","b.sp","sd.b.sp")

MCMC.params <- list(n.samples = 100000,
                   n.thin = 80,
                   n.burnin = 60000,
                   n.chains = 5)

```

The daily mean is a smooth function of days (1:90)

```{r setup-data}
daily.effort <- daily.counts <- daily.N <- matrix(nrow = 90, ncol = n.seasons)
Watch.Length <- Ver2.results$jags.data$Watch.Length
day.mat <- Ver2.results$jags.data$day
ns <- rbind(Ver2.results$jags.data$n[,1,], matrix(NA, nrow = 2, ncol = n.seasons))

for (y in 1:n.seasons){
  for (j in 1:90){
    if (j == 1 | j == 90){
      daily.effort[j,y] <- 0
      daily.counts[j,y] <- 0
      daily.N[j,y] <- 0
    } else {
      daily.effort[j, y] <- sum(Watch.Length[day.mat[,y] == j, y], na.rm = T)
      daily.counts[j, y] <- sum(ns[day.mat[,y] == j, y], na.rm = T)
      daily.N[j, y] <- ceiling(daily.counts[j,y]/daily.effort[j,y])
      
    }
  }
}
  
daily.N.df <- data.frame(daily.N) 
colnames(daily.N.df) <- seasons
daily.N.df %>% 
  pivot_longer(everything(),
               names_to = "Season", 
               values_to = "N") %>%  
  arrange(Season) -> daily.N.df.long

daily.N.df.long$days <- rep(1:90, times = n.seasons)
daily.N.df.long$year <- factor(rep(1:n.seasons, each = 90))

daily.N.df.long %>% 
  select(-Season) %>%
  na.omit() -> daily.N.df.long.noNA

# To use non-Gaussian family, I add 1 to all counts:
daily.N.df.long.noNA$Nplus1 <- daily.N.df.long.noNA$N + 1  
```



```{r run-jags}
basis.dim <- 10
#family <- "gaussian"
family <- "Gamma"
# This is the output file name
jagam.model <- paste0("models/jagam_dailyN_k", 
                      basis.dim, "_", family, ".jags")

jags.out.file <- paste0("RData/jagam_dailyN_k", 
                        basis.dim, "_", family, "_jags.rds")

jags.data.v2 <- Ver2.results$jags.data
if (!file.exists(jags.out.file)){
  
  # Use jagam function to create a model file and input data list
  jags.gam <- jagam(Nplus1 ~ s(days, by = year, k = basis.dim), 
                    family = family,
                    data = daily.N.df.long.noNA,
                    file = jagam.model)
  
  # Run short chains to get what I need for now
  jm.jagam <- rjags::jags.model(file = jagam.model,
                                data = jags.gam$jags.data,
                                inits = rep(list(jags.gam$jags.ini), 
                                            MCMC.params$n.chains),
                                n.chains = MCMC.params$n.chains,
                                n.adapt = 500)
  
  jm.jagam.samples <- rjags::jags.samples(jm.jagam, 
                                          c("mu", "b", "scale", "rho"),
                                          n.iter = 1000, 
                                          n.thin = 5)
  
  jam <- sim2jam(jm.jagam.samples, jags.gam$pregam)
  
  pred.data <- data.frame(days = rep(1:90, n.seasons),
                          year = rep(1:n.seasons, each = 90))
  
  # prediciton by jagam
  Xp.matrix <- predict(jam, newdata = pred.data, type = "lpmatrix")
  Xp.link <- predict(jam, newdata = pred.data, se.fit = T)
  
  # I stack all S's into an array in Model_dailyN_GAM_kx_Nmix_JAGS.txt
  jags.data.1 <- jags.gam$jags.data
  all.S <- jags.data.1[grepl("S", names(jags.data.1))] 
  S.df <- do.call("rbind", all.S)
  	#b.start.idx[k] <- S.start.idx[k] + 1  #(k-1) * (x-1) + 2
	#b.end.idx[k] <- S.end.idx[k] + 1      # k * (x-1) + 1

  jags.data <- list(x = basis.dim,
                    n = jags.data.v2$n,
                    n.station = jags.data.v2$n.station,
                    n.year = jags.data.v2$n.year,
                    n.obs = jags.data.v2$n.obs,
                    periods = jags.data.v2$periods,
                    obs = jags.data.v2$obs,
                    u = jags.data.v2$u,
                    vs = jags.data.v2$vs,
                    bf = jags.data.v2$bf,
                    day = jags.data.v2$day,
                    N = jags.data.v2$N,
                    Watch.Length = jags.data.v2$Watch.Length,
                    X = jags.data.1$X,
                    y = jags.data.1$y,
                    n.data = jags.data.1$n,
                    zero = jags.data.1$zero,
                    S = S.df,
                    t.start = (((1:n.seasons)-1) * 90) + 1,
                    col.idx = rbind(rep(1, times = n.seasons), 
                                    matrix(2:73, ncol = n.seasons)),
                    new.X = pred.data,
                    S.start.idx = seq(from = 1, 
                                      to = (n.seasons-1) * (basis.dim-1) + 1,
                                      by = basis.dim - 1),
                    S.end.idx = seq(from = 9, 
                                    to = n.seasons * (basis.dim-1), 
                                    by = basis.dim-1))
  
  inits.list <- list(mean.prob = 0.5,
                     BF.Fixed = 0,
                     VS.Fixed = 0,
                     mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
                     beta.sigma = c(1,1,1),#beta.sigma = c(7,7,7),
                     BF.Switch = 1,
                     VS.Switch = 1,
                     OBS.Switch = 1,
                     sigma.Obs = 1,
                     N = Ver2.results$N_inits,
                     b = jags.gam$jags.ini$b,
                     omega = jags.gam$jags.ini$lambda)
  
  Start_Time<-Sys.time()
  jm.all <- jags(jags.data,
                 inits = rep(list(inits.list), 
                             MCMC.params$n.chains),
                 parameters.to.save= jags.params,
                 model.file = "models/Model_dailyN_GAM_kx_Nmix_JAGS.txt",
                 n.chains = MCMC.params$n.chains,
                 n.burnin = MCMC.params$n.burnin,
                 n.thin = MCMC.params$n.thin,
                 n.iter = MCMC.params$n.samples,
                 DIC = T, parallel=T)

  Run_Time <- Sys.time() - Start_Time
  Ver2.results <- list(jags.data = jags.data,
                       inits = inits.list,
                       JAGS_out = jm,
                       Run_Time = Run_Time,
                       Run_Date = Sys.Date())
  saveRDS(Ver2.results, jags.out.file)
    
  jags.all <- list(jags.gam = jags.gam,
                   jm = jm, 
                   jm.samples = jm.samples,
                   run.date = Sys.Date(),
                   MCMC.params = MCMC.params)
  
  saveRDS(jags.all,
          file = jags.out.file)
  
} else {
  
  jags.all <- readRDS(jags.out.file)
}


#plot(jam)



  

# original data
#jags.data.v2 <- Ver2.results$jags.data 

# select data that are needed:
jags.data <- list(n = jags.gam$jags.data$n,
                  n.station = jags.data.v2$n.station,
                  n.year = jags.data.v2$n.year,
                  n.obs = jags.data.v2$n.obs,
                  periods = jags.data.v2$periods,
                  obs = jags.data.v2$obs,
                  u = jags.data.v2$u,
                  vs = jags.data.v2$vs,
                  bf = jags.data.v2$bf,
                  day = jags.data.v2$day,
                  N = jags.data.v2$N,
                  Watch.Length = jags.data.v2$Watch.Length,
                  X = jagam.out$jags.gam$jags.data$X,
                  y = jagam.out$jags.gam$jags.data$y,
                  n.data = jagam.out$jags.gam$jags.data$n,
                  zero = jagam.out$jags.gam$jags.data$zero,
                  S1 = jagam.out$jags.gam$jags.data$S1,
                  S2 = jagam.out$jags.gam$jags.data$S2,
                  S3 = jagam.out$jags.gam$jags.data$S3,
                  S4 = jagam.out$jags.gam$jags.data$S4,
                  S5 = jagam.out$jags.gam$jags.data$S5,
                  S6 = jagam.out$jags.gam$jags.data$S6,
                  S7 = jagam.out$jags.gam$jags.data$S7,
                  S8 = jagam.out$jags.gam$jags.data$S8,
                  t.start = (((1:x)-1) * 90) + 1,
                  col.idx = rbind(rep(1, times = x), 
                                  matrix(2:73, ncol = x)),
                  new.X = Xp)


# 
# jagam.out.file <- paste0("RData/jagam_dailyN_k", basis.dim,"_", family, ".rds")



  
```


The following part is to see how GAM fit to the data:

```{r}
# 
jam <- sim2jam(jm.gam, jags.gam$pregam)
#plot(jam)
pred.data <- data.frame(days = rep(1:90, x),
                        year = rep(1:x, each = 90))

Xp <- predict(jam, newdata = pred.data, type = "lpmatrix")
pred <- predict(jam, newdata = pred.data)
pred.data$prediction <- 1/pred

# # SEnd in the prediction values - they are now fixed
# # this is kinda like using cut function, no? 
# jags.data$Xp <- Xp
# jags.data$col.idx <- rbind(rep(1, times = x), 
#                            matrix(2:73, ncol = x))
# jags.data$b <- summary(jm.gam2$b, FUN = mean)$stat
# jags.data$t.start <- (((1:x)-1) * 90) + 1

# For Gamma family, the link is invese, so we have to invert the vector
# to get the right numbers. 
ggplot() +
  geom_path(data = pred.data,
            aes(x = days, y = prediction)) +
  geom_point(data = daily.N.df.long,
             aes(x = days, y = N)) +
  facet_wrap(vars(year))

ggsave(filename = paste0("figures/GAM_fit_k", 
                         basis.dim, "_", family, ".png"),
       device = "png", dpi = 600)
```


<!-- Run jags if it hasn't been run.. -->

<!-- ```{r} -->
<!-- if (!file.exists(paste0("RData/JAGS_GAM_Nmix_", x, "yr_v2.rds"))){ -->

<!--   #Run time:  -->
<!--   Start_Time<-Sys.time() -->
<!--   rjags::load.module("glm") # Simon Wood's example shows this but it doesn't work  -->
<!--   jm <- jags(jags.data, -->
<!--              inits = rep(list(inits.list), MCMC.params$n.chains), -->
<!--              parameters.to.save= jags.params, -->
<!--              model.file = 'models/Model_GAM_Nmix_JAGS.txt', -->
<!--              n.chains = MCMC.params$n.chains, -->
<!--              n.burnin = MCMC.params$n.burnin, -->
<!--              n.thin = MCMC.params$n.thin, -->
<!--              n.iter = MCMC.params$n.samples, -->
<!--              DIC = T, parallel=T) -->

<!--   Run_Time <- Sys.time() - Start_Time -->
<!--   Ver2.results <- list(jags.data = jags.data, -->
<!--                        N_inits = N_inits, -->
<!--                        JAGS_out = jm, -->
<!--                        Run_Time = Run_Time) -->
<!--   saveRDS(Ver2.results,  -->
<!--           paste0("RData/JAGS_GAM_Nmix_", x, "yr_v2.rds")) -->

<!-- } else { -->
<!--   Ver2.results <- readRDS(paste0("RData/JAGS_GAM_Nmix_", x, "yr_v2.rds")) -->

<!-- } -->
<!-- ``` -->


<!-- Make some plots: -->

<!-- ```{r} -->
<!-- seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011",  -->
<!--              "2014/2015", "2015/2016", "2019/2020", "2021/2022") -->

<!-- # Extract estimated counts -->
<!-- Daily.Est <- Ver2.results$JAGS_out$sims.list$Daily.Est -->
<!-- sp <- Ver2.results$JAGS_out$sims.list$sp -->
<!-- com <- Ver2.results$JAGS_out$sims.list$com -->
<!-- Corrected.Est <- Ver2.results$JAGS_out$sims.list$Corrected.Est -->

<!-- # Each one of them is (# samples) x (90 days) x (# years) -->
<!-- # To plot them using ggplot's facet, I need to convert -->
<!-- # these into 2D dataframes of statistics (upper and lower  -->
<!-- # CIs, median, etc.) -->
<!-- # Daily.Est.list <- sp.list <- com.list <- vector(mode = "list",  -->
<!-- #                                                 length = dim(Daily.Est)[3]) -->
<!-- #  -->
<!-- # Daily.Est.UCIs <- Daily.Est.LCIs <- vector(mode = "list", -->
<!-- #                                            length = dim(Daily.Est)[3]) -->

<!-- stats.list <- vector(mode = "list", -->
<!--                      length = dim(Daily.Est)[3]) -->

<!-- for (k in 1:dim(Daily.Est)[3]){ -->
<!--   # Daily.Est.list[[k]] <- Daily.Est[,,k] -->
<!--   # Daily.Est.UCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.975) -->
<!--   # Daily.Est.LCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.275) -->
<!--   #  -->
<!--   # sp.list[[k]] <- sp[,,k] -->
<!--   # com.list[[k]] <- com[,,k] -->

<!--   stats.list[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est[,,k], 2, -->
<!--                                                          median), -->
<!--                                 Daily.Est.LCL = apply(Daily.Est[,,k], 2, -->
<!--                                                       quantile,0.275), -->
<!--                                 Daily.Est.UCL = apply(Daily.Est[,,k], 2, -->
<!--                                                       quantile,0.975), -->
<!--                                 sp.median = apply(exp(sp[,,k]), 2, -->
<!--                                                   median), -->
<!--                                 sp.LCL = apply(exp(sp[,,k]), 2, -->
<!--                                                quantile,0.025), -->
<!--                                 sp.UCL = apply(exp(sp[,,k]), 2, -->
<!--                                                quantile,0.975), -->
<!--                                 com.median = apply(exp(com[,,k]), 2, -->
<!--                                                    median), -->
<!--                                 com.LCL = apply(exp(com[,,k]), 2, -->
<!--                                                 quantile,0.025), -->
<!--                                 com.UCL = apply(exp(com[,,k]), 2, -->
<!--                                                 quantile,0.975), -->
<!--                                 #total.median = apply(exp(sp[,,k]), 1, sum), -->
<!--                                 days = 1:dim(Daily.Est)[2], -->
<!--                                 year = seasons[k]) -->
<!-- } -->

<!-- all.stats <- do.call("rbind", stats.list) %>% group_by(year) -->

<!-- ggplot(data = all.stats) +  -->
<!--   geom_line(aes(x = days, y = sp.median)) +  -->
<!--   geom_line(aes(x = days, y = com.median)) + -->
<!--   geom_ribbon(aes(x = days,  -->
<!--                   ymin = sp.LCL,  -->
<!--                   ymax = sp.UCL), -->
<!--               fill = "orange",  -->
<!--               alpha = 0.5) + -->
<!--   geom_line(aes(x = days, y = com.median), -->
<!--             color = "red") + -->
<!--   facet_wrap(vars(year)) + -->
<!--   xlab("Days since December 1") +  -->
<!--   ylab("Whales per day") -->

<!-- ``` -->



<!-- ```{r} -->

<!-- #Com vs Sp -->
<!-- ggplot(data = all.stats) + -->
<!--   geom_line(aes(x = days, y = sp.median)) + -->
<!--   geom_ribbon(aes(x = days,  -->
<!--                   ymin = sp.LCL,  -->
<!--                   ymax = sp.UCL), -->
<!--               fill = "orange", alpha = 0.5) +  -->
<!--   facet_wrap(vars(year))+ -->
<!--   xlab("Days since December 1") +  -->
<!--   ylab("Whales per day (spline)") -->



<!-- ``` -->


<!-- ```{r} -->

<!-- ggplot(data = all.stats) + -->
<!--   geom_line(aes(x = days, y = com.median)) + -->
<!--   geom_ribbon(aes(x = days,  -->
<!--                   ymin = com.LCL,  -->
<!--                   ymax = com.UCL), -->
<!--               fill = "orange", alpha = 0.5) +  -->
<!--   facet_wrap(vars(year))+ -->
<!--   xlab("Days since December 1") +  -->
<!--   ylab("Whales per day (Normal)") -->

<!-- ``` -->


<!-- ```{r} -->

<!-- ggplot(data = all.stats) + -->
<!--   geom_line(aes(x = days, y = Daily.Est.median)) + -->
<!--   geom_ribbon(aes(x = days,  -->
<!--                   ymin = Daily.Est.LCL,  -->
<!--                   ymax = Daily.Est.UCL), -->
<!--               fill = "orange",  -->
<!--               alpha = 0.5) +  -->
<!--   facet_wrap(vars(year))+ -->
<!--   xlab("Days since December 1") +  -->
<!--   ylab("Whales per day") -->


<!-- ``` -->

<!-- Total abundance -->
<!-- ```{r} -->
<!-- abundance.df <- data.frame(total.median = apply(Corrected.Est,  -->
<!--                                                 FUN = median,  -->
<!--                                                 MARGIN = 2), -->
<!--                            total.LCL = apply(Corrected.Est,  -->
<!--                                              MARGIN = 2,  -->
<!--                                              FUN = quantile, 0.025), -->
<!--                            total.UCL = apply(Corrected.Est,  -->
<!--                                              MARGIN = 2,  -->
<!--                                              FUN = quantile, 0.975), -->
<!--                            Season = seasons) -->


<!-- ggplot(data = abundance.df) +  -->
<!--   geom_point(aes(x = Season, y = total.median)) +  -->
<!--   geom_errorbar(aes(x = Season, ymin = total.LCL, ymax = total.UCL)) -->

<!-- #median(apply(exp(sp[,,4]),1,sum)) -->

<!-- ``` -->

