---
title: "R Notebook"
output: html_notebook
---

This notebook describes abundance estimation of gray whales using Durban's WinBUGS code and Ver1.0 data extraction method. Ver1.0 was developed by John Durban and updated by Josh Stewart. In this notebook, the new data from the 2021/2022 season are added to the previous data and run with WinBUGS. 

Set up libraries

```{r}
rm(list=ls())
library(abind)
library(jagsUI)
library(tidyverse)
library(bayesplot)
# PARAMETER NAME TRANSFERS:

# JOSH     JOHN
#    Spline 
# b.sp     b1 
# beta.sp  beta1
# X.sp     X1

# # periods for the 6 survey years
periods <-c(136, 135, 164, 178, 179, 151)

# Observer list gets updated as new observers are added. 
obs.list <- read.csv("Data/Observer list.csv", header = T) 
colnames(obs.list) <- c("obs", "ID")
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

x <- length(seasons)
# CHANGE THIS LINE BACK
#Ver2.results <- readRDS(paste0("RData/WinBUGS_", x, "yr_v2.rds"))
Ver2.results <- readRDS(paste0("RData/WinBUGS_", x, "yr_100k_v2.rds"))

```

Get data prior to the 2021/2022 season:
```{r}
jags.data <- Ver2.results$jags.data
jags.params <- c("lambda","OBS.RF","OBS.Switch",
                "BF.Switch","BF.Fixed","VS.Switch",
                "VS.Fixed","mean.prob","mean.prob.com",
                "mean.prob.sp","BF.Fixed.com",
                "BF.Fixed.sp","VS.Fixed.com",
                "VS.Fixed.sp",
                "Corrected.Est","Raw.Est","z",
                "com","sp","Daily.Est","mean.beta",
                "beta.sigma","beta","beta.sp","b.sp","sd.b.sp")

MCMC.params <- list(n.samples = 20000,
                   n.thin = 50,
                   n.burnin = 5000,
                   n.chains = 5)

# ni <- 100000
# nt <- 80
# nb <- 60000
# nc <- 3
N_inits <- Ver2.results$N_inits
inits.list <- list(mean.prob = 0.5,
                   BF.Fixed = 0,
                   VS.Fixed = 0,
                   mean.prob.sp = 0.5,
                   BF.Fixed.sp = 0,
                   VS.Fixed.sp = 0,
                   mean.prob.com = 0.5,
                   BF.Fixed.com = 0,
                   VS.Fixed.com = 0,
                   mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
                   beta.sigma = c(1,1,1),#beta.sigma = c(7,7,7),
                   BF.Switch = 1,
                   VS.Switch = 1,
                   OBS.Switch = 1,
                   sigma.Obs = 1,
                   BF.Switch.sp = 1,
                   VS.Switch.sp = 1,
                   OBS.Switch.sp = 1,
                   sigma.Obs.sp = 1,
                   BF.Switch.com = 1,
                   VS.Switch.com = 1,
                   OBS.Switch.com = 1,
                   sigma.Obs.com = 1,
                   N = N_inits,
                   N.com = N_inits,
                   N.sp = N_inits,
                   #z = matrix(1,nrow=90,ncol=6),
                   beta.sp = array(data=0, dim=c(2,x)),
                   sd.b.sp = rep(1, times = x), #c(1,1,1,1,1,1),
                   z = matrix(1, nrow=90, ncol= x))
  
```

Run jags if it hasn't been run..

```{r}
if (!file.exists(paste0("RData/JAGS_Weighted_Com_Sp_", x, "yr_v2.rds"))){
  
  #Run time: 
  Start_Time<-Sys.time()
  jm <- jags(jags.data,
             inits = rep(list(inits.list), MCMC.params$n.chains),
             parameters.to.save= jags.params,
             model.file = 'models/Model_Weighted_Com_Sp_Nmix_JAGS.txt',
             n.chains = MCMC.params$n.chains,
             n.burnin = MCMC.params$n.burnin,
             n.thin = MCMC.params$n.thin,
             n.iter = MCMC.params$n.samples,
             DIC = T, parallel=T)
    
  Run_Time <- Sys.time() - Start_Time
  Ver2.results <- list(jags.data = jags.data,
                       N_inits = N_inits,
                       JAGS_out = jm,
                       Run_Time = Run_Time)
  saveRDS(Ver2.results, 
          paste0("RData/JAGS_Weighted_Com_Sp_", x, "yr_v2.rds"))
  
} else {
  Ver2.results <- readRDS(paste0("RData/JAGS_Weighted_Com_Sp_", x, "yr_v2.rds"))
  
}
```


Make some plots:

```{r}
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

# Extract estimated counts
Daily.Est <- Ver2.results$JAGS_out$sims.list$Daily.Est
sp <- Ver2.results$JAGS_out$sims.list$sp
com <- Ver2.results$JAGS_out$sims.list$com
Corrected.Est <- Ver2.results$JAGS_out$sims.list$Corrected.Est

# Each one of them is (# samples) x (90 days) x (# years)
# To plot them using ggplot's facet, I need to convert
# these into 2D dataframes of statistics (upper and lower 
# CIs, median, etc.)
# Daily.Est.list <- sp.list <- com.list <- vector(mode = "list", 
#                                                 length = dim(Daily.Est)[3])
# 
# Daily.Est.UCIs <- Daily.Est.LCIs <- vector(mode = "list",
#                                            length = dim(Daily.Est)[3])

stats.list <- vector(mode = "list",
                     length = dim(Daily.Est)[3])

for (k in 1:dim(Daily.Est)[3]){
  # Daily.Est.list[[k]] <- Daily.Est[,,k]
  # Daily.Est.UCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.975)
  # Daily.Est.LCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.275)
  # 
  # sp.list[[k]] <- sp[,,k]
  # com.list[[k]] <- com[,,k]
  
  stats.list[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est[,,k], 2,
                                                         median),
                                Daily.Est.LCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.275),
                                Daily.Est.UCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.975),
                                sp.median = apply(exp(sp[,,k]), 2,
                                                  median),
                                sp.LCL = apply(exp(sp[,,k]), 2,
                                               quantile,0.025),
                                sp.UCL = apply(exp(sp[,,k]), 2,
                                               quantile,0.975),
                                com.median = apply(exp(com[,,k]), 2,
                                                   median),
                                com.LCL = apply(exp(com[,,k]), 2,
                                                quantile,0.025),
                                com.UCL = apply(exp(com[,,k]), 2,
                                                quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est)[2],
                                year = seasons[k])
}

all.stats <- do.call("rbind", stats.list) %>% group_by(year)

ggplot(data = all.stats) + 
  geom_line(aes(x = days, y = sp.median)) + 
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", 
              alpha = 0.5) +
  geom_line(aes(x = days, y = com.median),
            color = "red") +
  facet_wrap(vars(year)) +
  xlab("Days since December 1") + 
  ylab("Whales per day")

```



```{r}

#Com vs Sp
ggplot(data = all.stats) +
  geom_line(aes(x = days, y = sp.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day (spline)")



```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = com.LCL, 
                  ymax = com.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day (Normal)")

```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = Daily.Est.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = Daily.Est.LCL, 
                  ymax = Daily.Est.UCL),
              fill = "orange", 
              alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day")


```

Total abundance
```{r}
abundance.df <- data.frame(total.median = apply(Corrected.Est, 
                                                FUN = median, 
                                                MARGIN = 2),
                           total.LCL = apply(Corrected.Est, 
                                             MARGIN = 2, 
                                             FUN = quantile, 0.025),
                           total.UCL = apply(Corrected.Est, 
                                             MARGIN = 2, 
                                             FUN = quantile, 0.975),
                           Season = seasons)


ggplot(data = abundance.df) + 
  geom_point(aes(x = Season, y = total.median)) + 
  geom_errorbar(aes(x = Season, ymin = total.LCL, ymax = total.UCL))

#median(apply(exp(sp[,,4]),1,sum))

```

