---
title: "R Notebook"
output: html_notebook
---

This notebook describes abundance estimation of gray whales using a modified version of Durban's WinBUGS code and Ver2.0 data extraction method. In this version, I replace the normal/spline curves with GAMs. 


Set up libraries

```{r}
rm(list=ls())
library(abind)
library(jagsUI)
library(tidyverse)
library(bayesplot)
library(mgcv)

# # periods for the 6 survey years
periods <-c(136, 135, 164, 178, 179, 151)

# Observer list gets updated as new observers are added. 
obs.list <- read.csv("Data/Observer list.csv", header = T) 
colnames(obs.list) <- c("obs", "ID")
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

x <- length(seasons)

# basis dimension - changes the jags script
basis.dim <- 10
#family <- "gaussian"
family <- "gamma"

# CHANGE THIS LINE BACK
#Ver2.results <- readRDS(paste0("RData/WinBUGS_", x, "yr_v2.rds"))
Ver2.results <- readRDS(paste0("RData/WinBUGS_", x, "yr_100k_v2.rds"))
jagam.out <- readRDS(paste0("RData/jagam_dailyN_k", basis.dim, "_", family, ".rds"))
jags.fit <- readRDS(paste0("RData/jagam_dailyN_k", basis.dim, "_", family, "_jags.rds"))

jam <- sim2jam(jags.fit$jm.gam, jagam.out$jags.gam$pregam)

#plot(jam)
pred.data <- data.frame(days = rep(1:90, x),
                        year = rep(1:x, each = 90))

Xp <- predict(jam, newdata = pred.data, type = "lpmatrix")

# original data
jags.data.v2 <- Ver2.results$jags.data 

# select data that are needed:
jags.data <- list(n = jags.data.v2$n,
                  n.station = jags.data.v2$n.station,
                  n.year = jags.data.v2$n.year,
                  n.obs = jags.data.v2$n.obs,
                  periods = jags.data.v2$periods,
                  obs = jags.data.v2$obs,
                  u = jags.data.v2$u,
                  vs = jags.data.v2$vs,
                  bf = jags.data.v2$bf,
                  day = jags.data.v2$day,
                  N = jags.data.v2$N,
                  Watch.Length = jags.data.v2$Watch.Length,
                  X = jagam.out$jags.gam$jags.data$X,
                  y = jagam.out$jags.gam$jags.data$y,
                  n.data = jagam.out$jags.gam$jags.data$n,
                  zero = jagam.out$jags.gam$jags.data$zero,
                  S1 = jagam.out$jags.gam$jags.data$S1,
                  S2 = jagam.out$jags.gam$jags.data$S2,
                  S3 = jagam.out$jags.gam$jags.data$S3,
                  S4 = jagam.out$jags.gam$jags.data$S4,
                  S5 = jagam.out$jags.gam$jags.data$S5,
                  S6 = jagam.out$jags.gam$jags.data$S6,
                  S7 = jagam.out$jags.gam$jags.data$S7,
                  S8 = jagam.out$jags.gam$jags.data$S8,
                  t.start = (((1:x)-1) * 90) + 1,
                  col.idx = rbind(rep(1, times = x), 
                                  matrix(2:73, ncol = x)),
                  new.X = Xp)


jags.params <- c("lambda", "OBS.RF", "OBS.Switch",
                "BF.Switch", "BF.Fixed", "VS.Switch",
                "VS.Fixed", "mean.prob", "mean.prob.com",
                "Corrected.Est", "Raw.Est",
                "gam.fit", "mean.beta",
                "beta.sigma","beta", "scale.daily.N",
                "b")

MCMC.params <- list(n.samples = 10000,
                   n.thin = 50,
                   n.burnin = 5000,
                   n.chains = 5)

# ni <- 100000
# nt <- 80
# nb <- 60000
# nc <- 3

inits.list <- list(mean.prob = 0.5,
                   BF.Fixed = 0,
                   VS.Fixed = 0,
                   mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
                   beta.sigma = c(1,1,1),#beta.sigma = c(7,7,7),
                   BF.Switch = 1,
                   VS.Switch = 1,
                   OBS.Switch = 1,
                   sigma.Obs = 1,
                   N = Ver2.results$N_inits,
                   b = jagam.out$jags.gam$jags.ini$b,
                   omega = jagam.out$jags.gam$jags.ini$lambda)
                   #z = matrix(1,nrow=90,ncol=6),

  
```


For the seasonal curves, a separate analysis was done in JAGS GAM Ver2.Rmd, where GAM was fitted to the observed numbers of whales per day. The 'jagam' function from mgcv was used to create a jags GAM script and a data list. The script was incorporated into the script that fits counts from search periods (1.5 hr shifts). 

Run jags if it hasn't been run..

```{r}
if (!file.exists(paste0("RData/JAGS_dailyN_GAM_k", 
                        basis.dim, "_", family, 
                        "_Nmix_", x, "yr_v2.rds"))){
  
  #Run time: 
  Start_Time<-Sys.time()
  jm <- jags(jags.data,
             inits = rep(list(inits.list), MCMC.params$n.chains),
             parameters.to.save= jags.params,
             model.file = paste0("models/Model_dailyN_GAM_k", 
                                 basis.dim, "_", family, "_Nmix_JAGS.txt"),
             n.chains = MCMC.params$n.chains,
             n.burnin = MCMC.params$n.burnin,
             n.thin = MCMC.params$n.thin,
             n.iter = MCMC.params$n.samples,
             DIC = T, parallel=T)
    
  Run_Time <- Sys.time() - Start_Time
  Ver2.results <- list(jags.data = jags.data,
                       inits = inits.list,
                       JAGS_out = jm,
                       Run_Time = Run_Time,
                       Run_Date = Sys.Date())
  saveRDS(Ver2.results, 
          paste0("RData/JAGS_dailyN_GAM_k",
                 basis.dim, "_", family, "_Nmix_", x, "yr_v2.rds"))
  
} else {
  Ver2.results <- readRDS(paste0("RData/JAGS_dailyN_GAM_k", 
                                 basis.dim, "_", family, 
                                 "_Nmix_", x, "yr_v2.rds"))
  
}
```


Make some plots:

```{r}
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

# Extract estimated counts
# Daily.Est <- Ver2.results$JAGS_out$sims.list$Daily.Est
# sp <- Ver2.results$JAGS_out$sims.list$sp
# com <- Ver2.results$JAGS_out$sims.list$com
Corrected.Est <- Ver2.results$JAGS_out$sims.list$Corrected.Est

# Each one of them is (# samples) x (90 days) x (# years)
# To plot them using ggplot's facet, I need to convert
# these into 2D dataframes of statistics (upper and lower 
# CIs, median, etc.)
# Daily.Est.list <- sp.list <- com.list <- vector(mode = "list", 
#                                                 length = dim(Daily.Est)[3])
# 
# Daily.Est.UCIs <- Daily.Est.LCIs <- vector(mode = "list",
#                                            length = dim(Daily.Est)[3])

stats.list <- vector(mode = "list",
                     length = dim(Corrected.Est)[3])

for (k in 1:dim(Daily.Est)[3]){
  # Daily.Est.list[[k]] <- Daily.Est[,,k]
  # Daily.Est.UCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.975)
  # Daily.Est.LCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.275)
  # 
  # sp.list[[k]] <- sp[,,k]
  # com.list[[k]] <- com[,,k]
  
  stats.list[[k]] <- data.frame(Corrected.Est.median = apply(Daily.Est[,,k], 2,
                                                         median),
                                Corrected.Est.LCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.275),
                                Corrected.Est.UCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est)[2],
                                year = seasons[k])
}

all.stats <- do.call("rbind", stats.list) %>% group_by(year)

ggplot(data = all.stats) + 
  geom_line(aes(x = days, y = sp.median)) + 
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", 
              alpha = 0.5) +
  geom_line(aes(x = days, y = com.median),
            color = "red") +
  facet_wrap(vars(year)) +
  xlab("Days since December 1") + 
  ylab("Whales per day")

```



```{r}

#Com vs Sp
ggplot(data = all.stats) +
  geom_line(aes(x = days, y = sp.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day (spline)")



```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = com.LCL, 
                  ymax = com.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day (Normal)")

```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = Daily.Est.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = Daily.Est.LCL, 
                  ymax = Daily.Est.UCL),
              fill = "orange", 
              alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day")


```

Total abundance
```{r}
abundance.df <- data.frame(total.median = apply(Corrected.Est, 
                                                FUN = median, 
                                                MARGIN = 2),
                           total.LCL = apply(Corrected.Est, 
                                             MARGIN = 2, 
                                             FUN = quantile, 0.025),
                           total.UCL = apply(Corrected.Est, 
                                             MARGIN = 2, 
                                             FUN = quantile, 0.975),
                           Season = seasons)


ggplot(data = abundance.df) + 
  geom_point(aes(x = Season, y = total.median)) + 
  geom_errorbar(aes(x = Season, ymin = total.LCL, ymax = total.UCL))

#median(apply(exp(sp[,,4]),1,sum))

```

