---
title: "Richards HSSM results summary"
output: html_notebook
---

This notebook shows various results from the Hierarchial State Space Model approach using the Richards function to estimate the abundance of gray whales. The models should be fitted using 'Jags_Richards_AllData_n_models.R'. The script saves the results in .rds files to the '/RData' directory. This notebook uses those outputs to create various summaries of the analysis.

```{r}
# Whether or not to save the figures
save.fig <- T

source("Granite_Canyon_Counts_fcns.R")
source("Richards_HSSM_model_definition.R")

library(tidyverse)
library(lubridate)
library(flextable)
library(jagsUI)
library(bayesplot)
library(ggpubr)
#library(R2WinBUGS)
library(abind)
library(rmarkdown)
library(loo)
library(ggridges)
library(rstanarm)
library(mgcv)

# Define the latest year to be analyzed
YEAR <- 2026
# Minimum shift duration to be included
min.dur <- 60#10 
# dpi for output figures
dpi.set <- 600

# The date when WinBUGS was run on the same dataset. This is run in the GrayWhaleAbundance project.
WinBUGS.Run.Date <- "2026-02-20"  

# Create a function to save figures if requested.
save.fig.fcn <- function(fig, file.name, replace = F, 
                         dpi = dpi.set, device = "png",
                         height = 3, width = 3, units = "in",
                         bg = "white"){
  if (isTRUE(replace) | (!isTRUE(replace) & !file.exists(file.name)))
    ggsave(fig, 
           filename = file.name, 
           dpi = dpi, device = device,
           height = height, width = width, units = units,
           bg = bg)
  
}

# Define models - this can be adjusted if needed or possible
# If new mdoels were added or the definitions chnaged, make sure to edit 'Richards_HSSM_model_definition.R'
all.model.names <- c("M1a1", "M2a1", "M3a1", "M4a1","M1a2", "M2a2", "M3a2", "M4a2")
model.defs <- data.frame(ID = all.model.names,
                         Lkhd = c(rep("POI", 4), rep("NB", 4)),
                         S1 = rep(c("Season", "Constant"), 4),
                         S2 = rep(c("Season", "Constant", "Constant", "Season"), 2))

```

Once the models were fit, the model comparison step is done using 'Jags_Richards_ModelComparison.R'. Results are saved in a .rds file. Bring in the output and create a table. Then bring in the best model output for inference. 

```{r}
out.table <- readRDS(paste0("RData/Richards_ModelComparison_", YEAR, ".rds"))
out.table %>% 
  select(model, n.params, n.big.Rhat, p.big.Rhat, n.bad.Pareto, p.bad.Pareto, min.ESS.bulk, min.ESS.tail, dLOOIC) %>%
  flextable() %>%
  colformat_double(j = c("dLOOIC", "p.big.Rhat", "p.bad.Pareto",
                         "min.ESS.bulk", "min.ESS.tail"),
                   digits = 2) %>%
  compose(part = "header", j = "dLOOIC",
          value = as_paragraph(as_chunk("\u0394"), as_chunk("LOOIC"))) -> ft.models.comp


ver <- out.table[1, "model"] # "M1a2" #5 #out.table$model[1] #5
jm.out <- readRDS(paste0("RData/JAGS_Richards_HSSM_",
                         ver,
                         "_1968to", YEAR, "_min", min.dur,
                         "_NoBUGS.rds"))

# bring in the convergence statistics:
# This is a list of length 4. They are LOOIC, Rhat, ESS.bulk and ESS.tail
# Each item has a length equals to the number of models ran. 
ver.idx <- which(all.model.names == ver)

# The convergence statistics are saved by 'Jags_Richards_ModelComparison.R'. 
# Rather than re-calculating those statistics, bring in the output.
list.convergence <- readRDS(paste0("RData/Richards_Convergence_", YEAR, ".rds"))

# LOOIC and Pareto-k stats
LOOIC.n <- list.convergence$LOOIC[ver.idx] 

# better ESS computations:
summary.posterior <- jm.out$posterior.summary

ESS.bulk <- list.convergence$ESS.bulk[[ver.idx]]
ESS.tail <- list.convergence$ESS.tail[[ver.idx]]

# Match Pareto-k and data. Note that days 1 and 100 should be removed.
n.array <- jm.out$jags.input$jags.data$n
days <- jm.out$jags.input$jags.data$day
watch.length <- jm.out$jags.input$jags.data$watch.length

all.start.year <- c(jm.out$jags.input$jags.input.Laake$all.start.year,
                    jm.out$jags.input$jags.input.new$start.years)

# Create a dataframe with all years, including unsampled years.
all.years <- data.frame(start.year = seq(min(all.start.year), max(all.start.year))) %>%
  mutate(Season = paste0(start.year, "/", start.year + 1))

# Look at the annual abundance estimates:
Nhat. <- data.frame(Season = paste0(all.start.year, "/",
                                    all.start.year+1),
                    Nhat = jm.out$jm$q50$Corrected.Est,
                    LCL = jm.out$jm$q2.5$Corrected.Est,
                    UCL = jm.out$jm$q97.5$Corrected.Est) %>%
  right_join(all.years, by = "Season") %>%
  arrange(start.year) %>%
  mutate(Method = paste0("Eguchi ", ver))

# Create an observed counts dataframe
obs.day <- jm.out$jags.input$jags.data$day[,1,]
obs.n <- jm.out$jags.input$jags.data$n[,1,]
obs.prop <- jm.out$jags.input$jags.data$watch.length[,1,]
obs.prop <- rbind(c(rep(0, ncol(obs.n))),
                  obs.prop,
                  c(rep(0, ncol(obs.n))))

obs.n.df <- data.frame(start.year = rep(all.start.year, each = 100),
                       Season = rep(paste0(all.start.year, "/",
                                           all.start.year+1),
                                    each = 100),
                       Day = rep(1:100, times = length(all.start.year)),
                       n = NA,
                       prop = NA)

k1 <- k2 <- 1
for (k1 in 1:length(all.start.year)){
  obs.day.1 <- obs.day[, k1]
  obs.n.1 <- obs.n[, k1]
  obs.day.1.uniq <- na.omit(unique(obs.day.1))
  for (k2 in 1:length(obs.day.1.uniq)){
    obs.n.df[obs.n.df$start.year == all.start.year[k1] &
               obs.n.df$Day == obs.day.1.uniq[k2], "n"] <- sum(obs.n[obs.day[,k1] == obs.day.1.uniq[k2], k1], na.rm = T)
    obs.n.df[obs.n.df$start.year == all.start.year[k1] &
               obs.n.df$Day == obs.day.1.uniq[k2], "prop"] <- sum(obs.prop[obs.day[,k1] == obs.day.1.uniq[k2], k1], na.rm = T)
  }
  
}

# This is for daily estimates
N.hats.day <- data.frame(Season = rep(paste0(all.start.year, 
                                             "/", all.start.year+1),
                                      
                                      each = nrow(jm.out$jm$mean$mean.N)), 
                         start.year = rep(all.start.year,
                                          each = nrow(jm.out$jm$mean$mean.N)),
                         Day = rep(1:nrow(jm.out$jm$mean$mean.N), 
                                   times = length(all.start.year)),
                         Mean = as.vector(jm.out$jm$mean$mean.N),
                         LCL = as.vector(jm.out$jm$q2.5$mean.N),
                         UCL = as.vector(jm.out$jm$q97.5$mean.N),
                         obs.n = obs.n.df$n,
                         prop = obs.n.df$prop,
                         Method = paste0("Eguchi ", ver)) %>%
  mutate(Nhat = obs.n/prop)

# Daily estimates plots
p.daily.Richards <- ggplot(N.hats.day %>% 
                             group_by(Season) %>%
                             arrange(by = Day)) + 
  geom_ribbon(aes(x = Day, ymin = LCL, ymax = UCL),
              fill = "blue", alpha = 0.5) +
  geom_path(aes(x = Day, y = Mean)) + 
  geom_point(aes(x = Day, y = Nhat), color = "darkgreen") +
  facet_wrap(~ Season) +
  theme_bw()

# These same plots were also created with NewGrayWhaleAbundanceEstimation_vx.Rmd
# In order to differentiate plots from this .Rmd file, I changed the output figure 
# file names.
save.fig.fcn(fig = p.daily.Richards, 
             file.name = paste0("figures/Daily_HSSM_Eguchi_", ver, "_", ".png"), 
             replace = T, 
             dpi = dpi.set, device = "png",
             height = 10, width = 10, units = "in",
             bg = "white")

# These estimates are from the annual report (created in the
# GrayWhaleAbundance project and using WinBUGS output). The file is
# created when an annual report is written. This needs to be fixed
# because the report may not be written using the same script. 
Reported.estimates <- read.csv(file = paste0("Data/Nhats_", YEAR, ".csv")) %>%  
  transmute(Season = Season,
            Nhat = Nhat,
            LCL = LCL,
            UCL = UCL,
            Method = paste0(Method, "-Reported")) %>%
  right_join(all.years, by = "Season") %>%
  arrange(start.year) %>%
  relocate(Method, .after = start.year)

```

Bring in the Laake analysis output 

```{r}
# The following needs output from LaakeAnalysis_NewData.R
# Run the LaakeAnalysis_NewData.R script with the newly acquired data.
# In the script, change the years variable to include the most
# up-to-date data. It should read all extracted data and run the
# Laake's analysis and write results in a .csv file.
Laake.Run.Date <- "2026-02-23"
Laake.abundance.new <- read.csv(file = paste0("Data/all_estimates_Laake_", 
                                              YEAR, "_", Laake.Run.Date, ".csv")) %>%
  mutate(LCL = CL.low,
         UCL = CL.high) %>%
  select(c(Season, Nhat, LCL, UCL)) %>%
  right_join(all.years, by = "Season") %>%
  arrange(start.year) %>%
  mutate(Method = "Laake")

#Laake.output <- read_rds(file = "RData/Laake_abundance_estimates_2024.rds")

# In reported estimates, there are two 2006/2007.
Reported.estimates %>%
  na.omit() %>%
  select(Season) %>% 
  unique() -> sampled.seasons 

# Reported estimates are not exactly identical to the reanalysis. For consistency,
# I use the reported estimates.  
# Laake.abundance.new %>%
#   rbind(Durban.abundance.df) %>%
Reported.estimates %>%
  mutate(Method1 = sub("-.*", "", Method)) -> tmp.1

tmp.1[grep("NA", tmp.1$Method1) & tmp.1$start.year < 2006, 
      "Method1"] <- "Laake"
tmp.1[grep("NA", tmp.1$Method1) & tmp.1$start.year > 2007, 
      "Method1"] <- "Durban"

tmp.1 %>%
  select(Season, Nhat, LCL, UCL, start.year, Method1) %>%
  rename(Method = Method1) %>% #-> tmp.2
  rbind(Laake.abundance.new %>% 
          filter(start.year > 2006)) %>%
  rbind(Nhat.) %>%
  mutate(Method2 = factor(Method, 
                          levels = c("Laake", 
                                     "Durban", 
                                     paste0("Eguchi ", ver)))) %>%
  select(Season, Nhat, LCL, UCL, start.year, Method2) %>%
  mutate(CI.width = UCL - LCL) %>%
  rename(Method = Method2) -> all.estimates

#  rbind(spline.Nhat) 
  #rbind(Reported.estimates %>% na.omit()) -> all.estimates

p.Nhats <- ggplot(all.estimates) +
  geom_point(aes(x = start.year, y = Nhat,
                 color = Method),
             alpha = 0.5) +
  geom_errorbar(aes(x = start.year, ymin = LCL, ymax = UCL,
                    color = Method)) +
  ylim(5000, 35000) +
  xlab("") + ylab("Abundance") +
  theme_bw() +
  theme(legend.position = "top") 

if (save.fig)
  save.fig.fcn(fig = p.Nhats,
               file.name = paste0("figures/Nhats_HSSM_", 
                                  ver, "_min", min.dur, 
                                  "_", dpi.set, "dpi.png"),
               replace = T,
               height = 5, width = 7,
               dpi = dpi.set)

Nhat. %>% 
  select(Season, start.year, Nhat, LCL, UCL) %>%
  rename(Nhat.Eguchi = Nhat,
         LCL.Eguchi = LCL,
         UCL.Eguchi = UCL) %>%
  cbind(Laake.abundance.new %>%
          select(Nhat, LCL, UCL) %>%
          rename(Nhat.Laake = Nhat,
                 LCL.Laake = LCL,
                 UCL.Laake = UCL)) %>%
  cbind(Durban.abundance.df %>%
          select(Nhat, LCL, UCL) %>%
          rename(Nhat.Durban = Nhat,
                 LCL.Durban = LCL,
                 UCL.Durban = UCL)) %>%
  mutate(d.Laake.Eguchi = Nhat.Laake - Nhat.Eguchi,
         d.Durban.Eguchi = Nhat.Durban - Nhat.Eguchi) -> Nhat.all.wide


```

Look at the marginal posterior distributions of the Richards function

```{r}
#Posteriors on P, S1, and S2
posteriors.P <- plot.trace.dens(jm.out$jm, 
                                var.name = "^P\\[")

posteriors.P$df %>%
  mutate(start.year = all.start.year[numeric.index],
         start.year.f = as.factor(start.year)) %>% 
  rename(P = sample)-> P.df

p.posteriors.P <- ggplot(P.df, aes(x = P, y = start.year.f)) +
  geom_density_ridges() +
  xlab("Ordinal Day of Season") +
  ylab("Start year") +
  theme_bw()

if (save.fig)
  save.fig.fcn(fig = p.posteriors.P,
               file.name = paste0("figures/posterior_P_HSSM_", 
                                  ver, "_", dpi.set, "dpi.png"),
               replace = T,
               height = 5, width = 7,
               dpi = dpi.set)

# Need to include \\[ to remove S1.alpha and S1.beta
if (length(grep("M1|M3", ver)) > 0){
  posteriors.S1 <- plot.trace.dens(jm.out$jm,
                                   var.name = "S1\\[")
  posteriors.S1$df %>%
  mutate(start.year = all.start.year[numeric.index],
         start.year.f = as.factor(start.year)) %>% 
  rename(S1 = sample)-> S1.df

  p.posteriors.S1 <- ggplot(S1.df, aes(x = S1, y = start.year.f)) +
    geom_density_ridges() +
    xlim(c(1, 10)) +
    xlab(expression(S[1])) +
    ylab("Start year") +
  theme_bw()
  
} else {
  posteriors.S1 <- plot.trace.dens(jm.out$jm,
                                   var.name = "^S1\\b(?!\\.\\w*)")
  p.posteriors.S1 <- posteriors.S1$p.dens
}

if (save.fig)
  save.fig.fcn(fig = p.posteriors.S1,
               file.name = paste0("figures/posteriors_S1_HSSM_", 
                                  ver, "_", dpi.set, "dpi.png"),
               replace = T,
               height = 5, width = 7,
               dpi = dpi.set)

if (length(grep("M2|M3", ver)) > 0){
  posteriors.S2 <- plot.trace.dens(jm.out$jm, 
                                   var.name = "S2\\b(?!\\.\\w*)")
  p.posteriors.S2 <- posteriors.S2$p.dens
} else {
  posteriors.S2 <- plot.trace.dens(jm.out$jm, var.name = "S2\\[")
  posteriors.S2$df %>%
    mutate(start.year = all.start.year[numeric.index],
           start.year.f = as.factor(start.year)) %>% 
    rename(S2 = sample)-> S2.df
  
  p.posteriors.S2 <- ggplot(S2.df, aes(x = S2, y = start.year.f)) +
    geom_density_ridges() +
    xlim(c(1, 10)) +
    xlab(expression(S[2])) +
    ylab("Start year") +
  theme_bw()
  
} 

if (save.fig)
  save.fig.fcn(fig = p.posteriors.S2,
               file.name = paste0("figures/posteriors_S2_HSSM_", 
                                  ver, "_", dpi.set, "dpi.png"),
               replace = T,
               height = 5, width = 7,
               dpi = dpi.set)

posterior.theta <- plot.trace.dens(jm.out$jm,
                                   var.name = "^r")

posterior.theta$p.dens +
  labs(x = expression(theta)) -> p.posterior.theta

if (save.fig)
  save.fig.fcn(fig = p.posterior.theta,
               file.name = paste0("figures/posterior_theta_HSSM_", 
                                  ver, "_", dpi.set, "dpi.png"),
               replace = T,
               height = 5, width = 7,
               dpi = dpi.set)

```

