---
title: "R Notebook"
output: html_notebook
---

This notebook describes abundance estimation of gray whales using a modified version of Durban's WinBUGS code and Ver2.0 data extraction method. In this version, I replace the normal/spline curves with GAMs. 


Set up libraries

```{r}
rm(list=ls())
library(abind)
library(jagsUI)
library(tidyverse)
library(bayesplot)
library(mgcv)

# # periods for the 6 survey years
periods <-c(136, 135, 164, 178, 179, 151)

# Observer list gets updated as new observers are added. 
obs.list <- read.csv("Data/Observer list.csv", header = T) 
colnames(obs.list) <- c("obs", "ID")
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

x <- length(seasons)

# basis dimension - changes the jags script
basis.dim <- 10
#family <- "gaussian"
family <- "gamma"

# CHANGE THIS LINE BACK
#Ver2.results <- readRDS(paste0("RData/WinBUGS_", x, "yr_v2.rds"))
Ver2.results <- readRDS(paste0("RData/WinBUGS_", x, "yr_100k_v2.rds"))
jagam.out <- readRDS(paste0("RData/jagam_dailyN_k", basis.dim, "_", family, ".rds"))
jags.fit <- readRDS(paste0("RData/jagam_dailyN_k", basis.dim, "_", family, "_jags.rds"))

jam <- sim2jam(jags.fit$jm.gam, jagam.out$jags.gam$pregam)

#plot(jam)
pred.data <- data.frame(days = rep(1:90, x),
                        year = rep(1:x, each = 90))

Xp <- predict(jam, newdata = pred.data, type = "lpmatrix")

# original data
jags.data.v2 <- Ver2.results$jags.data 

# select data that are needed:
jags.data <- list(n = jags.data.v2$n,
                  n.station = jags.data.v2$n.station,
                  n.year = jags.data.v2$n.year,
                  n.obs = jags.data.v2$n.obs,
                  periods = jags.data.v2$periods,
                  obs = jags.data.v2$obs,
                  u = jags.data.v2$u,
                  vs = jags.data.v2$vs,
                  bf = jags.data.v2$bf,
                  day = jags.data.v2$day,
                  N = jags.data.v2$N,
                  Watch.Length = jags.data.v2$Watch.Length,
                  X = jagam.out$jags.gam$jags.data$X,
                  y = jagam.out$jags.gam$jags.data$y,
                  n.data = jagam.out$jags.gam$jags.data$n,
                  zero = jagam.out$jags.gam$jags.data$zero,
                  S1 = jagam.out$jags.gam$jags.data$S1,
                  S2 = jagam.out$jags.gam$jags.data$S2,
                  S3 = jagam.out$jags.gam$jags.data$S3,
                  S4 = jagam.out$jags.gam$jags.data$S4,
                  S5 = jagam.out$jags.gam$jags.data$S5,
                  S6 = jagam.out$jags.gam$jags.data$S6,
                  S7 = jagam.out$jags.gam$jags.data$S7,
                  S8 = jagam.out$jags.gam$jags.data$S8,
                  t.start = (((1:x)-1) * 90) + 1,
                  col.idx = rbind(rep(1, times = x), 
                                  matrix(2:73, ncol = x)),
                  new.X = Xp)


jags.params <- c("lambda", "OBS.RF", "OBS.Switch",
                "BF.Switch", "BF.Fixed", "VS.Switch",
                "VS.Fixed", "mean.prob", "mean.prob.com",
                "Corrected.Est", "Raw.Est",
                "gam.fit", "mean.beta",
                "beta.sigma","beta", "scale.daily.N",
                "b")

MCMC.params <- list(n.samples = 200000,
                   n.thin = 10,
                   n.burnin = 50000,
                   n.chains = 5)

# ni <- 100000
# nt <- 80
# nb <- 60000
# nc <- 3

inits.list <- list(mean.prob = 0.5,
                   BF.Fixed = 0,
                   VS.Fixed = 0,
                   mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
                   beta.sigma = c(1,1,1),#beta.sigma = c(7,7,7),
                   BF.Switch = 1,
                   VS.Switch = 1,
                   OBS.Switch = 1,
                   sigma.Obs = 1,
                   N = Ver2.results$N_inits,
                   b = jagam.out$jags.gam$jags.ini$b,
                   omega = jagam.out$jags.gam$jags.ini$lambda)
                   #z = matrix(1,nrow=90,ncol=6),

  
```


For the seasonal curves, a separate analysis was done in JAGS GAM Ver2.Rmd, where GAM was fitted to the observed numbers of whales per day. The 'jagam' function from mgcv was used to create a jags GAM script and a data list. The script was incorporated into the script that fits counts from search periods (1.5 hr shifts). 

Run jags if it hasn't been run..

```{r}
if (!file.exists(paste0("RData/JAGS_dailyN_GAM_k", 
                        basis.dim, "_", family, 
                        "_Nmix_", x, "yr_v2.rds"))){
  
  #Run time: 
  Start_Time<-Sys.time()
  jm <- jags(jags.data,
             inits = rep(list(inits.list), MCMC.params$n.chains),
             parameters.to.save= jags.params,
             model.file = paste0("models/Model_dailyN_GAM_k", 
                                 basis.dim, "_", family, "_Nmix_JAGS.txt"),
             n.chains = MCMC.params$n.chains,
             n.burnin = MCMC.params$n.burnin,
             n.thin = MCMC.params$n.thin,
             n.iter = MCMC.params$n.samples,
             DIC = T, parallel=T)
    
  Run_Time <- Sys.time() - Start_Time
  Ver2.results <- list(jags.data = jags.data,
                       inits = inits.list,
                       JAGS_out = jm,
                       Run_Time = Run_Time,
                       Run_Date = Sys.Date())
  saveRDS(Ver2.results, 
          paste0("RData/JAGS_dailyN_GAM_k",
                 basis.dim, "_", family, "_Nmix_", x, "yr_v2.rds"))
  
} else {
  Ver2.results <- readRDS(paste0("RData/JAGS_dailyN_GAM_k", 
                                 basis.dim, "_", family, 
                                 "_Nmix_", x, "yr_v2.rds"))
  
}
```


Make some plots:

```{r}
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

# Extract estimated counts
# Daily.Est <- Ver2.results$JAGS_out$sims.list$Daily.Est
# sp <- Ver2.results$JAGS_out$sims.list$sp
# com <- Ver2.results$JAGS_out$sims.list$com
Corrected.Est <- Ver2.results$JAGS_out$sims.list$Corrected.Est

Corrected.Est.Summary <- data.frame(median = apply(Corrected.Est, 
                                                   MARGIN = 2, 
                                                   FUN = median),
                                    LCL = apply(Corrected.Est, 
                                                MARGIN = 2,
                                                FUN = quantile, 0.025),
                                    UCL = apply(Corrected.Est, 
                                                MARGIN = 2,
                                                FUN = quantile, 0.975),
                                    year = seasons)

ggplot(Corrected.Est.Summary) +
  geom_point(aes(x = year, y = median)) +
  geom_errorbar(aes(x = year, ymin = LCL, ymax = UCL))
```


```{r}

gam.fit <- Ver2.results$JAGS_out$sims.list$gam.fit

median.daily <- apply(gam.fit, MARGIN = 2:3, FUN = median) %>%
  data.frame() 
colnames(median.daily) <- seasons
median.daily %>% pivot_longer(cols = everything(),
                              names_to = "Season",
                              values_to = "Median") %>%
  arrange(Season) -> median.df

LCL.daily <- apply(gam.fit, MARGIN = 2:3, FUN = quantile, 0.025) %>%
  data.frame()
colnames(LCL.daily) <- seasons
LCL.daily %>% pivot_longer(cols = everything(),
                              names_to = "Season",
                              values_to = "LCL") %>%
  arrange(Season) -> LCL.df

UCL.daily <- apply(gam.fit, MARGIN = 2:3, FUN = quantile, 0.975) %>%
  data.frame()
colnames(UCL.daily) <- seasons
UCL.daily %>% pivot_longer(cols = everything(),
                              names_to = "Season",
                              values_to = "UCL") %>%
  arrange(Season) -> UCL.df

gam.daily <- data.frame(Season = median.df$Season,
                        median = median.df$Median,
                        LCL = LCL.df$LCL,
                        UCL = UCL.df$UCL,
                        days = rep(1:90, times = x))

# put together observed n -> total N per day
daily.effort <- daily.counts <- daily.N <- matrix(nrow = 90, ncol = x)
Watch.Length <- Ver2.results$jags.data$Watch.Length
day.mat <- Ver2.results$jags.data$day
ns <- rbind(Ver2.results$jags.data$n[,1,], matrix(NA, nrow = 2, ncol = x))

for (y in 1:x){
  for (j in 1:90){
    if (j == 1 | j == 90){
      daily.effort[j,y] <- 0
      daily.counts[j,y] <- 0
      daily.N[j,y] <- 0
    } else {
      daily.effort[j, y] <- sum(Watch.Length[day.mat[,y] == j, y], na.rm = T)
      daily.counts[j, y] <- sum(ns[day.mat[,y] == j, y], na.rm = T)
      daily.N[j, y] <- ceiling(daily.counts[j,y]/daily.effort[j,y])
      
    }
  }
}
  
daily.N.df <- data.frame(daily.N) 
colnames(daily.N.df) <- seasons
daily.N.df %>% pivot_longer(everything(),
                            names_to = "Season", 
                            values_to = "N") %>%  
  arrange(Season) -> daily.N.df.long

daily.N.df.long$days <- rep(1:90, times = x)

ggplot(data = gam.daily) + 
  geom_line(aes(x = days, y = median)) + 
  geom_ribbon(aes(x = days, 
                  ymin = LCL, 
                  ymax = UCL),
              fill = "orange", 
              alpha = 0.5) +
  geom_point(data = daily.N.df.long,
             aes(x = days, y = N)) +
  facet_wrap(vars(Season)) +
  xlab("Days since December 1") + 
  ylab("Whales per day")

```



