---
title: "R Notebook"
output: html_notebook
---

This notebook describes abundance estimation of gray whales using Durban's WinBUGS code and Ver2.0 data extraction method. Ver1.0 was developed by John Durban and updated by Josh Stewart. I created Ver2.0, which corrected some errors in Ver1.0. In this notebook, the new data from the 2021/2022 and 2022/2023 seasons were added to the previous data and run with WinBUGS. 

Set up libraries

```{r}
rm(list=ls())
library(R2jags)
library(abind)
library(R2WinBUGS)
library(tidyverse)
# PARAMETER NAME TRANSFERS:

# JOSH     JOHN
#    Spline 
# b.sp     b1 
# beta.sp  beta1
# X.sp     X1

WinBUGS.dir <- paste0(Sys.getenv("HOME"), "/WinBUGS14")

```

Get data prior to the 2021/2022 season:

```{r}
# this file contains all necessary inputs:
data.0 <- readRDS("RData/2006-2019_GC_Formatted_Data.RDS")

# output from Ver2.0 extraction
years <- c("2015", "2016", "2020", "2022", "2023")

#out.v2 <- lapply(years, FUN = function(x) readRDS(paste0("RData/V2.1_Aug2022/out_", x, "_Tomo_v2.rds")))

# 85 or 30
min.duration <- 85 #30

# Change file names accordingly:
Nhats.filename <- paste0("Data/abundance_", years[length(years)], "_", min.duration, "min.csv")
#Nhats.filename <- paste0("Data/abundance_", years[length(years)], "_", min.duration, "min_8weeks.csv")

# Analysis with minimum of 85 or 30 minutes sampling duration:
out.v2 <- lapply(years, FUN = function(x) readRDS(paste0("RData/V2.1_Mar2023/out_", x,
                                                         "_min", min.duration, "_Tomo_v2.rds")))

# just for 8 weeks in 2023
#out.v2 <- lapply(years, FUN = function(x) readRDS(paste0("RData/V2.1_Mar2023_8weeks/out_", x, 
#                                                         "_min", min.duration, "_Tomo_v2.rds")))

begin. <- lapply(out.v2, FUN = function(x) x$Final_Data$begin)
end. <- lapply(out.v2, FUN = function(x) x$Final_Data$end)

# Number of watch periods in each year's survey - before the 2019/2020 season
# plus the new ones
periods <-c(136, 135, 164, 178, 
            lapply(begin., FUN = function(x) length(x)) %>% unlist)
# Shorten data to first x years only to replicate 
# the analysis in Durban et al 2016:
x <- length(periods)

#out.file.name <- paste0("RData/WinBUGS_", x, "yr_v2_min", min.duration, ".rds")
#out.file.name <- paste0("RData/WinBUGS_", x, "yr_v2_min", min.duration, "_8weeks_2023.rds")
out.file.name <- paste0("RData/WinBUGS_", x, "yr_v2_min", min.duration, "_2023.rds")

Watch.Length. <- list()

for (k in 1:length(begin.)){
  Watch.Length.[[k]] <- end.[[k]] - begin.[[k]]
}

# 
# #whale counts
n <- labelled::remove_attributes(data.0$n, "dimnames")
n <- n[,,1:4]  # take the first four years
n <- abind(n, array(0, replace(dim(n), 1, max(periods) - nrow(n))), along = 1)

# the u data is whether there were observers on watch. 
# 0 counts are often associated with years/shifts with 
# no second observer. So if u=0, it will fix observation probability at 0
# the second column for each year is for the second station - not the second
# observer.
u <- data.0$u[,,1:4]
u <- abind(u, 
           array(0, replace(dim(u), 1, max(periods) - nrow(u))), along = 1)

# #visibility
vs. <- lapply(out.v2, FUN = function(x) x$Final_Data$vs)
vs <- rbind(data.0$vs[,1:4], 
                  array(NA, dim = c(max(periods) - nrow(data.0$vs), 4))) %>%
    labelled::remove_attributes("dimnames")

# Beaufort
bf. <- lapply(out.v2, FUN = function(x) x$Final_Data$bf)
bf <- rbind(data.0$bf[,1:4], 
                  array(NA, dim = c(max(periods) - nrow(data.0$bf), 4)))%>%
    labelled::remove_attributes("dimnames")

# observers
# need to convert observer initials into numbers for all years
obs.list <- read.csv(file = "Data/ObserverList2023.csv")

obs <- data.0$obs[,,1:4]
obs <- abind(obs, 
           array(36, replace(dim(obs), 1, max(periods) - nrow(obs))), along = 1)

Watch.Length <- rbind(data.0$Watch.Length[,1:4], 
                      array(NA, dim = c(max(periods) - 
                                          nrow(data.0$Watch.Length), 4)))%>%
    labelled::remove_attributes("dimnames")

day <- rbind(data.0$day[,1:4], 
                   array(NA, dim = c(max(periods) - nrow(data.0$day), 4)))%>%
    labelled::remove_attributes("dimnames")

k <- 1
for (k in 1:length(begin.)){
  n <- abind(n, 
             cbind(c(out.v2[[k]]$Final_Data$n, 
                     rep(0, times = (dim(n)[1] - length(out.v2[[k]]$Final_Data$n)))), 
                   rep(0, times = dim(n)[1]))) %>%
    labelled::remove_attributes("dimnames")
  
  u <- abind(u, 
             cbind(c(rep(1, times = length(out.v2[[k]]$Final_Data$n)), 
                     rep(0, times = (dim(u)[1] - length(out.v2[[k]]$Final_Data$n)))), 
                   rep(0, times = dim(u)[1])))
  
  vs <- cbind(vs, c(out.v2[[k]]$Final_Data$vs, 
                    rep(NA, times = max(periods - length(out.v2[[k]]$Final_Data$vs)))))  %>%
    labelled::remove_attributes("dimnames")
  
  
  bf <- cbind(bf, c(out.v2[[k]]$Final_Data$bf,
                    rep(NA, times = max(periods - length(out.v2[[k]]$Final_Data$bf)))))  %>%
    labelled::remove_attributes("dimnames")
  
  obs.year <- data.frame(obs = out.v2[[k]]$Final_Data$obs) %>% 
    left_join(obs.list, by = "obs")
  
  obs <- abind(obs, 
               cbind(c(obs.year$ID, 
                       rep(36, times = (max(periods) - length(obs.year$ID)))), 
                     rep(36, times = max(periods))))

  Watch.Length <- cbind(Watch.Length,
                        c(Watch.Length.[[k]], 
                          rep(NA, 
                              times = max(periods) - length(Watch.Length.[[k]])))) %>%
    labelled::remove_attributes("dimnames")
  
  day <- cbind(day, 
               c(floor(begin.[[k]]), 
               rep(NA, times = max(periods) - length(begin.[[k]])))) %>%
    labelled::remove_attributes("dimnames")
  
}


#we're going to make N a partially observed data object with anchor points at day 1 and 90
# TE: I don't know how these numbers were created... they are generally 2x n (not all)
# N_inits <- as.matrix(read.table("Data/Initial Values/N_inits.txt",
#                                 header=T))

N_inits1 <- n[, 1,] * 2 + 2
N_inits2 <- n[, 2,] * 2 + 2 
            
N_inits <- N_inits1
N_inits[N_inits1 < N_inits2] <- N_inits2[N_inits1 < N_inits2]

N_inits <- rbind(N_inits,
                 matrix(data = NA, nrow = 2, ncol = length(periods)))

for (k in 1:length(periods)){
  N_inits[(periods[k]+1):nrow(N_inits), k] <- NA  
}

#The 'data' has to be the inverse of the inits, 
# with NAs for all of the estimated Ns, and 0s for the days 1 and 90
N <- matrix(NA, nrow=max(periods)+2, ncol=length(periods)) 

for(i in 1:length(periods)){
  N[(periods[i]+1):(periods[i]+2),i] <- 0 #True number of whales passing fixed at 0 for day 1 and 90
}
```

Set up BUGS:

```{r}
# Watch.Length <- cbind(rbind(data.0$Watch.Length[,1:4], 
#                             array(NA, dim = c(max(periods) - 
#                                                 nrow(data.0$Watch.Length), 4))),
#                       c(Watch.Length.2015, 
#                         rep(NA, 
#                             times = max(periods) -                             
#                               length(Watch.Length.2015))),
#                       c(Watch.Length.2016, 
#                         rep(NA, 
#                             times = max(periods) -                              
#                               length(Watch.Length.2016))),
#                       c(Watch.Length.2020, 
#                         rep(NA, 
#                             times = max(periods) -                              
#                               length(Watch.Length.2020))),
#                       c(Watch.Length.2022, 
#                         rep(NA, 
#                             times = max(periods) -                 
#                               length(Watch.Length.2022)))) %>%
#   labelled::remove_attributes("dimnames")

#specify survey days associated with each watch period
# day <- cbind(rbind(data.0$day[,1:4], 
#                    array(NA, dim = c(max(periods) - nrow(data.0$day), 4))),
#              c(floor(begin.2015), 
#                rep(NA, times = max(periods) - length(begin.2015))),
#              c(floor(begin.2016), 
#                rep(NA, times = max(periods) - length(begin.2016))),
#              c(floor(begin.2020), 
#                rep(NA, times = max(periods) - length(begin.2020))),
#              c(floor(begin.2022), 
#                rep(NA, times = max(periods) - length(begin.2022)))) %>%
#   labelled::remove_attributes("dimnames")

end <- Watch.Length + day

#t <- round((begin+end)/2)
t <- round((day+end)/2)

# #Add a couple of extra rows of NAs to the end of the day index reference to match up with the fixed 0s in N (above), assigning them to days 1 and 90
day <- rbind(as.matrix(day),
             matrix(NA, nrow=2, ncol=length(periods)))
#
#
for(i in 1:length(periods)){ #Set the anchor points: days 1 and 90
  day[(periods[i]+1):(periods[i]+2),i] <- c(1,90)
}

t <- rbind(as.matrix(t),
           matrix(NA,nrow=2,ncol=length(periods)))
for(i in 1:length(periods)){ #Set the anchor points: days 1 and 90
  t[(periods[i]+1):(periods[i]+2),i] <- c(1,90)
}

#Place 36s for 'no observer' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
#this will force it to the mean observation probability with no observer effect

Watch.Length <- rbind(as.matrix(Watch.Length),
                      matrix(NA, nrow=2, ncol=length(periods)))

for(i in 1:length(periods)){
  Watch.Length[(periods[i]+1):(periods[i]+2),i] <- 1
}

#### Two separate models each fit to data, 
# and then a third replicate of data to select
# 'best fit' (same structure as Durban et al)


# n <- n[,,1:x]
# obs <- obs[,,1:x]
# periods <- periods[1:x]
# u <- u[,,1:x]
# vs <- vs[,1:x]
# bf <- bf[,1:x]
# day <-day[,1:x]
# t <- t[,1:x]
# N <- N[,1:x]
# N_inits <- N_inits[,1:x]
# Watch.Length <- Watch.Length[,1:x]

BUGS.data <- list(n = n[1:max(periods[1:x]),,1:x],
                  n.com = n[1:max(periods[1:x]),,1:x],
                  n.sp = n[1:max(periods[1:x]),,1:x],
                  n.station = dim(n[1:max(periods[1:x]),,1:x])[2],
                  n.year = dim(n[1:max(periods[1:x]),,1:x])[3],
                  n.obs = max(obs[1:max(periods[1:x]),,1:x]),
                  periods = periods[1:x],
                  obs = obs[1:max(periods[1:x]),,1:x],
                  #Watch.Length = 0.0625,
                  u = u[1:max(periods[1:x]),,1:x],
                  vs = vs[1:max(periods[1:x]),1:x],
                  bf = bf[1:max(periods[1:x]),1:x],
                  #day=day,
                  day = t[1:(max(periods[1:x])+2),1:x],
                  N = N[,1:x],
                  N.com = N[,1:x],
                  N.sp = N[,1:x],
                  knot = c(-1.46,-1.26,-1.02,-0.78,
                         -0.58,-0.34,-0.10,0.10,
                         0.34,0.57,0.78,1.02,1.26,1.46),
                  n.knots=14,
                  #begin=begin,
                  #end=end,
                  Watch.Length=Watch.Length[1:(max(periods[1:x])+2), 1:x])

BUGS.inits <- function() list(mean.prob = 0.5,
                              BF.Fixed = 0,
                              VS.Fixed = 0,
                              mean.prob.sp = 0.5,
                              BF.Fixed.sp = 0,
                              VS.Fixed.sp = 0,
                              mean.prob.com = 0.5,
                              BF.Fixed.com = 0,
                              VS.Fixed.com = 0,
                              mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
                              beta.sigma = c(1,1,1),#beta.sigma = c(7,7,7),
                              BF.Switch = 1,
                              VS.Switch = 1,
                              OBS.Switch = 1,
                              sigma.Obs = 1,
                              BF.Switch.sp = 1,
                              VS.Switch.sp = 1,
                              OBS.Switch.sp = 1,
                              sigma.Obs.sp = 1,
                              BF.Switch.com = 1,
                              VS.Switch.com = 1,
                              OBS.Switch.com = 1,
                              sigma.Obs.com = 1,
                              N = N_inits,
                              N.com = N_inits,
                              N.sp = N_inits,
                              #z = matrix(1,nrow=90,ncol=6),
                              beta.sp = array(data=0, dim=c(2,x)),
                              sd.b.sp = rep(1, times = x), #c(1,1,1,1,1,1),
                              z = matrix(1, nrow=90, ncol= x))

#### To run 2006-2019 data, load .RDS object and name it BUGS.data, 
# then update initial values:
# Files are 2006-2019_GC_Formatted_Data and 2006-2019_GC_N_inits

# WinBUGS gives errors when N inits are set to 0. 
# Try setting them to 1 instead (seems to work):
# But this makes no sense to have it here... 2022-10-05
#N_inits[which(N_inits==0,arr.ind = T)] <- 1

parameters <- c("lambda","OBS.RF","OBS.Switch",
                "BF.Switch","BF.Fixed","VS.Switch",
                "VS.Fixed","mean.prob","mean.prob.com",
                "mean.prob.sp","BF.Fixed.com",
                "BF.Fixed.sp","VS.Fixed.com",
                "VS.Fixed.sp",
                "Corrected.Est","Raw.Est","z",
                "com","sp","Daily.Est","mean.beta",
                "beta.sigma","beta","beta.sp","b.sp","sd.b.sp")

# Good to run with 10k/6k as a practice run to see if WinBUGS completes computation.
# Then, increase iterations/burnin to 100k/60k. 

MCMCparams <- list(n.iter = 100,
                   n.thin = 5,
                   n.burnin = 60,
                   n.chains = 5)

# MCMCparams <- list(n.iter = 100000,
#                    n.thin = 80,
#                    n.burnin = 60000,
#                    n.chains = 5)

```

Run WinBUGS if it hasn't been run..

```{r}
#out.file.name <- paste0("RData/WinBUGS_", x, "yr_v2.rds")

if (!file.exists(out.file.name)){
  
  #Run time: 
  Start_Time<-Sys.time()
  
  BUGS_out <- bugs(data = BUGS.data,
                  inits = BUGS.inits,
                  parameters = parameters,
                  model.file="GW_Nmix_Orig.bugs",
                  n.chains = MCMCparams$n.chains,
                  n.iter = MCMCparams$n.iter, 
                  n.burnin = MCMCparams$n.burnin, 
                  n.thin = MCMCparams$n.thin,
                  debug=F,
                  bugs.directory = WinBUGS.dir,
                  DIC = FALSE)
  
  # 2023-03-03 (problem solved but leavee the comments below for future reference)
  # ERROR: NIL dereference (read). According to the user manual, this error may
  # happen "at compilation in some circumstances when an inappropriate
  # transformation is made, for example an array into a scalar." 
  # (https://www.mrc-bsu.cam.ac.uk/wp-content/uploads/manual14.pdf)
  
  # DIC problems: Surprisingly, sometimes when getting a trap (including one with the very
  # informative title “NIL dereference (read)”), setting the argument DIC = FALSE in the bugs()
  # function has helped. (https://www.mbr-pwrc.usgs.gov/software/kerybook/AppendixA_list_of_WinBUGS_tricks.pdf)
  
  # I only changed the effort (Watch.Length) for this analysis using 30 minutes as the 
  # minimum observation duration. That changed the size of data arrays, which shouldn't be an issue. 
  
  # It turned out a new observer (JSD) was introduced when a shorter minimum was used to filter
  # observation period. "JSD" was only found in 2020 (200109_080043). A strange thing happened for that day.
  # During the 0800 shift, JSD/JWG changed to SJC/JDS on 0900, then the shift continued until 0930. The new
  # cutoff time (30 min) picked up the first (1 hr) and the second (30 min) as separate observation periods.

  Run_Time <- Sys.time() - Start_Time
  Ver2.results <-list(BUGS.data = BUGS.data,
                      N_inits = N_inits,
                      BUGS_out = BUGS_out,
                      Run_Time = Run_Time,
                      Sys.info = Sys.info())
  
  saveRDS(Ver2.results, file = out.file.name)
  
} else {
  Ver2.results <- readRDS(out.file.name)
}

# The following line is for saving the data list and inits list. This was done to check why simulation
# data were resulting in error - "undefined real result" 
# (ran with n.iter = 1000, n.thin = 8, n.burnin = 600, n.chains = 3) as well as n.iter = 10000,
# n.thin = 80, n.burnin = 6000, n.chains = 5
# data.worked <- list(BUGS.data = BUGS.data,
#                     BUGS.inits = BUGS.inits,
#                     N_inits = N_inits,
#                     x = x,
#                     parameters = parameters,
#                     MCMC.params = MCMCparams,
#                     model.file = "GW_Nmix_Orig.bugs",
#                     WinBUGS.dir = WinBUGS.dir)
# 
# saveRDS(file = paste0("RData/BUGS_data_runs_", Sys.Date(), ".rds"), 
#         data.worked)

# clear the workspace, bring in the data that worked.
#rm(list = ls())
#
# Runs from the saved list? Yes.
# data.worked <- readRDS(paste0("RData/BUGS_data_runs_2022-10-21.rds"))
# N_inits <- data.worked$N_inits
# x <- data.worked$x
# bm <- bugs(data = data.worked$BUGS.data,
#              inits = data.worked$BUGS.inits,
#              parameters = data.worked$parameters,
#              model.file = data.worked$model.file,
#              n.chains = data.worked$MCMC.params$n.chains,
#              n.iter = data.worked$MCMC.params$n.iter,
#              n.burnin = data.worked$MCMC.params$n.burnin,
#              n.thin = data.worked$MCMC.params$n.thin,
#              debug=T,
#              bugs.directory = data.worked$WinBUGS.dir)




```
array index is greater than array upper bound for Watch.Length
inits(1,C:/Users/TOMO~1.EGU/AppData/Local/Temp/Rtmpy436t0/init

Make some plots:

```{r}
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022",
             "2022/2023")

# Extract estimated counts
Daily.Est <- Ver2.results$BUGS_out$sims.list$Daily.Est
sp <- Ver2.results$BUGS_out$sims.list$sp
com <- Ver2.results$BUGS_out$sims.list$com
Corrected.Est <- Ver2.results$BUGS_out$sims.list$Corrected.Est

# Each one of them is (# samples) x (90 days) x (# years)
# To plot them using ggplot's facet, I need to convert
# these into 2D dataframes of statistics (upper and lower 
# CIs, median, etc.)
# Daily.Est.list <- sp.list <- com.list <- vector(mode = "list", 
#                                                 length = dim(Daily.Est)[3])
# 
# Daily.Est.UCIs <- Daily.Est.LCIs <- vector(mode = "list",
#                                            length = dim(Daily.Est)[3])

stats.list <- vector(mode = "list",
                     length = dim(Daily.Est)[3])

for (k in 1:dim(Daily.Est)[3]){
  # Daily.Est.list[[k]] <- Daily.Est[,,k]
  # Daily.Est.UCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.975)
  # Daily.Est.LCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.275)
  # 
  # sp.list[[k]] <- sp[,,k]
  # com.list[[k]] <- com[,,k]
  
  stats.list[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est[,,k], 2,
                                                         median),
                                Daily.Est.LCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.275),
                                Daily.Est.UCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.975),
                                sp.median = apply(exp(sp[,,k]), 2,
                                                  median),
                                sp.LCL = apply(exp(sp[,,k]), 2,
                                               quantile,0.025),
                                sp.UCL = apply(exp(sp[,,k]), 2,
                                               quantile,0.975),
                                com.median = apply(exp(com[,,k]), 2,
                                                   median),
                                com.LCL = apply(exp(com[,,k]), 2,
                                                quantile,0.025),
                                com.UCL = apply(exp(com[,,k]), 2,
                                                quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est)[2],
                                year = seasons[k])
}

all.stats <- do.call("rbind", stats.list) %>% group_by(year)

ggplot(data = all.stats) + 
  geom_line(aes(x = days, y = sp.median)) + 
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", 
              alpha = 0.5) +
  geom_line(aes(x = days, y = com.median),
            color = "red") +
  facet_wrap(vars(year)) +
  xlab("Days since December 1") + 
  ylab("Whales per day")

```



```{r}

#Com vs Sp
ggplot(data = all.stats) +
  geom_line(aes(x = days, y = sp.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day (spline)")



```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = com.LCL, 
                  ymax = com.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day (Normal)")

```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = Daily.Est.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = Daily.Est.LCL, 
                  ymax = Daily.Est.UCL),
              fill = "orange", 
              alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day")


```

Total abundance
```{r}
abundance.df <- data.frame(Season = seasons,
                           total.mean = apply(Corrected.Est,
                                              FUN = mean,
                                              MARGIN = 2),
                           total.CV = apply(Corrected.Est,
                                            FUN = function(x) 100*sqrt(var(x))/mean(x),
                                            MARGIN = 2),
                           total.median = apply(Corrected.Est, 
                                                FUN = median, 
                                                MARGIN = 2),
                           total.LCL = apply(Corrected.Est, 
                                             MARGIN = 2, 
                                             FUN = quantile, 0.025),
                           total.UCL = apply(Corrected.Est, 
                                             MARGIN = 2, 
                                             FUN = quantile, 0.975))


ggplot(data = abundance.df) + 
  geom_point(aes(x = Season, y = total.median)) + 
  geom_errorbar(aes(x = Season, ymin = total.LCL, ymax = total.UCL))

#median(apply(exp(sp[,,4]),1,sum))


write_csv(abundance.df, file = Nhats.filename)

```

```{r}
ggplot(data = abundance.df) + 
  geom_point(aes(x = Season, y = total.mean)) + 
  geom_errorbar(aes(x = Season, ymin = total.LCL, ymax = total.UCL))

#ggsave(filename = "figures/abundance_9yrs.png", device = "png", dpi = 600)
```
