---
title: "R Notebook"
output: html_notebook
---

This notebook describes abundance estimation of gray whales using Durban's WinBUGS code and Ver1.0 data extraction method. Ver1.0 was developed by John Durban and updated by Josh Stewart. In this notebook, the new data from the 2021/2022 season are added to the previous data and run with WinBUGS. 

Set up libraries

```{r}
rm(list=ls())
library(R2jags)
library(abind)
library(R2WinBUGS)
library(tidyverse)
# PARAMETER NAME TRANSFERS:

# JOSH     JOHN
#    Spline 
# b.sp     b1 
# beta.sp  beta1
# X.sp     X1


```

Get data prior to the 2021/2022 season:

```{r}
# this file contains all necessary inputs:
data.0 <- readRDS("RData/2006-2019_GC_Formatted_Data.RDS")

# output from Ver1.0 (or Josh's) extraction
out.V1.2020 <- readRDS("RData/out_2020_Joshs.rds")
out.V1.2022 <- readRDS("RData/out_2022_Joshs.rds")

begin.2020 <- data.frame(begin.2020 = out.V1.2020$FinalData$begin) %>%
  rownames_to_column(var = "rowname") %>%
  mutate(rowname = as.integer(rowname))

begin.2022 <- data.frame(begin.2022 = out.V1.2022$FinalData$begin) %>%
  rownames_to_column(var = "rowname") %>%
  mutate(rowname = as.integer(rowname))

end.2020 <- data.frame(begin.2020 = out.V1.2020$FinalData$end) %>%
  rownames_to_column(var = "rowname") %>%
  mutate(rowname = as.integer(rowname))

end.2022 <- data.frame(begin.2022 = out.V1.2022$FinalData$end) %>%
  rownames_to_column(var = "rowname") %>%
  mutate(rowname = as.integer(rowname))

# Number of watch periods in each year's survey - before the 2019/2020 season
# plus the new ones
periods <-c(136, 135, 164, 178, 179, 151, nrow(begin.2020), nrow(begin.2022))

# begin and end need to be brought in - only 6 columns. So, the 2019/2020 season
# needs to be added. 
# #Watch start times, as fraction of a day
# TE changed begin.txt to begin_a.txt; modified headers from begin[,i] to begin.YEAR
# and removed END at the end of the file. 2022-04-05
begin <- read.table("Data/begin_a.txt",
                    header=T) %>%
  rownames_to_column(var = "rowname") %>%
  mutate(rowname = as.integer(rowname))

# Use merge to combine different lengths dataframes
begin <- merge(begin,
               begin.2020, by = "rowname",
               all = TRUE)

begin <- merge(begin, 
               begin.2022, by = "rowname",
               all = TRUE) %>%
  select(-rowname)

# watch end times - same modificaitons made here as begin.
end <- read.table("Data/end_a.txt",
                    header=T) %>%
  rownames_to_column(var = "rowname") %>%
  mutate(rowname = as.integer(rowname))

end <- merge(end,
             end.2020, by = "rowname",
             all = TRUE)

end <- merge(end, 
             end.2022, by = "rowname",
             all = TRUE) %>% 
  select(-rowname)


# 
# #whale counts
n <- data.0$n
# this has 7 years worth. 
n.2020 <- n[,,7]

# those from Ver1.0 extraction:
n.2020a <- out.V1.2020$FinalData$n # slightly different but mostly the same...

n.2022a <- out.V1.2022$FinalData$n

# use abind for combining multi-dimensional arrays
n <- abind(n, cbind(c(n.2022a, rep(0, times = (dim(n)[1] - length(n.2022a)))), 
                      rep(0, times = dim(n)[1])))

# n <- as.matrix(read.table("Data/n.txt",
#                           header=T,
#                           nrows = max(periods)))
# 
# dim(n) <- c(179,2,6) #convert this back to a 3D array
# #n <- abind(n,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the sightings array (this is for the day 1 and day 90 zero-whale anchor points)
# 
# 

# the u data is whether there were observers on watch. 
# 0 counts are often associated with years/shifts with 
# no second observer. So if u=0, it will fix observation probability at 0
# the second column for each year is for the second station - not the second
# observer. 
u <- abind(data.0$u, 
           cbind(c(rep(1, times = length(n.2022a)), 
                   rep(0, times = (dim(n)[1] - length(n.2022a)))), 
                 rep(0, times = dim(n)[1])))

# u <- as.matrix(read.table("Data/u.txt",
#                           header=T,
#                           nrows = max(periods)))
# dim(u) <- c(179,2,6) #convert this back to a 3D array
# #u <- abind(u,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the effort on/off array
# #for(i in 1:length(periods)){ #Place 1's for 'effort on' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
# #  u[(periods[i]+1):(periods[i]+2),,i] <- 1
# #}
# 
# 
# #visibility
vs <- cbind(data.0$vs,
            c(out.V1.2022$FinalData$vs, 
              rep(NA, 
                  times = nrow(data.0$vs) - length(out.V1.2022$FinalData$vs))))
# vs <- as.matrix(read.table("Data/vs.txt",
#                            header=T,
#                           nrows = max(periods)))
# #vs <- rbind(vs,matrix(NA,nrow=2,ncol=6)) #Add two trailing NAs (this is for the day 1 and day 90 zero-whale anchor points)
# 
# #beaufort
bf <- cbind(data.0$bf,
            c(out.V1.2022$FinalData$vs, 
              rep(NA, 
                  times = nrow(data.0$vs) - length(out.V1.2022$FinalData$vs))))
# bf <- as.matrix(read.table("Data/bf.txt",
#                            header=T,
#                           nrows = max(periods)))
# #bf <- rbind(bf,matrix(NA,nrow=2,ncol=6)) #Add two trailing NAs (this is for the day 1 and day 90 zero-whale anchor points)
# 
# #observer numbers
obs <- data.0$obs

# need to convert observer initials into numbers for 2022
obs.2022 <- data.frame(obs = out.V1.2022$FinalData$obs)
obs.list <- read.csv(file = "Data/Observer list 2022.csv")

obs.2022 %>% left_join(obs.list, by = "obs") -> obs.2022.ID

obs <- abind(data.0$obs, 
             cbind(c(obs.2022.ID$ID, 
                     rep(36, times = (dim(data.0$obs)[1] - nrow(obs.2022.ID)))),
                   rep(36, times = dim(data.0$obs)[1])))

# obs <- as.matrix(read.table("Data/obs.txt",
#                             header=T,
#                           nrows = max(periods)))
# dim(obs) <- c(179,2,6) #convert this back to a 3D array
# #obs <- abind(obs,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the effort on/off array
# #for(i in 1:length(periods)){ #Place 36s for 'no observer' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
# #  obs[(periods[i]+1):(periods[i]+2),,i] <- 36 #this will force it to the mean observation probability with no observer effect
# #}

#we're going to make N a partially observed data object with anchor points at day 1 and 90
# TE: I don't know how these numbers were created... 
# N_inits <- as.matrix(read.table("Data/Initial Values/N_inits.txt",
#                                 header=T))

N_inits <- matrix(data = NA, nrow = (max(periods)+2), ncol = length(periods))
for(i in 1:length(periods)){
  N_inits[(periods[i]+1):(periods[i]+2),i] <- NA 
}

#The 'data' has to be the inverse of the inits, 
# with NAs for all of the estimated Ns, and 0s for the days 1 and 90
N <- matrix(NA, nrow=max(periods)+2, ncol=length(periods)) 

for(i in 1:length(periods)){
  N[(periods[i]+1):(periods[i]+2),i] <- 0 #True number of whales passing fixed at 0 for day 1 and 90
}
```

Set up BUGS:

```{r}

#specify survey days associated with each watch period
day <- floor(begin)
t <- round((begin+end)/2)
#Add a couple of extra rows of NAs to the end of the day index reference to match up with the fixed 0s in N (above), assigning them to days 1 and 90
day <- rbind(as.matrix(day), matrix(NA, nrow=2, ncol=length(periods)))
for(i in 1:length(periods)){ #Set the anchor points: days 1 and 90
  day[(periods[i]+1):(periods[i]+2),i] <- c(1,90)
}

t <- rbind(as.matrix(t),matrix(NA,nrow=2,ncol=length(periods)))
for(i in 1:length(periods)){ #Set the anchor points: days 1 and 90
  t[(periods[i]+1):(periods[i]+2),i] <- c(1,90)
}

#### Two separate models each fit to data, 
# and then a third replicate of data to select
# 'best fit' (same structure as Durban et al)

# Shorten data to first 4 years only to replicate 
# the analysis in Durban et al 2016:
# n.short <- n[,,1:4]
# obs.short <- obs[,,1:4]
# periods.short <- periods[1:4]
# u.short <- u[,,1:4]
# vs.short <- vs[,1:4]
# bf.short <- bf[,1:4]
# day.short <-day[,1:4]
# t.short <- t[,1:4]
# N.short <- N[,1:4]
# N_inits.short <- N_inits[,1:4]

Watch.Length <- rbind(as.matrix(end),
                      matrix(NA,nrow=2,ncol=length(periods))) - 
  rbind(as.matrix(begin), matrix(NA,nrow=2,ncol=length(periods)))
for(i in 1:length(periods)){ #Place 36s for 'no observer' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
  Watch.Length[(periods[i]+1):(periods[i]+2),i] <- 1 #this will force it to the mean observation probability with no observer effect
}

jags.data <- list(n = n,
                  n.com = n,
                  n.sp = n,
                  n.station = dim(n)[2],
                  n.year = dim(n)[3],
                  n.obs = max(obs),
                  periods = periods,
                  obs = obs,
                  #Watch.Length = 0.0625,
                  u = u,
                  vs = vs,
                  bf = bf,
                  #day=day,
                  day = t,
                  N = N,
                  N.com = N,
                  N.sp = N,
                  knot = c(-1.46,-1.26,-1.02,-0.78,
                         -0.58,-0.34,-0.10,0.10,
                         0.34,0.57,0.78,1.02,1.26,1.46),
                  n.knots=14,
                  #begin=begin,
                  #end=end,
                  Watch.Length=Watch.Length)

jags.inits <- function() list(mean.prob = 0.5,
                              BF.Fixed = 0,
                              VS.Fixed = 0,
                              mean.prob.sp = 0.5,
                              BF.Fixed.sp = 0,
                              VS.Fixed.sp = 0,
                              mean.prob.com = 0.5,
                              BF.Fixed.com = 0,
                              VS.Fixed.com = 0,
                              mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
                              beta.sigma = c(1,1,1),#beta.sigma = c(7,7,7),
                              BF.Switch = 1,
                              VS.Switch = 1,
                              OBS.Switch = 1,
                              sigma.Obs = 1,
                              BF.Switch.sp = 1,
                              VS.Switch.sp = 1,
                              OBS.Switch.sp = 1,
                              sigma.Obs.sp = 1,
                              BF.Switch.com = 1,
                              VS.Switch.com = 1,
                              OBS.Switch.com = 1,
                              sigma.Obs.com = 1,
                              N = N_inits,
                              N.com = N_inits,
                              N.sp = N_inits,
                              #z = matrix(1,nrow=90,ncol=6),
                              beta.sp = array(data=0,dim=c(2,length(periods))),
                              sd.b.sp = rep(1, times = length(periods)), #c(1,1,1,1,1,1),
                              z=matrix(1,nrow=90,ncol=length(periods)))

#### To run 2006-2019 data, load .RDS object and name it jags.data, 
# then update intitial values:
# Files are 2006-2019_GC_Formatted_Data and 2006-2019_GC_N_inits

# WinBUGS gives errors when N inits are set to 0. 
# Try setting them to 1 instead (seems to work):
N_inits[which(N_inits==0,arr.ind = T)] <- 1

parameters <- c("lambda","OBS.RF","OBS.Switch",
                "BF.Switch","BF.Fixed","VS.Switch",
                "VS.Fixed","mean.prob","mean.prob.com",
                "mean.prob.sp","BF.Fixed.com",
                "BF.Fixed.sp","VS.Fixed.com",
                "VS.Fixed.sp",
                "Corrected.Est","Raw.Est","z",
                "com","sp","Daily.Est","mean.beta",
                "beta.sigma","beta","beta.sp","b.sp","sd.b.sp")

ni <- 100000
nt <- 80
nb <- 60000
nc <- 3

```

Run WinBUGS if it hasn't been run..

```{r}
if (!file.exists("RData/GW BUGS 8yr 100k.RData")){
  
  #Run time: 
  Start_Time<-Sys.time()
  
  GW_Nmix <- bugs(data = jags.data,
                  inits = jags.inits,
                  parameters = parameters,
                  model.file="GW_Nmix_Orig.bugs",
                  n.chains = nc,
                  n.iter = ni, 
                  n.burnin = nb, 
                  n.thin = nt,
                  debug=T,
                  bugs.directory = "C:/Users/tomo.eguchi/Documents/WinBUGS14")
  
  Run_Time <- Sys.time() - Start_Time
  save.image("RData/GW BUGS 8yr 100k.RData")
  
} else {
  load("RData/GW BUGS 8yr 100k.RData")
}
```


Make some plots:

```{r}
# Extract estimated counts
Daily.Est <- GW_Nmix$sims.list$Daily.Est
sp <- GW_Nmix$sims.list$sp
com <- GW_Nmix$sims.list$com

# Each one of them is (# samples) x (90 days) x (# years)
# To plot them using ggplot's facet, I need to convert
# these into 2D dataframes of statistics (upper and lower 
# CIs, median, etc.)
# Daily.Est.list <- sp.list <- com.list <- vector(mode = "list", 
#                                                 length = dim(Daily.Est)[3])
# 
# Daily.Est.UCIs <- Daily.Est.LCIs <- vector(mode = "list",
#                                            length = dim(Daily.Est)[3])

stats.list <- vector(mode = "list",
                     length = dim(Daily.Est)[3])

for (k in 1:dim(Daily.Est)[3]){
  # Daily.Est.list[[k]] <- Daily.Est[,,k]
  # Daily.Est.UCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.975)
  # Daily.Est.LCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.275)
  # 
  # sp.list[[k]] <- sp[,,k]
  # com.list[[k]] <- com[,,k]
  
  stats.list[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est[,,k], 2,
                                                         median),
                                Daily.Est.LCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.275),
                                Daily.Est.UCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.975),
                                sp.median = apply(exp(sp[,,k]), 2,
                                                  median),
                                sp.LCL = apply(exp(sp[,,k]), 2,
                                               quantile,0.025),
                                sp.UCL = apply(exp(sp[,,k]), 2,
                                               quantile,0.975),
                                com.median = apply(exp(com[,,k]), 2,
                                                   median),
                                com.LCL = apply(exp(com[,,k]), 2,
                                                quantile,0.025),
                                com.UCL = apply(exp(com[,,k]), 2,
                                                quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est)[2],
                                year = k)
}

all.stats <- do.call("rbind", stats.list) %>% group_by(year)

ggplot(data = all.stats) + 
  geom_line(aes(x = days, y = sp.median)) + 
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", 
              alpha = 0.5) +
  geom_line(aes(x = days, y = com.median),
            color = "red") +
  facet_wrap(vars(year)) +
  xlab("Days since December 1") + 
  ylab("Whales per day")

```



```{r}

#Com vs Sp
ggplot(data = all.stats) +
  geom_line(aes(x = days, y = sp.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day")



```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = com.LCL, 
                  ymax = com.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day")

```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = Daily.Est.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = Daily.Est.LCL, 
                  ymax = Daily.Est.UCL),
              fill = "orange", 
              alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day")


```

Total abundance
```{r}
abundance.list <- vector(mode = "list", length = dim(sp)[3])
for (y in 1:dim(sp)[3]){
  abundance.list[[y]] <- data.frame(median = median(apply(exp(sp[,,y]), 1, sum)),
                                    LCL = quantile(apply(exp(sp[,,y]), 1, sum), 0.025),
                                    UCL = quantile(apply(exp(sp[,,y]), 1, sum), 0.975),
                                    year = y)
  
}

abundance.df <- do.call("rbind", abundance.list)

ggplot(data = abundance.df) + 
  geom_line(aes(x = year, y = median)) + 
  geom_errorbar(aes(x = year, ymin = LCL, ymax = UCL))

#median(apply(exp(sp[,,4]),1,sum))

```

