---
title: "R Notebook"
output: html_notebook
---

This notebook describes abundance estimation of gray whales using Durban's WinBUGS code and Ver1.0 data extraction method. Ver1.0 was developed by John Durban and updated by Josh Stewart. In this notebook, the new data from the 2021/2022 season are added to the previous data and run with WinBUGS. 

Set up libraries

```{r}
rm(list=ls())
library(R2jags)
library(abind)
library(R2WinBUGS)
library(tidyverse)
# PARAMETER NAME TRANSFERS:

# JOSH     JOHN
#    Spline 
# b.sp     b1 
# beta.sp  beta1
# X.sp     X1

WinBUGS.dir <- paste0(Sys.getenv("HOME"), "/WinBUGS14")

```

Get data prior to the 2021/2022 season:

```{r}
# this file contains all necessary inputs:
data.0 <- readRDS("RData/2006-2019_GC_Formatted_Data.RDS")

# output from Ver1.0 (or Josh's) extraction
out.V2.2015 <- readRDS("RData/out_2015_Tomo_v2.rds")
out.V2.2016 <- readRDS("RData/out_2016_Tomo_v2.rds")
out.V2.2020 <- readRDS("RData/out_2020_Tomo_v2.rds")
out.V2.2022 <- readRDS("RData/out_2022_Tomo_v2.rds")

begin.2015 <- out.V2.2015$FinalData$begin
begin.2016 <- out.V2.2016$FinalData$begin
begin.2020 <- out.V2.2020$FinalData$begin
begin.2022 <- out.V2.2022$FinalData$begin

end.2015 <- out.V2.2015$FinalData$end
end.2016 <- out.V2.2016$FinalData$end
end.2020 <- out.V2.2020$FinalData$end
end.2022 <- out.V2.2022$FinalData$end

# Number of watch periods in each year's survey - before the 2019/2020 season
# plus the new ones
periods <-c(136, 135, 164, 178, 
            length(begin.2015), 
            length(begin.2016), 
            length(begin.2020), 
            length(begin.2022))

Watch.Length.2022 <- end.2022 - begin.2022
Watch.Length.2020 <- end.2020 - begin.2020
Watch.Length.2016 <- end.2016 - begin.2016
Watch.Length.2015 <- end.2015 - begin.2015

# 
# #whale counts
n <- labelled::remove_attributes(data.0$n, "dimnames")
n <- n[,,1:4]  # take the first four years
n <- abind(n, array(0, replace(dim(n), 1, max(periods) - nrow(n))), along = 1)

n.2015 <- out.V2.2015$FinalData$n
# use abind for combining multi-dimensional arrays
n <- abind(n, cbind(c(n.2015, rep(0, times = (dim(n)[1] - length(n.2015)))), 
                      rep(0, times = dim(n)[1]))) %>%
  labelled::remove_attributes("dimnames")

n.2016 <- out.V2.2016$FinalData$n
# use abind for combining multi-dimensional arrays
n <- abind(n, cbind(c(n.2016, rep(0, times = (dim(n)[1] - length(n.2016)))), 
                      rep(0, times = dim(n)[1]))) %>%
  labelled::remove_attributes("dimnames")

n.2020 <- out.V2.2020$FinalData$n
# use abind for combining multi-dimensional arrays
n <- abind(n, cbind(c(n.2020, rep(0, times = (dim(n)[1] - length(n.2020)))), 
                      rep(0, times = dim(n)[1]))) %>%
  labelled::remove_attributes("dimnames")

n.2022 <- out.V2.2022$FinalData$n
# use abind for combining multi-dimensional arrays
n <- abind(n, cbind(c(n.2022, rep(0, times = (dim(n)[1] - length(n.2022)))), 
                      rep(0, times = dim(n)[1]))) %>%
  labelled::remove_attributes("dimnames")


# #n <- abind(n,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the sightings array (this is for the day 1 and day 90 zero-whale anchor points)
# 
# 

# the u data is whether there were observers on watch. 
# 0 counts are often associated with years/shifts with 
# no second observer. So if u=0, it will fix observation probability at 0
# the second column for each year is for the second station - not the second
# observer.
u <- data.0$u[,,1:4]
u <- abind(u, 
           array(0, replace(dim(u), 1, max(periods) - nrow(u))), along = 1)

# use abind for combining multi-dimensional arrays
u <- abind(u, 
           cbind(c(rep(1, times = length(n.2015)), 
                   rep(0, times = (dim(u)[1] - length(n.2015)))), 
                 rep(0, times = dim(u)[1])))

u <- abind(u, 
           cbind(c(rep(1, times = length(n.2016)), 
                   rep(0, times = (dim(u)[1] - length(n.2016)))), 
                 rep(0, times = dim(u)[1])))

u <- abind(u, 
           cbind(c(rep(1, times = length(n.2020)), 
                   rep(0, times = (dim(u)[1] - length(n.2020)))), 
                 rep(0, times = dim(u)[1])))

u <- abind(u, 
           cbind(c(rep(1, times = length(n.2022)), 
                   rep(0, times = (dim(u)[1] - length(n.2022)))), 
                 rep(0, times = dim(u)[1])))  %>%
  labelled::remove_attributes("dimnames")

# 
# #visibility
vs <- cbind(rbind(data.0$vs[,1:4], 
                  array(NA, dim = c(max(periods) - nrow(data.0$vs), 4))),
            c(out.V2.2015$FinalData$vs, 
                  rep(NA, times = max(periods) - length(out.V2.2015$FinalData$vs))),
            c(out.V2.2016$FinalData$vs, 
                  rep(NA, times = max(periods) - length(out.V2.2016$FinalData$vs))),
            c(out.V2.2020$FinalData$vs, 
                  rep(NA, times = max(periods) - length(out.V2.2020$FinalData$vs))),
            c(out.V2.2022$FinalData$vs, 
                  rep(NA, times = max(periods) - length(out.V2.2022$FinalData$vs)))) %>%
  labelled::remove_attributes("dimnames")
# vs <- as.matrix(read.table("Data/vs.txt",
#                            header=T,
#                           nrows = max(periods)))

#Add two trailing NAs (this is for the day 1 and day 90 zero-whale anchor points)
#vs <- rbind(vs, matrix(NA,nrow=2, ncol=length(periods))) 
# 
# #beaufort
bf <- cbind(rbind(data.0$bf[,1:4], 
                  array(NA, dim = c(max(periods) - nrow(data.0$bf), 4))),
            c(out.V2.2015$FinalData$bf, 
                  rep(NA, times = max(periods) - length(out.V2.2015$FinalData$bf))),
            c(out.V2.2016$FinalData$bf, 
                  rep(NA, times = max(periods) - length(out.V2.2016$FinalData$bf))),
            c(out.V2.2020$FinalData$bf, 
                  rep(NA, times = max(periods) - length(out.V2.2020$FinalData$bf))),
            c(out.V2.2022$FinalData$bf, 
                  rep(NA, times = max(periods) - length(out.V2.2022$FinalData$bf)))) %>%
  labelled::remove_attributes("dimnames")
# bf <- as.matrix(read.table("Data/bf.txt",
#                            header=T,
#                           nrows = max(periods)))
#Add two trailing NAs (this is for the day 1 and day 90 zero-whale anchor points)
#bf <- rbind(bf, matrix(NA,nrow=2,ncol=length(periods))) 
# 

# observers
# need to convert observer initials into numbers for all years
obs.list <- read.csv(file = "Data/Observer list 2022.csv")

obs.2015 <- data.frame(obs = out.V2.2015$FinalData$obs) %>% 
  left_join(obs.list, by = "obs") 
obs.2016 <- data.frame(obs = out.V2.2016$FinalData$obs)%>% 
  left_join(obs.list, by = "obs") 
obs.2020 <- data.frame(obs = out.V2.2020$FinalData$obs)%>% 
  left_join(obs.list, by = "obs") 
obs.2022 <- data.frame(obs = out.V2.2022$FinalData$obs)%>% 
  left_join(obs.list, by = "obs") 

obs <- data.0$obs[,,1:4]
obs <- abind(obs, 
           array(36, replace(dim(obs), 1, max(periods) - nrow(obs))), along = 1)

obs <- abind(obs, 
             cbind(c(obs.2015$ID, 
                     rep(36, times = (max(periods) - length(obs.2015$ID)))), 
                   rep(36, times = max(periods))))

obs <- abind(obs, 
             cbind(c(obs.2016$ID, 
                     rep(36, times = (max(periods) - length(obs.2016$ID)))), 
                   rep(36, times = max(periods))))

obs <- abind(obs, 
             cbind(c(obs.2020$ID, 
                     rep(36, times = (max(periods) - length(obs.2020$ID)))), 
                   rep(36, times = max(periods))))

obs <- abind(obs, 
             cbind(c(obs.2022$ID, 
                     rep(36, times = (max(periods) - length(obs.2022$ID)))), 
                   rep(36, times = max(periods))))

#obs <- abind(obs,array(0,dim=c(2,2,6)), along=1) #add two trailing 0s to the end of the effort on/off array
# #for(i in 1:length(periods)){ #Place 36s for 'no observer' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
# #  obs[(periods[i]+1):(periods[i]+2),,i] <- 36 #this will force it to the mean observation probability with no observer effect
# #}

#we're going to make N a partially observed data object with anchor points at day 1 and 90
# TE: I don't know how these numbers were created... they are generally 2x n (not all)
# N_inits <- as.matrix(read.table("Data/Initial Values/N_inits.txt",
#                                 header=T))

N_inits1 <- n[, 1,] * 2 + 2
N_inits2 <- n[, 2,] * 2 + 2 
            
N_inits <- N_inits1
N_inits[N_inits1 < N_inits2] <- N_inits2[N_inits1 < N_inits2]

N_inits <- rbind(N_inits,
                 matrix(data = NA, nrow = 2, ncol = length(periods)))

for (k in 1:length(periods)){
  N_inits[(periods[k]+1):nrow(N_inits), k] <- NA  
}

#The 'data' has to be the inverse of the inits, 
# with NAs for all of the estimated Ns, and 0s for the days 1 and 90
N <- matrix(NA, nrow=max(periods)+2, ncol=length(periods)) 

for(i in 1:length(periods)){
  N[(periods[i]+1):(periods[i]+2),i] <- 0 #True number of whales passing fixed at 0 for day 1 and 90
}
```

Set up BUGS:

```{r}
Watch.Length <- cbind(rbind(data.0$Watch.Length[,1:4], 
                            array(NA, dim = c(max(periods) - 
                                                nrow(data.0$Watch.Length), 4))),
                      c(Watch.Length.2015, 
                        rep(NA, 
                            times = max(periods) -                             
                              length(Watch.Length.2015))),
                      c(Watch.Length.2016, 
                        rep(NA, 
                            times = max(periods) -                              
                              length(Watch.Length.2016))),
                      c(Watch.Length.2020, 
                        rep(NA, 
                            times = max(periods) -                              
                              length(Watch.Length.2020))),
                      c(Watch.Length.2022, 
                        rep(NA, 
                            times = max(periods) -                 
                              length(Watch.Length.2022)))) %>%
  labelled::remove_attributes("dimnames")

#specify survey days associated with each watch period
day <- cbind(rbind(data.0$day[,1:4], 
                   array(NA, dim = c(max(periods) - nrow(data.0$day), 4))),
             c(floor(begin.2015), 
               rep(NA, times = max(periods) - length(begin.2015))),
             c(floor(begin.2016), 
               rep(NA, times = max(periods) - length(begin.2016))),
             c(floor(begin.2020), 
               rep(NA, times = max(periods) - length(begin.2020))),
             c(floor(begin.2022), 
               rep(NA, times = max(periods) - length(begin.2022)))) %>%
  labelled::remove_attributes("dimnames")

end <- Watch.Length + day

#t <- round((begin+end)/2)
t <- round((day+end)/2)

# #Add a couple of extra rows of NAs to the end of the day index reference to match up with the fixed 0s in N (above), assigning them to days 1 and 90
day <- rbind(as.matrix(day),
             matrix(NA, nrow=2, ncol=length(periods)))
#
#
for(i in 1:length(periods)){ #Set the anchor points: days 1 and 90
  day[(periods[i]+1):(periods[i]+2),i] <- c(1,90)
}

t <- rbind(as.matrix(t),
           matrix(NA,nrow=2,ncol=length(periods)))
for(i in 1:length(periods)){ #Set the anchor points: days 1 and 90
  t[(periods[i]+1):(periods[i]+2),i] <- c(1,90)
}

#Place 36s for 'no observer' for the two periods following the end of true watches (this is for the day 1 and day 90 zero-whale anchor points)
#this will force it to the mean observation probability with no observer effect

Watch.Length <- rbind(as.matrix(Watch.Length),
                      matrix(NA, nrow=2, ncol=length(periods)))

for(i in 1:length(periods)){
  Watch.Length[(periods[i]+1):(periods[i]+2),i] <- 1
}

#### Two separate models each fit to data, 
# and then a third replicate of data to select
# 'best fit' (same structure as Durban et al)

# Shorten data to first x years only to replicate 
# the analysis in Durban et al 2016:
x <- 8

# n <- n[,,1:x]
# obs <- obs[,,1:x]
# periods <- periods[1:x]
# u <- u[,,1:x]
# vs <- vs[,1:x]
# bf <- bf[,1:x]
# day <-day[,1:x]
# t <- t[,1:x]
# N <- N[,1:x]
# N_inits <- N_inits[,1:x]
# Watch.Length <- Watch.Length[,1:x]

jags.data <- list(n = n[1:max(periods[1:x]),,1:x],
                  n.com = n[1:max(periods[1:x]),,1:x],
                  n.sp = n[1:max(periods[1:x]),,1:x],
                  n.station = dim(n[1:max(periods[1:x]),,1:x])[2],
                  n.year = dim(n[1:max(periods[1:x]),,1:x])[3],
                  n.obs = max(obs[1:max(periods[1:x]),,1:x]),
                  periods = periods[1:x],
                  obs = obs[1:max(periods[1:x]),,1:x],
                  #Watch.Length = 0.0625,
                  u = u[1:max(periods[1:x]),,1:x],
                  vs = vs[1:max(periods[1:x]),1:x],
                  bf = bf[1:max(periods[1:x]),1:x],
                  #day=day,
                  day = t[1:(max(periods[1:x])+2),1:x],
                  N = N[,1:x],
                  N.com = N[,1:x],
                  N.sp = N[,1:x],
                  knot = c(-1.46,-1.26,-1.02,-0.78,
                         -0.58,-0.34,-0.10,0.10,
                         0.34,0.57,0.78,1.02,1.26,1.46),
                  n.knots=14,
                  #begin=begin,
                  #end=end,
                  Watch.Length=Watch.Length[1:(max(periods[1:x])+2), 1:x])

jags.inits <- function() list(mean.prob = 0.5,
                              BF.Fixed = 0,
                              VS.Fixed = 0,
                              mean.prob.sp = 0.5,
                              BF.Fixed.sp = 0,
                              VS.Fixed.sp = 0,
                              mean.prob.com = 0.5,
                              BF.Fixed.com = 0,
                              VS.Fixed.com = 0,
                              mean.beta = c(0,0,0), #mean.beta = c(5,0.14,-3.5),
                              beta.sigma = c(1,1,1),#beta.sigma = c(7,7,7),
                              BF.Switch = 1,
                              VS.Switch = 1,
                              OBS.Switch = 1,
                              sigma.Obs = 1,
                              BF.Switch.sp = 1,
                              VS.Switch.sp = 1,
                              OBS.Switch.sp = 1,
                              sigma.Obs.sp = 1,
                              BF.Switch.com = 1,
                              VS.Switch.com = 1,
                              OBS.Switch.com = 1,
                              sigma.Obs.com = 1,
                              N = N_inits,
                              N.com = N_inits,
                              N.sp = N_inits,
                              #z = matrix(1,nrow=90,ncol=6),
                              beta.sp = array(data=0, dim=c(2,x)),
                              sd.b.sp = rep(1, times = x), #c(1,1,1,1,1,1),
                              z = matrix(1, nrow=90, ncol= x))

#### To run 2006-2019 data, load .RDS object and name it jags.data, 
# then update initial values:
# Files are 2006-2019_GC_Formatted_Data and 2006-2019_GC_N_inits

# WinBUGS gives errors when N inits are set to 0. 
# Try setting them to 1 instead (seems to work):
# But this makes no sense to have it here... 2022-10-05
#N_inits[which(N_inits==0,arr.ind = T)] <- 1

parameters <- c("lambda","OBS.RF","OBS.Switch",
                "BF.Switch","BF.Fixed","VS.Switch",
                "VS.Fixed","mean.prob","mean.prob.com",
                "mean.prob.sp","BF.Fixed.com",
                "BF.Fixed.sp","VS.Fixed.com",
                "VS.Fixed.sp",
                "Corrected.Est","Raw.Est","z",
                "com","sp","Daily.Est","mean.beta",
                "beta.sigma","beta","beta.sp","b.sp","sd.b.sp")

MCMCparams <- list(n.iter = 10000, #0,
                   n.thin = 80,
                   n.burnin = 6000, #0,
                   n.chains = 5)


```

Run WinBUGS if it hasn't been run..

```{r}
if (!file.exists(paste0("RData/WinBUGS_", x, "yr_v2.rds"))){
  
  #Run time: 
  Start_Time<-Sys.time()
  
  BUGS_out <- bugs(data = jags.data,
                  inits = jags.inits,
                  parameters = parameters,
                  model.file="GW_Nmix_Orig.bugs",
                  n.chains = MCMCparams$n.chains,
                  n.iter = MCMCparams$n.iter, 
                  n.burnin = MCMCparams$n.burnin, 
                  n.thin = MCMCparams$n.thin,
                  debug=F,
                  bugs.directory = WinBUGS.dir)
  
  Run_Time <- Sys.time() - Start_Time
  Ver2.results <-list(jags.data = jags.data,
                      N_inits = N_inits,
                      BUGS_out = BUGS_out)
  saveRDS(Ver2.results, 
          paste0("RData/WinBUGS_", x, "yr_v2.rds"))
  
} else {
  Ver2.results <- readRDS(paste0("RData/WinBUGS_", x, "yr_v2.rds"))
}

# The following line is for saving the data list and inits list. This was done to check why simulation
# data were resulting in error - "undefined real result" 
# (ran with n.iter = 1000, n.thin = 8, n.burnin = 600, n.chains = 3) as well as n.iter = 10000,
# n.thin = 80, n.burnin = 6000, n.chains = 5
# data.worked <- list(BUGS.data = jags.data,
#                     BUGS.inits = jags.inits,
#                     N_inits = N_inits,
#                     x = x,
#                     parameters = parameters,
#                     MCMC.params = MCMCparams,
#                     model.file = "GW_Nmix_Orig.bugs",
#                     WinBUGS.dir = WinBUGS.dir)
# 
# saveRDS(file = paste0("RData/BUGS_data_runs_", Sys.Date(), ".rds"), 
#         data.worked)

# clear the workspace, bring in the data that worked.
rm(list = ls())
#
# Runs from the saved list? Yes.
# data.worked <- readRDS(paste0("RData/BUGS_data_runs_2022-10-21.rds"))
# N_inits <- data.worked$N_inits
# x <- data.worked$x
# bm <- bugs(data = data.worked$BUGS.data,
#              inits = data.worked$BUGS.inits,
#              parameters = data.worked$parameters,
#              model.file = data.worked$model.file,
#              n.chains = data.worked$MCMC.params$n.chains,
#              n.iter = data.worked$MCMC.params$n.iter,
#              n.burnin = data.worked$MCMC.params$n.burnin,
#              n.thin = data.worked$MCMC.params$n.thin,
#              debug=T,
#              bugs.directory = data.worked$WinBUGS.dir)

```
array index is greater than array upper bound for Watch.Length
inits(1,C:/Users/TOMO~1.EGU/AppData/Local/Temp/Rtmpy436t0/init

Make some plots:

```{r}
seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011", 
             "2014/2015", "2015/2016", "2019/2020", "2021/2022")

# Extract estimated counts
Daily.Est <- Ver2.results$BUGS_out$sims.list$Daily.Est
sp <- Ver2.results$BUGS_out$sims.list$sp
com <- Ver2.results$BUGS_out$sims.list$com
Corrected.Est <- Ver2.results$BUGS_out$sims.list$Corrected.Est

# Each one of them is (# samples) x (90 days) x (# years)
# To plot them using ggplot's facet, I need to convert
# these into 2D dataframes of statistics (upper and lower 
# CIs, median, etc.)
# Daily.Est.list <- sp.list <- com.list <- vector(mode = "list", 
#                                                 length = dim(Daily.Est)[3])
# 
# Daily.Est.UCIs <- Daily.Est.LCIs <- vector(mode = "list",
#                                            length = dim(Daily.Est)[3])

stats.list <- vector(mode = "list",
                     length = dim(Daily.Est)[3])

for (k in 1:dim(Daily.Est)[3]){
  # Daily.Est.list[[k]] <- Daily.Est[,,k]
  # Daily.Est.UCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.975)
  # Daily.Est.LCIs[[k]] <- apply(Daily.Est[,,k],2,quantile,0.275)
  # 
  # sp.list[[k]] <- sp[,,k]
  # com.list[[k]] <- com[,,k]
  
  stats.list[[k]] <- data.frame(Daily.Est.median = apply(Daily.Est[,,k], 2,
                                                         median),
                                Daily.Est.LCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.275),
                                Daily.Est.UCL = apply(Daily.Est[,,k], 2,
                                                      quantile,0.975),
                                sp.median = apply(exp(sp[,,k]), 2,
                                                  median),
                                sp.LCL = apply(exp(sp[,,k]), 2,
                                               quantile,0.025),
                                sp.UCL = apply(exp(sp[,,k]), 2,
                                               quantile,0.975),
                                com.median = apply(exp(com[,,k]), 2,
                                                   median),
                                com.LCL = apply(exp(com[,,k]), 2,
                                                quantile,0.025),
                                com.UCL = apply(exp(com[,,k]), 2,
                                                quantile,0.975),
                                #total.median = apply(exp(sp[,,k]), 1, sum),
                                days = 1:dim(Daily.Est)[2],
                                year = seasons[k])
}

all.stats <- do.call("rbind", stats.list) %>% group_by(year)

ggplot(data = all.stats) + 
  geom_line(aes(x = days, y = sp.median)) + 
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", 
              alpha = 0.5) +
  geom_line(aes(x = days, y = com.median),
            color = "red") +
  facet_wrap(vars(year)) +
  xlab("Days since December 1") + 
  ylab("Whales per day")

```



```{r}

#Com vs Sp
ggplot(data = all.stats) +
  geom_line(aes(x = days, y = sp.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = sp.LCL, 
                  ymax = sp.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day (spline)")



```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = com.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = com.LCL, 
                  ymax = com.UCL),
              fill = "orange", alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day (Normal)")

```


```{r}

ggplot(data = all.stats) +
  geom_line(aes(x = days, y = Daily.Est.median)) +
  geom_ribbon(aes(x = days, 
                  ymin = Daily.Est.LCL, 
                  ymax = Daily.Est.UCL),
              fill = "orange", 
              alpha = 0.5) + 
  facet_wrap(vars(year))+
  xlab("Days since December 1") + 
  ylab("Whales per day")


```

Total abundance
```{r}
abundance.df <- data.frame(total.median = apply(Corrected.Est, 
                                                FUN = median, 
                                                MARGIN = 2),
                           total.LCL = apply(Corrected.Est, 
                                             MARGIN = 2, 
                                             FUN = quantile, 0.025),
                           total.UCL = apply(Corrected.Est, 
                                             MARGIN = 2, 
                                             FUN = quantile, 0.975),
                           Season = seasons)


ggplot(data = abundance.df) + 
  geom_point(aes(x = Season, y = total.median)) + 
  geom_errorbar(aes(x = Season, ymin = total.LCL, ymax = total.UCL))

#median(apply(exp(sp[,,4]),1,sum))

```

